### 2021-06-07 jsdse/Manuscript.tex
starting to address comments of final reviews

%&latex
\documentclass[12pt]{article}
\usepackage{amsthm,amsmath}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{url} % not crucial - just used below for the URL
\usepackage{amsthm,amsmath}
\usepackage[utf8]{inputenc}

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{1}

% DON'T change margins - should be 1 inch all around.
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-.5in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.3in}%
\addtolength{\topmargin}{-.8in}%

\begin{document}

%\bibliographystyle{natbib}
%\bibliographystyle{bmc-mathphys}
\bibliographystyle{agsm}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\if0\blind
{
  \title{\bf Approachable case studies support learning and reproducibility in data science: An example from evolutionary biology}
  \author{Luna L. Sanchez Reyes\thanks{
    The authors gratefully acknowledge ``Sustaining the Open Tree of Life'',
    NSF ABI No. 1759838, and ABI No. 1759846. They are also also deeply grateful towards
    ``The Carpentries'' organization; without its invaluable volunteers and resources
    this project could not have been possible}\hspace{.2cm}\\
    School of Natural Sciences, University of California, Merced\\
    and \\
    Emily Jane McTavish \\
    School of Natural Sciences, University of California, Merced}
  \maketitle
} \fi

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Approachable case studies support learning and reproducibility in data science: An example from evolutionary biology}
\end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
Research reproducibility is essential for scientific development. Yet, rates of reproducibility are low, especially in the natural sciences. 
As increasingly more research relies on computers and software, efforts for improving reproducibility rates have focused on making available research workflows as computer code, as well as raw and processed data in computer readable form. 
However, research products that are digitally available are not necessarily friendly for learners and interested parties with little to no experience in the field. 
This renders research products unapproachable, counteracts availability, and hinders reproducibility short and long term. 
To improve long term adoption of reproducible workflows in research, they need to be made approachable for learners, the researchers of the future.

Using a case study within evolutionary biology, we identify aspects of research workflows that make them unapproachable to the general audience: 
use of highly specialized language; 
unclear goals and high cognitive load; and 
content-focused descriptions. 
We propose a set of principles to improve the unapproachable aspects of research workflows, and illustrate their application in a tutorial. 
We elaborate on the general application of these principles for documenting research workflows and products, to provide present learners and future researchers with tools for successful scientific reproducibility. 
%> ** add a sentence on implications for teaching reproducibility

%% Projects that address availability of code and data in the natural sciences
%% have focused on producing documentation that describes the nature and usage of the
%% resources.
%% However, it often communicates in a highly specialized language that is largely
%% inaccessible for the average target user. Additionally, baseline
%% computational knowledge and required skills to access scientific results are often
%% not found among potential users, and keep on increasing.

\end{abstract}

\noindent%
{\it Keywords:}  open science, R programming language, phylogenetics, Open Tree of Life, pedagogy, phylogenetics, datelife
\vfill

\newpage
\spacingset{1.45} % DON'T change the spacing!
\section*{Introduction}
\label{sec:intro}

Research reproducibility --the extent to which consistent results are obtained when
a scientific experiment or research workflow is repeated \citep{repdef2021}-- is a key aspect of the advancement of science, as it constitutes a minimum standard that allows understanding research products, i.e., methods, data, analysis, results, etc. \citep{piwowar2013value}, to determine their reliability and generality, and eventually build up scientific knowledge and applications based on those products \citep{king1995replication, peng2011reproducible, powers2019open}.
In the natural sciences, rates of reproducibility are low \citep{ioannidis2005most, prinz2011believe}, which has elicited concerns about a crisis in the field \citep{baker2016reproducibility}.

In response, the scientific community has been developing new principles and standards to incentivize cultural changes that support a long term improvement of reproducibility rates in the natural sciences \citep{peng2015reproducibility, wilkinson2016fair, miyakawa2020no}.
A standard for reproducibility that has received much attention is availability, which we define as a property denoting that a research product can be reached (acquired, copied, analyzed, processed and/or reused) at no financial, legal or technical cost \citep{arnold2019turing}, and without geographic, demographic, social or temporal barriers for the population \citep{fecher2014open}.

In this paper, we argue that research products that are digitally available are often unapproachable in practice, because they are not friendly for learners and interested parties with different levels of experience in the field. 
Research products that are unapproachable counteract availability, and hinder reproducibility short and long term.
To support long term adoption of reproducible practices in the natural sciences, research workflows need to be made approachable for learners, the researchers of the future \citep{roland2002think}.

To elaborate on our thesis, we designed a case study within the research field of phylogenetics, a discipline within evolutionary biology.
We use our case study to identify barriers that have made research workflows largely unapproachable to a general audience in the natural sciences.
Then, we propose some principles for researchers to address these barriers, and create research workflows that are reproducible by a larger audience.
% ** describe/introduce reproducibility curriculum here
The principles proposed here can be generalized and integrated into the undergraduate and graduate school STEM curriculum, either for courses specialized in reproducibility or within other subject areas, as a necessary component of successful and impactful science.

%% \bigskip
%% \bigskip
%%
%% \noindent\fbox{%
%%     \parbox{\textwidth}{%
%%     \textbf{Box 1. The role of approachable research workflows in teaching reproducibility.}
%%       % Availability is defined as the fact or possibility that something can be bought,
%%       % used, or reached by someone \citep{availability2022cambridge}.
%%       % Accessibility is defined as "the fact of being able to be reached or obtained easily"
%%       % \citep{accessibility2022cambridge}
%%       % Yet, in practice, availability does not imply accessibility.
%%       Availability and accessibility are conceptually accepted as synonyms for ``the fact
%%       of being able to be reached" \citep{available-accessible2022cambridge}.
%%       However, accessibility's definition goes further and describes it as the
%%       ``fact of being able to be reached \textbf{easily}" \citep{accessibility2022cambridge}.
%%       A secondary definition of accessibility elaborates more on this: ``the quality
%%       of being \textbf{easy to understand}''.
%%
%%       What we realize from these definitions is that, in practice, something might
%%       be available but not necessarily accessible,
%%       hindering the realization of availability.
%%       To illustrate the practical difference between availability and accessibility we
%%       will use an allegory that goes like this:
%%
%%       ** Change marshmallow story to cake
%%
%%       ** Explain different types of barriers to accessibility within the example, then within science
%%
%%       ** Explain why we focus on psychological/social/pedagogical barriers
%%
%%       ** Gauging accessibility of a resource is tricky because it depends on an
%%       individual experience of what is in effect \textbf{easy} for someone.
%%       \textbf{Easy} is a subjective quality that fully depends on individual perceptions
%%       resulting from individual contexts.
%%       As a society we are moving towards validating and welcoming all human experiences.
%%       To improve accessibility levels in the natural sciences, it
%%       is important to account for the factors that are creating different experiences
%%       of what is \textbf{easy}.
%%
%%       ** Now, explain different types of barriers to accessibility in science
%%       (Jessica: Maybe say that lack of accessibility can be due to a number of
%%       factors (physical, resources, economic), but one critical factor is the use
%%       of impenetrable expert language.)
%%     }%
%% }
%% \bigskip

\section*{A case study from phylogenetics}
\label{sec:case}

Phylogenetics is a key discipline within evolutionary biology \citep{dobzhansky1973nothing}.
It focuses on investigating the history of shared ancestry of living and extinct organisms using biological data, and represents this evolutionary history with a diagram known as a phylogeny or phylogenetic tree (because it grows through time and appears to have branches; Figure \ref{fig:tree}).
Phylogenies provide the basis to study and understand all biological processes in an evolutionary context \citep{dobzhansky1973nothing}.
Hence, it follows that improving reproducibility rates in phylogenetics has the potential to positively impact research across the natural sciences.

\begin{figure}
\begin{center}
\includegraphics[width=3in]{fig-tree.png}
\end{center}
\caption{A phylogenetic tree from our tutorial. It was extracted using OpenTree of Life resources \citep{opentreeoflife2019synth} wrapped in the \texttt{rotl} R package \citep{michonneau2016rotl}. \label{fig:tree}}
\end{figure}

To explore barriers to approachable phylogenetics, we develop a case study that touches on three common problems within the field: standardizing organism names in phylogenies, obtaining current phylogenetic knowledge for a group of organisms, and summarizing this phylogenetic knowledge in a meaningful way.
To address these problems, we propose a research workflow that relies on resources from the Open Tree of Life (OpenTree), an open source project that provides digital availability of phylogenetic results from published, peer-reviewed research, which is considered as vetted and state-of-the-art knowledge in the field.
OpenTree phylogenies are stored in a public database, the Phylesystem \citep{mctavish2015phylesystem}, and are downloadable as various computer-readable file types, which is key for reusability and reproducible workflows \citep{wilson2017good}.
OpenTree also provides access to a single standard for organism names (taxonomic standard) that is applied to the stored phylogenies \citep{rees2017automated}, which are then used to summarize a single phylogenetic tree encompassing all life \citep{opentreeoflife2019synth}.

All of these resources 
%% (the stored phylogenies, the taxonomic standard, the synthetic tree, and all
%% tools developed to process and summarize stored phylogenies),
are available for download and use from OpenTree free of cost to any user.
One way to access OpenTree resources is through its Graphical User Interface (GUI; aka, a website or application that allows users to access and use functionalities with mouse or keyboard clicks).
However, reducing as many manual steps as possible in research workflows is key for reproducibility, as manual data manipulation scales poorly and is prone to error \citep{bakken2019journey}.
OpenTree's resources are also programmatically available through its Application Programming Interface services (APIs; aka, computer code that implements computer functionalities, in this case OpenTree resources, that can be used by programmers
to build more functionalities), which provide scalability and reproducibility \citep{opentreeAPIv3}.
However, this comes at a high cost for the user, which requires considerably more computer programming experience and literacy to be able to successfully use APIs.
%% In recent years, OpenTree API services have been wrapped into software packages for R and Python
%% \citep{michonneau2016rotl, mctavish2021opentree}, open source, free of cost
%% programming languages that represent two of the most widely used programming languages
%% in the sciences \citep{baker2017scientific}. These packages should contribute to making
%% OpenTree resources more accessible to a wider user audience.
The \texttt{rotl} R package \citep{michonneau2016rotl} and the \texttt{opentree} Python module \citep{mctavish2021opentree} have been developed to wrap OpenTree's API services
and make them available through the R and Python programming languages.
R and Python are open source and free of cost, and represent two of the most widely used programming languages in the sciences today \citep{eglen2009quick, baker2017scientific}.
As such, \texttt{rotl} and \texttt{opentree} software packages should contribute to making OpenTree resources more accessible to a wider programming user audience.

However, while learners in the natural sciences have been engaging independently with R and Python programming languages, computer programming is not traditionally a core skill formally taught to biologists and naturalists \citep{sayres2018bioinformatics, wright2019the, williams2019barriers}.
As computers continue to play a larger role in most scientific disciplines \citep{piccolo2016tools}, higher baseline computational skills are required across all natural sciences not only to develop an original research workflow, but to be able to follow and reproduce research workflows from other researchers.

Thus, efforts to increase reproducibility rates long term in the natural sciences would benefit from addressing specific barriers for learners in the field, to support them in acquiring the skills needed to reproduce research workflows that rely heavily on computer code \citep{peng2011reproducible, sandve2013ten, powers2019open}.

In the next section, we describe (in no particular order) the barriers to approachable research workflows that we identified on our case study.
Then we develop a set of principles to address these barriers, and apply the latter to a set of teaching materials that are available at https://mctavishlab.github.io/R\_OpenTree\_tutorials/.

\section*{Identifying barriers to approachable research workflows}
\label{sec:identifying}

% Cite carpentries and Mosi

The main goal of our case study is to obtain a single phylogeny summarizing data from a set of published phylogenies for the canids (the family of dogs, coyotes, wolves, etc.), our organism of study.
All analysis for our case study can be completely accomplished using functions from the R package \texttt{rotl} or the Python module \texttt{opentree}. 
If a researcher were to use the proposed analysis workflow in a publication, they would typically describe it in the methodology section as ``The canid summary phylogeny was obtained using functions from X package, details are available as supplementary files".
This is usual practice, mainly because journals do not have space to publish all code used for an analysis in the methods section. Yet, supplementary files and data have the misfortune to not be peer-reviewed as thoroughly (or at all) as the main manuscript \citep{pop2015use}.
They are also prone to the dreaded promise ``available upon request", which has very low rates of fulfillment \citep{krawczyk2012available}.
Without the primary data and code that was used to perform an analysis it is impossible to reproduce said analysis \citep{miyakawa2020no}. When the code is available, other issues can complicate reproduction of the analysis, to the point of completely obstructing reproducibility.
For example, some questions that are often unaddressed in research workflows are: what software do I need to open the code files?
Can I run the code with the same software I used to open the files, or do I need a different one?
Do I need additional software that the analysis depends on?
What does the code even mean?

These questions can usually be answered by referring to the software documentation, which are usually publicly available and can be accessed by any potential user.
As opposed to code, software documentation is written in natural language (i.e., any known human language, e.g., English, Spanish, Chinese), and is considered a key element for successful adoption of software by target users \citep{karimzadeh2018top}. This might explain why documentation for software addressed to academic users is also usually written using highly specialized computational language or jargon (i.e., computationally specific concepts, words, and phrases) as well as formal scientific and academic language.
We identify this as \textit{barrier 1} to approachable research workflows.-- \textit{Specialized language is intimidating}.
While scientific jargon might have an important role for formal acceptance of software by the scientific and academic community, it can be perceived as cold and/or intimidating language, that often slows down or even obstructs examination, application, and adoption of code by a wider audience \citep{ball2017its}.
In contrast, introducing information without the use of jargon, supports learners' conceptual understanding of new ideas and concepts \citep{mcdonnell2016concepts, pan2019online}.
% jargon creates a hostile environment that does not foster learning \citep{CITATION}.

Another element of good software documentation is that it has to be thorough \citep{karimzadeh2018top}.
Meaning that it should describe general usage of individual functions, as well as arguments and variables that said function can take \citep{karimzadeh2018top}.
Individual documentation for each function is usually presented in alphabetic order, and does not have a specific analysis goal.
Moreover, most software has numerous functions, so documentation is usually very lengthy and it is hard to navigate.
This can have the effect of increasing the amount of information that needs to be simultaneously processed by the users, which can lead to overload of the finite amount of working memory any one possesses, known as cognitive load \citep{sweller1988cognitive}.
In this context, identifying connections across functions that are meant to work on the same analysis workflow can become a very difficult task.
We recognize this as \textit{barrier 2.-- Lack of specific goals leading to high cognitive load}.
High cognitive load is know to have a negative effect in learning software \citep{chandler1996cognitive, van2005research, lambert2009student}.

A third important aspect of software documentation are examples that demonstrate usage of individual functions \citep{karimzadeh2018top}.
Examples presented in software documentation are usually worked to perfection, as they are intended to showcase the ideal or minimal case in which a function works well.
Perfectly worked examples ignore the user experience by maintaining focus on the software content, and fail to provide users with expert and clear advice on how to troubleshoot if needed.
We identify this as \textit{barrier 3.-- Content-focused examples}.
Error management training is an approach that focuses on framing mistakes as beneficial to learning complex tasks, to give learners the opportunity to
actively explore a task with a positive mindset \citep{frese1995error}.
Providing examples that showcase potential errors supports users performance \citep{steele2014error}, and can greatly improve learners ability to troubleshoot outside the classroom \citep{shannon2015live, nederbragt2020ten}.

%% Last but not least, software documentation is usually explained verbally in a document, very
%% rarely is it approached using diagrams or other allegories.
%% We recognize this as \textit{barrier 4.-- Little to no diversity of representation of information}.

In sum, best practices for good software documentation are not enough to promote reproducibility of published research workflows that rely heavily on code. 
In the following section we describe some principles that can help to reduce or remove the identified barriers, to create research workflows that are more approachable and hence more reproducible by a larger audience.

\section*{Principles for approachable research workflows}
\label{sec:addressing}

\subsubsection*{Principle 1. Use friendly, relatable and respectful language}

% avoid specialized, intimidating and cold language

Besides avoiding formal language, and incorporating elements of pop culture, such as picture character icons known as ``emojis``, to make the language more familiar to a broader target audience (Figure \ref{fig:error}), we made an effort to specifically complement the primary documentation by identifying computational concepts that were assumed or were not explained in depth.
We vetted the tutorials through feedback from workshop participants as well as individual users to identify such specialized concepts.

\begin{figure}
\begin{center}
\includegraphics[width=3in]{fig-error.png}
\end{center}
\caption{Snapshot of a section of the tutorial website, where we demonstrate a common error. \label{fig:error}}
\end{figure}

\subsubsection*{Principle 2. Reduce cognitive load by providing specific and clear goals with literate programming}

Cognitive load can be greatly reduced for learners by applying an active learning strategy such as linking usage to a ``real world'' or ``human'' application \citep{felder2009active}.
Programming computer languages are by themselves quite abstract and represent a learning subject with a potentially high cognitive load for most learners.
Pedagogical research shows that active learning practices are one of the most effective ways to take on abstract subjects \citep{freeman2014active}.
A story-like narrative that links code usage in an integrative example, invites learners to try the code, which can lead them to remember what they are doing and why they are doing it.
This ``literate programming'' paradigm \citep{knuth1984literate, fritzson2002mathmodelica} makes code more approachable, as it integrates narratives with computer code in the same document, supporting learners in actively following the code usage, supporting memory and understanding \citep{piccolo2016tools}.

We propose that documents developed with ``literate programming'' can be made more accessible by choosing narratives that are relatable to a more general audience.
An easy way to do this in biology is choosing a charismatic taxon as a model organism.
For a research group this can be the group they are studying. For the general audience, a highly charismatic group such as dinosaurs should work well.
For example, when we presented our tutorial to the Amphibia Web Organization \citep{van2002amphibiaweb} in January 2020, we tailored all examples to frogs and their allies.
% \citep{cite-aw-workshop}

We examined available software documentation for the package \texttt{rotl}, and designed a narrative that requires the usage of as many functions as possible.
We demonstrate code applications that are commonly requested by OpenTree users, but that are not demonstrated in the documentation of the package.
By framing the function workflow using highly requested uses, the documentation acquires a narrative arc that is easier to follow and remember by users. This can also facilitate translating the code application to other use cases of interest for learners in biology.

%% Maybe Figure!! In the tutorial showcased here, we used the commonly requested
%% use case of obtaining
%% a phylogenetic tree for all lineages within a specific taxonomic rank.

\subsubsection*{Principle 3. Provide examples that are user-focused by demonstrating errors and warnings}

% approachable practice: user focused examples (instead of content focus)
% demonstrate errors and warnings thoroughly

An activity that has become increasingly widespread in programming-language education is live programming. During live programming, and instructor writes code and executes it in a way that is visible to all learners through a screen \citep{guzdial2013lure, selvaraj2021live}.
One benefit of this practice is that typos and mistakes occur, normalizing them for learners. 
Watching an instructor handling errors, demonstrates learners on how to solve them when they are outside theclassroom \citep{shannon2015live, nederbragt2020ten}.
When co-author McTavish was a postdoc, teaching an introductory programming workshop as a volunteer with the Carpentries (a non-profit group that teaches foundational coding and data science skills to researchers worldwide) \citep{wilson2006swc, SWCwebsite},
a senior faculty member taking the workshop complained that the typos were slowing things down and interfering with the pedagogy.
McTavish replied "the typos ARE the pedagogy".
This has become a slogan of sorts at the Carpentries, capturing the idea that embracing and discussing mistakes is essential to teaching programming \citep{wilson2019teaching}.
Yet, working through mistakes is rarely done on written pedagogical materials.
Software documentation focuses on demonstrating usage function with examples that work seamlessly, without errors. 
We argue that the opposite is needed to support adoption of reproducible workflows and support long term independence in learners and users performance \citep{gaspar2007restoring, steele2014error}.

In our tutorials, we apply this principle by demonstrating examples that do not work as expected, and exemplifying ways to address them (Figure \ref{fig:error}).
For example, we identified inputs that would give a wide range of warnings and errors. We then focus on providing explanations for these errors and warning messages. We believe this supports users and learners to be less intimidated by the messages, and to practice taking useful information out of them.
% This is probably part of the computer science curriculum, but it is not at all
% part of the natural sciences curriculum.

We also demonstrate ways to evaluate inputs to determine if they will trigger an error or warning, and design and demonstrate alternative analysis routes on what to do when faced with an error or warning.
One of the most essential skills in programming is interpreting and moving forward from errors.
This has two pedagogical benefits. First, it provides users and learners with the means to troubleshoot their own warnings and errors. Second, it allows them to understand with more depth what the function is doing.

% \subsubsection*{Principle 4. Demonstrate one thing in different ways}

% - and little to no diversity of representation of information (say the same thing with different images, metaphors, stories)

% We designed ways to access the different elements of the outputs.

\section*{Conclusion}
\label{sec:conclusion}

Response from the community has been invaluable in gauging success of our teaching materials.
Senior researchers often comment on the usefulness of the tutorials for their research, as well as how they have supported students in using the demosntrated R packages more independently.

When developing our tutorials, we not only applied the principles elaborated here,
but we followed basic recommendations for successful reproducibility \citep{sandve2013ten}.
For example, we published the tutorials on a persistent and public website \citep{RopentreeTutorials} that adheres to the four r's of openness, by being free license, free of cost, and thus free for use, reuse, redistribute, revise and remix \citep{hilton2010four}.
To make the website persistent, any updates to the tutorial are published as new versions.
Versions presented at workshops are a copy from the original repository, and constitute a temporally stable snapshot of functions and workflows presented during a live workshop \citep{wilson2006swc, SWCwebsite}.
This ensures that the tutorials are available for the users to return to any time they need it, and to be shared with other users and learners (Figure \ref{fig:schedule}).

\begin{figure}
\begin{center}
\includegraphics[width=3in]{fig-schedule.png}
\end{center}
\caption{Snapshot of the home to our tutorial website, showing part of the schedule.
 Our tutorial website was constructed using The Software Carpentries workshop template \citep{swc-workshop-template}. \label{fig:schedule}}
\end{figure}

We note that making approachable research workflows has all the advantages of reproducible research workflows.
It saves time during explanation and training, when analyses are run by new collaborators and students.
It can also save research time for yourself, when analyses are run again with more data, a different dataset, a different organism or biological model.
It contributes to scientific efforts that can build off of each other.

% \citep{roland2002think}
Some universities and academic groups have started to incorporate reproducibility as a subject into their curriculum
\citep{uwlibraries2022, nigms2022, psyteachr2022}.
The focus of these resources has been for students to acquire and practice skills to document their work.
The principles identified and outlined here can be used to set learning goals and outcomes for new reproducibility syllabi.
Ultimately, the long term improvement of reproducibility rates in science will depend on our ability to intentionally integrate best practices for achieving reproducibility into the educational framework of future data scientists \citep{nasem2018data}.
Inclusion of reproducibility into the data acumen of undergraduate curriculum will provide college learners and future researchers with the tools to develop the fundamental skills needed to successfully create reproducible scientific workflows and research products.

The principles to create tutorials described here facilitate adoption of software and analysis workflows among researchers at different academic levels, from undergrads to established researchers.
It will also help close the gap between students that had access to computational resources (and computational training) from an early age and students that did not.
Late access to computational resources and training can occur due to lack of economic resources, often occurring in households from underrepresented communities and minorities \citep{google2016diversity, warner2021quantifying}.
It can also be due to gender-biased parental and community pressures, in which male identifying individuals are more often encouraged to perform activities related to computers, while female identifying individuals are discouraged \citep{warner2021quantifying}.
These principles can be used to improve not only reproducibility practices, but also software adoption in the natural sciences.

\bigskip
\begin{center}
{\large\bf SUPPLEMENTARY MATERIAL}
\end{center}

\begin{description}

\item[Title:] Website and GitHub repository containing the complete teaching materials developed and demonstrated here.

\item[GitHub repository link:] \url{https://github.com/McTavishLab/R_OpenTree_tutorials}

\item[Website link:] \url{https://mctavishlab.github.io/R_OpenTree_tutorials}

\end{description}

\bibliography{Manuscript-bibliography}

\end{document}

### 2021-09-16 jsdse/TAS-template.tex
first latex version for jsdse

%&latex
\documentclass[12pt]{article}
\usepackage{amsthm,amsmath}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{url} % not crucial - just used below for the URL
\usepackage{amsthm,amsmath}
\usepackage[utf8]{inputenc}

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{0}

% DON'T change margins - should be 1 inch all around.
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-.5in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.3in}%
\addtolength{\topmargin}{-.8in}%

\begin{document}

%\bibliographystyle{natbib}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\if0\blind
{
  \title{\bf Promoting analysis reproducibility with accessibility: An example in evolutionary biology}
  \author{Luna L. Sanchez Reyes\thanks{
    The authors gratefully acknowledge "Sustaining the Open Tree of Life", NSF ABI No. 1759838, and ABI No. 1759846.}\hspace{.2cm}\\
    School of Natural Sciences, University of California, Merced\\
    and \\
    Emily Jane McTavish \\
    School of Natural Sciences, University of California, Merced}
  \maketitle
} \fi

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Promoting analysis reproducibility with accessibility: An example in evolutionary biology}
\end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
The text of your abstract.  200 or fewer words.
\end{abstract}

\noindent%
{\it Keywords:}  open science, education, R, phylogenetics, tutorials
\vfill

\newpage
\spacingset{1.45} % DON'T change the spacing!
\section{Introduction}
\label{sec:intro}

Reproducibility --the extent to which consistent results are obtained when a scientific experiment is repeated (\cite{repdef2021})-- is a key aspect for the advancement of science, as it constitutes a minimum standard to understand scientific results, to determine their reliability and generality, and eventually to build more scientific knowledge upon those results (\cite{king1995replication, peng2011reproducible, powers2019open}).
Reproducibility rates in the natural sciences are low (\cite{ioannidis2005most, prinz2011believe}), prompting concerns about a reproducibility crisis in the field (\cite{baker2016reproducibility}). The scientific community has united to incentivize cultural changes that will improve reproducibility rates long term, such as transparency, availability, and workflow automatization, to name a few (\cite{peng2015reproducibility}).
We argue that accesibility is a key aspect that must accompany availability (Box 1) in order to achieve full workflow automatization and reproducibility.

\bigskip
\bigskip

\noindent\fbox{%
    \parbox{\textwidth}{%
    \textbf{Box 1.}
      Availability does not imply accessibility. One example we really like is the
      marshmallows in an office story.
      The marshmallow bags were there, availabe for anyone in the office to eat
      the marshmallows inside. But the marshmallows stayed there for days, weeks even.
      And it was not until someone opened the bag and put the marshmallows in a tray,
      that people started actually eating them. They were gone in a matter of hours.
      Code is not marshmallows. But the point is that by
      sharing your code and documenting it might not be enough for the general researcher
      to reproduce results. Researchers will certainly be able to eventually figure it out,
      but the time needed
      to do so might not be worthy for them, or might not be something they can invest
      in, even if it would be useful for them long term, the short term investment
      is too intense.
    }%
}
\bigskip

We focus on identifying factors that have specifically affected accessibility in the natural sciences, and ways researchers can address it to ensure reproducible and automatic workflows.
We use an example in evolutionary biology. This example focuses on understanding the shared ancestry among organisms through evolutionary trees. These phylogenetic trees provide the basis to study and understand biological processes (\cite{dobzhansky1973nothing}). Hence, improving reproducibility and availability of phylogenetic research is relevant for many aspects of biological research.
The Open Tree of Life project (OpenTree) has developed a platform that facilitates availability of results from phylogenetic research, by standardizing and storing phylogenetic data with the goal of synthesizing a single phylogenetic tree encompassing all life (\cite{opentreeoflife2019synth}).
All data in OpenTree is open access and available programmatically through its many Application Programming Interface (API) services (\cite{opentreeAPIs}).
Additionally, R packages (\cite{michonneau2016rotl}) and Python libraries (\cite{mctavish2021opentree}) have been developed as wrappers for OpenTree API services to make them available to a wider programming audience.
The R and Python OpenTree wrappers have been utilized by computer-literate individuals, to seamlessly establish reproducible workflows to use and reuse expert phylogenetic knowledge for biological research (\cite{sanchez2019datelife}) and education (\cite{nguyen2020phylotastic, phylotasticedtools, galacticedtools}).

The OpenTree project demonstrates that efforts to increase reproducibility and availability have also increased the baseline required computational knowledge and skills in phylogenetic research.
These computational skills requirements are likely increasing across all natural sciences.
Computing is not traditionally a core skill taught to biologists and naturalists.
However, many students are now being trained in R as a statistics and data analysis environment.
In order for reproducible computational tools to be adopted for research, they much be accessible to reseachers.
Data and code availability is a core requirement for reproducibility research (\cite{peng2011reproducible, sandve2013ten, powers2019open}).
However, the utility of data resources is limited by the technical challenges of accessing the data.
In order to motivate reproducible research, that gap needs to be bridged.
In this work we focus on improving accessibility of code examples and documentation.
In particular, we identify the necessity for documentation that is written down using language that is common to the target audience to facilitate examination, application, and adoption of code by the wider audience.

We present a set of principles to generate documentation that improves accessibility of code and documentation. We applied these principles to a series of tutorials and vignettes for the OpenTree project.

\begin{figure}
\begin{center}
\includegraphics[width=3in]{fig1.pdf}
\end{center}
\caption{Principle 1. \label{fig:first}}
\end{figure}

\section{Methods}
\label{sec:meth}
\subsection*{Identifying hurdles to accesibility}

Good primary documentation for code describes general usage of individual functions, the components and variables a function can take, and it should be accompanied with function usage examples on how to apply it.
As opposed to code, primary documentation is written in natural language (i.e., any known human language, e.g., English, Spanish, Chinese) and usually makes use of highly specialized computational jargon (computationally specific concepts, words, and phrases) as well as formal language, which often slows down or even obstructs examination, application, and adoption of code by external individuals.
Because primary documentation is considered a professional document, acceptance of the research by the scientific community could be reduced if a more informal language is used.
Secondary types of documentation, such as vignettes and tutorials, demonstrate additional cases of individual function usage, and describe analysis workflows in more detail, as well as function associations to generate a specific analysis and results. While secondary documentation has become more common practice, it is still often generated using highly specialized language.

\subsection*{Adressing hurdles to accesibility: the principles}

\subsubsection*{1. Demonstrate integration of function usage with motivating examples}

We examined primary documentation from the OpenTree API R wrapper package `rotl`.

We designed a workflow that visits as many functions as possible, while answering to uses that are most commonly requested by users of OpenTree.
By framing the use cases in highly requested tasks, the documenation follows a narrative conceptual arc, making the translation to biological use cases clear.

> [name=Emily Jane McTavish]Brief para or figure on what the motivating example is here

\subsubsection*{2. Demonstrate errors and warnings thoroughly}

Primary documentation focuses on demonstrating usage function with examples that work seamlessly without errors. We argue that the opposite is needed to support user adoption of reproducibile workflows: demonstrate examples that do not work as expected and exemplify ways to address them. We identify inputs that would give a wide range of warnings and errors, focusing on demonstrating these cases. This helps users to not be afraid of errors and warnings, but instead to use them to their advantage (CITE carpentries).
We also identify effects of warnings and errors downstream of the workflow.

We identify ways to evaluate inputs to know if they will produce an error, and design alternatives on what to do when faced with an error or warning, and demonstrate these alternatives.
One of the most essential skills in programming is interpreting and moving forward from errors.
Many finely honed tutorials do not trigger errors, which preculdes helping studnts to develop the tools to understand and address errors when the do encounter them, as they inevitably will!

> [name=Emily Jane McTavish]Show where you deal with an error in your tutorial

\subsubsection*{3. Avoid jargon and expert language}

We focused on complementing the primary documentation by identifying computational concepts that were assumed or were not explained in depth.
We vetted the tutorials with an audience on workshops as well as individual user

We choose examples that are charismatic for the audience.

> [name=Emily Jane McTavish]Specify what one you used!

\subsubsection*{4. Make it stable through time}

We published the tutorials on a public, free license, free of cost, and free for use and reuse repository and persistent website.
The tutorial is available for the users to go back to it any time they need it, and to be passed on to other users.

Following the carpentries [@CITE], we created a main version of the tutorial that is updated. Versions presented on workshops are a copy from the original repository, and represent a stable and temporal snapshot of the functions and workflows presented in the tutorial.

> [name=Emily Jane McTavish] Show the page, or the layout, or something :P

\subsubsection*{5. The (now) classics of computational reproducibility}

Provide all information on package version and system capabilities.

> [name=Emily Jane McTavish]link to where/and how you did that

\section{Results and Discussion}
\label{sec:results}

We explain the warnings and errors and design ways to avoid them, and detect them beforehand (i.e., before using an input that would give an error). We explain the consequences of warnings.
We designed ways to access the different elements of the outputs.
We have received emails from senior researchers thanking us for this materials, and students have been able to engage using the packages with less hep from the PIs.

The principles to create tutorials described here facilitate adoption of software and analysis wokflows among researchers at different academic levels, from undergrads to established researchers.
It will also help closing the gap between students that had access to computational resources (and computational training) from an early age and students that did not. Late access to computational resources and training can occur due to lack of economic resources, often ocurring in households from underrepresented communities and minorites. It can also be due to gender-biased parental and community pressures, in which males are more often encouraged to perform activities related to computers, while females are discouraged.
How to balance software acceptance VS. adoption?
These principles can be used to aide not only reproducibility, but also software adoption in the natural sciences.
Discuss: why address accessibility and not other apects of reproducibility?

\section{Conclusion}
\label{sec:conc}

Making accessible reproducible workflows has several advantages:
save explanation/training time when analyses are run again by students and collaborators.
save research time for yourself when analyses are run again with more data, a different dataset, a different organism or biological model.
scientific efforts can build off of each other

\bigskip
\begin{center}
{\large\bf SUPPLEMENTARY MATERIAL}
\end{center}

\begin{description}

\item[Title:] Website and GitHub repository containing the complete teaching materials developed and demonstrated here.

\item[GitHub repository link:] \url{https://github.com/McTavishLab/R_OpenTree_tutorials}

\item[Website link:] \url{https://mctavishlab.github.io/R_OpenTree_tutorials}

\end{description}

\bibliographystyle{agsm}

\bibliography{Bibliography-MM-MC}

\end{document}

### 2021-09-16 jsdse/TAS-template.tex
jsdse submission

@@
 % NOTE: To produce blinded version, replace "0" with "1" below.
-\newcommand{\blind}{0}
+\newcommand{\blind}{1}
 
@@
 \begin{abstract}
-The text of your abstract.  200 or fewer words.
+Reproducibility is essential for scientific development. Efforts
+to increase computational reproducibility have focused on increasing availability
+of code and data. However, availability does not imply accessibility, and the latter
+is infrequently addressed, even if it is key to achieve full workflow automatization
+and reproducibility. Using an example in evolutionary biology, we identify factors that have specifically affected accessibility in
+the natural sciences, and ways researchers can address it to ensure reproducible
+and automatic workflows.
+The Open Tree of Life project (OpenTree) has developed a platform that facilitates
+availability of results from evolutionary biology research. However, baseline required
+computational knowledge and skills to
+access them are often not found among the target users. While documentation
+is available, it is often written using highly specialized language that is also inaccessible for the
+average target user.
+We present a set of principles to generate documentation that improves accessibility
+of code and documentation and present a tutorial were we apply those principles.
+
 \end{abstract}
@@
 
-Reproducibility --the extent to which consistent results are obtained when a scientific experiment is repeated (\cite{repdef2021})-- is a key aspect for the advancement of science, as it constitutes a minimum standard to understand scientific results, to determine their reliability and generality, and eventually to build more scientific knowledge upon those results (\cite{king1995replication, peng2011reproducible, powers2019open}).
-Reproducibility rates in the natural sciences are low (\cite{ioannidis2005most, prinz2011believe}), prompting concerns about a reproducibility crisis in the field (\cite{baker2016reproducibility}).
-The scientific community has united to incentivize cultural changes that will improve reproducibility rates long term, such as transparency, availability, and workflow automatization, to name a few (\cite{peng2015reproducibility}).
-We argue that accesibility is a key aspect that must accompany availability (Box 1) in order to achieve full workflow automatization and reproducibility.
+Reproducibility --the extent to which consistent results are obtained when a scientific
+experiment is repeated (\cite{repdef2021})-- is a key aspect for the advancement
+of science, as it constitutes a minimum standard to understand scientific results,
+to determine their reliability and generality, and eventually to build more scientific
+knowledge upon those results (\cite{king1995replication, peng2011reproducible, powers2019open}).
+Reproducibility rates in the natural sciences are low (\cite{ioannidis2005most, prinz2011believe}),
+prompting concerns about a reproducibility crisis in the field (\cite{baker2016reproducibility}).
+The scientific community has united to incentivize cultural changes that will improve
+reproducibility rates long term, such as transparency, availability, and workflow
+automatization, to name a few (\cite{peng2015reproducibility}).
+We argue that accesibility is a key aspect that must accompany availability (Box 1)
+in order to achieve full workflow automatization and reproducibility.
 
@@
 
-We focus on identifying factors that have specifically affected accessibility in the natural sciences, and ways researchers can address it to ensure reproducible and automatic workflows.
-We use an example in evolutionary biology.
-This example focuses on understanding the shared ancestry among organisms through evolutionary trees.
-These phylogenetic trees provide the basis to study and understand biological processes (\cite{dobzhansky1973nothing}).
-Hence, improving reproducibility and availability of phylogenetic research is relevant for many aspects of biological research.
-The Open Tree of Life project (OpenTree) has developed a platform that facilitates availability of results from phylogenetic research, by standardizing and storing phylogenetic data with the goal of synthesizing a single phylogenetic tree encompassing all life (\cite{opentreeoflife2019synth}).
-All data in OpenTree is open access and available programmatically through its many Application Programming Interface (API) services (\cite{opentreeAPIs}).
-Additionally, R packages (\cite{michonneau2016rotl}) and Python libraries (\cite{mctavish2021opentree}) have been developed as wrappers for OpenTree API services to make them available to a wider programming audience.
-The R and Python OpenTree wrappers have been utilized by computer-literate individuals, to seamlessly establish reproducible workflows to use and reuse expert phylogenetic knowledge for biological research (\cite{sanchez2019datelife}) and education (\cite{nguyen2020phylotastic, phylotasticedtools, galacticedtools}).
+We focus on identifying factors that have specifically affected accessibility in
+the natural sciences, and ways researchers can address it to ensure reproducible
+and automatic workflows.
+We use an example in evolutionary biology. This example focuses on understanding
+the shared ancestry among organisms through evolutionary trees. These
+trees provide the basis to study and understand biological processes
+(\cite{dobzhansky1973nothing}). Hence, improving reproducibility and availability
+of phylogenetic research is relevant for many aspects of biological research.
+The Open Tree of Life project (OpenTree) has developed a platform that facilitates
+availability of results from phylogenetic research, by standardizing and storing
+phylogenetic data with the goal of synthesizing a single phylogenetic tree encompassing
+all life (\cite{opentreeoflife2019synth}).
+All data in OpenTree is open access and available programmatically through its many
+Application Programming Interface (API) services (\cite{opentreeAPIs}).
+Additionally, R packages (\cite{michonneau2016rotl}) and Python libraries
+(\cite{mctavish2021opentree}) have been developed as wrappers for OpenTree API services
+to make them available to a wider programming audience.
+The R and Python OpenTree wrappers have been utilized by computer-literate individuals,
+to seamlessly establish reproducible workflows to use and reuse expert phylogenetic
+knowledge for biological research (\cite{sanchez2019datelife}) and education
+(\cite{nguyen2020phylotastic, phylotasticedtools, galacticedtools}).
 
@@
 However, many students are now being trained in R as a statistics and data analysis environment.
-In order for reproducible computational tools to be adopted for research, they much be accessible to reseachers.
+In order for reproducible computational tools to be adopted for research, they need to be much more accessible to reseachers.
 Data and code availability is a core requirement for reproducibility research (\cite{peng2011reproducible, sandve2013ten, powers2019open}).
@@
 We applied these principles to a series of tutorials and vignettes for the OpenTree project.
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=3in]{fig1.pdf}
-\end{center}
-\caption{Principle 1. \label{fig:first}}
-\end{figure}
 
@@
 
-We examined primary documentation from the OpenTree API R wrapper package `rotl`.
+We examined available primary documentation for the OpenTree API R wrapper (the package `rotl`),
+and designed a workflow that visits as many functions as possible, and demonstrate
+uses commonly requested by OpenTree users that are not demonstrated elsewhere.
+By framing the function workflow using highly requested uses, the documentation acquires a
+narrative arc that is easier to follow and remember by users. This facilitates the translation
+of functions to specfic use cases in biology.
 
-We designed a workflow that visits as many functions as possible, while answering to uses that are most commonly requested by users of OpenTree.
-By framing the use cases in highly requested tasks, the documenation follows a narrative conceptual arc, making the translation to biological use cases clear.
-
-> [name=Emily Jane McTavish]Brief para or figure on what the motivating example is here
+For the tutorial demonstrated here, we used the commonly requested use case of obtaining
+a phylogenetic tree for all lineages within a specific taxonomic rank.
 
@@
 
-Primary documentation focuses on demonstrating usage function with examples that work seamlessly without errors.
-We argue that the opposite is needed to support user adoption of reproducibile workflows: demonstrate examples that do not work as expected and exemplify ways to address them.
-We identify inputs that would give a wide range of warnings and errors, focusing on demonstrating these cases.
-This helps users to not be afraid of errors and warnings, but instead to use them to their advantage (CITE carpentries).
+Primary documentation focuses on demonstrating usage function with examples that
+work seamlessly, without errors. We argue that the opposite is needed to support
+user adoption of reproducibile workflows: demonstrate examples that do not work
+as expected and exemplify ways to address them (Figure \ref{fig:first}). We identify inputs that would give
+a wide range of warnings and errors, focusing on demonstrating these cases. This
+helps users to not be afraid of errors and warnings, but instead to use them to
+their advantage.
 We also identify effects of warnings and errors downstream of the workflow.
 
-We identify ways to evaluate inputs to know if they will produce an error, and design alternatives on what to do when faced with an error or warning, and demonstrate these alternatives.
-One of the most essential skills in programming is interpreting and moving forward from errors.
-Many finely honed tutorials do not trigger errors, which preculdes helping studnts to develop the tools to understand and address errors when the do encounter them, as they inevitably will!
+We identify ways to evaluate inputs to know if they will produce an error, and design
+alternatives on what to do when faced with an error or warning, and demonstrate
+these alternatives.
+One of the most essential skills in programming is interpreting and moving forward
+from errors.
+Many finely honed tutorials do not trigger errors, which precludes helping students
+to develop the tools to understand and address errors when they do encounter them,
+as they inevitably will.
 
-> [name=Emily Jane McTavish]Show where you deal with an error in your tutorial
+\begin{figure}
+\begin{center}
+\includegraphics[width=3in]{fig1.png}
+\end{center}
+\caption{Snapshot of a section of the tutorial website, where we demonstrate a common error. \label{fig:first}}
+\end{figure}
 
@@
 
-We focused on complementing the primary documentation by identifying computational concepts that were assumed or were not explained in depth.
-We vetted the tutorials with an audience on workshops as well as individual user
-
+Besides avoiding formal language, and incorporating elements of pop culture, such as picture
+character icons known as "emojis", to make the language used more familiar to the
+target audience (see Figure \ref{fig:first}), we made an effort to specifically
+complement the primary documentation by identifying
+computational concepts that were assumed or were not explained in depth.
+We vetted the tutorials with an audience on workshops as well as individual user.
 We choose examples that are charismatic for the audience.
-
-> [name=Emily Jane McTavish]Specify what one you used!
+For example, when we presented the tutorial for a team specialized in Amphibians,
+we tailored the examples using frogs and their allies.
 
@@
 
-We published the tutorials on a public, free license, free of cost, and free for use and reuse repository and persistent website.
-The tutorial is available for the users to go back to it any time they need it, and to be passed on to other users.
+We published the tutorials on a public, free license, free of cost, and free for
+use and reuse repository and persistent website (\cite{RopentreeTutorials, RopentreeTutorialsWebsite}).
+The tutorial is available for the users to go back to it any time they need it,
+and to be passed on to other users (Figure \ref{fig:second}).
 
-Following the carpentries [@CITE], we created a main version of the tutorial that is updated.
-Versions presented on workshops are a copy from the original repository, and represent a stable and temporal snapshot of the functions and workflows presented in the tutorial.
+\begin{figure}
+\begin{center}
+\includegraphics[width=3in]{fig2.png}
+\end{center}
+\caption{Snapshot of the home to our tutorial website, showing part of the schedule. \label{fig:second}}
+\end{figure}
 
-> [name=Emily Jane McTavish] Show the page, or the layout, or something :P
+Following the carpentries, we created a main version of the tutorial that
+is updated. Versions presented on workshops are a copy from the original repository,
+and represent a stable and temporal snapshot of the functions and workflows presented
+in the tutorial.
 
-\subsubsection*{5. The (now) classics of computational reproducibility}
-
-Provide all information on package version and system capabilities.
-
-> [name=Emily Jane McTavish]link to where/and how you did that
+%%% \subsubsection*{5. The (now) classics of computational reproducibility}
+%%%
+%%% Provide all information on package version and system capabilities.
+%%%
+%%% > [name=Emily Jane McTavish]link to where/and how you did that
 

### 2022-02-02 jsdse/Manuscript.tex
updating file structure and readme

@@
 %\bibliographystyle{natbib}
-%\bibliographystyle{bmc-mathphys}
-\bibliographystyle{agsm}
 
@@
 {
-  \title{\bf Approachable case studies support learning and reproducibility in data science: An example from evolutionary biology}
+  \title{\bf Promoting analysis reproducibility with accessibility: An example in evolutionary biology}
   \author{Luna L. Sanchez Reyes\thanks{
-    The authors gratefully acknowledge ``Sustaining the Open Tree of Life'',
-    NSF ABI No. 1759838, and ABI No. 1759846. They are also also deeply grateful towards
-    ``The Carpentries'' organization; without its invaluable volunteers and resources
-    this project could not have been possible}\hspace{.2cm}\\
+    The authors gratefully acknowledge "Sustaining the Open Tree of Life", NSF ABI No. 1759838, and ABI No. 1759846.}\hspace{.2cm}\\
     School of Natural Sciences, University of California, Merced\\
@@
   \begin{center}
-    {\LARGE\bf Approachable case studies support learning and reproducibility in data science: An example from evolutionary biology}
+    {\LARGE\bf Promoting analysis reproducibility with accessibility: An example in evolutionary biology}
 \end{center}
@@
 \begin{abstract}
-Research reproducibility is essential for scientific development. Yet, rates of reproducibility are low, especially in the natural sciences. 
-As increasingly more research relies on computers and software, efforts for improving reproducibility rates have focused on making available research workflows as computer code, as well as raw and processed data in computer readable form.
-However, research products that are digitally available are not necessarily friendly for learners and interested parties with little to no experience in the field. 
-This renders research products unapproachable, counteracts availability, and hinders reproducibility short and long term. 
-To improve long term adoption of reproducible workflows in research, they need to be made approachable for learners, the researchers of the future.
-
-Using a case study within evolutionary biology, we identify aspects of research workflows that make them unapproachable to the general audience: 
-use of highly specialized language; 
-unclear goals and high cognitive load; and 
-content-focused descriptions. 
-We propose a set of principles to improve the unapproachable aspects of research workflows, and illustrate their application in a tutorial. 
-We elaborate on the general application of these principles for documenting research workflows and products, to provide present learners and future researchers with tools for successful scientific reproducibility.
-%> ** add a sentence on implications for teaching reproducibility
-
-%% Projects that address availability of code and data in the natural sciences
-%% have focused on producing documentation that describes the nature and usage of the
-%% resources.
-%% However, it often communicates in a highly specialized language that is largely
-%% inaccessible for the average target user. Additionally, baseline
-%% computational knowledge and required skills to access scientific results are often
-%% not found among potential users, and keep on increasing.
+Reproducibility is essential for scientific development. Efforts
+to increase computational reproducibility have focused on increasing availability
+of code and data. However, availability does not imply accessibility, and the latter
+is infrequently addressed, even if it is key to achieve full workflow automatization
+and reproducibility. Using an example in evolutionary biology, we identify factors
+that have specifically affected accessibility in
+the natural sciences, and ways researchers can address this to ensure reproducible
+and automatic workflows.
+The Open Tree of Life project (OpenTree) has developed a platform that facilitates
+availability of results from evolutionary biology research. However, baseline
+computational knowledge and skills required to
+access scientific results are often not found among target users. While documentation
+is available, it is often written using highly specialized language that is also
+ inaccessible for the average target user.
+We present a set of principles to generate documentation that improves accessibility
+of code and documentation and present a tutorial were we apply those principles.
 
@@
 \noindent%
-{\it Keywords:}  open science, R programming language, phylogenetics, Open Tree of Life, pedagogy, phylogenetics, datelife
+{\it Keywords:}  open science, education, R, phylogenetics, tutorials
 \vfill
@@
 \spacingset{1.45} % DON'T change the spacing!
-\section*{Introduction}
+\section{Introduction}
 \label{sec:intro}
 
-Research reproducibility --the extent to which consistent results are obtained when
-a scientific experiment or research workflow is repeated \citep{repdef2021}-- is a key aspect of the advancement of science, as it constitutes a minimum standard that allows understanding research products, i.e., methods, data, analysis, results, etc.
-\citep{piwowar2013value}, to determine their reliability and generality, and eventually build up scientific knowledge and applications based on those products \citep{king1995replication, peng2011reproducible, powers2019open}.
-In the natural sciences, rates of reproducibility are low \citep{ioannidis2005most, prinz2011believe}, which has elicited concerns about a crisis in the field \citep{baker2016reproducibility}.
+Reproducibility --the extent to which consistent results are obtained when a scientific
+experiment is repeated (\cite{repdef2021})-- is a key aspect for the advancement
+of science, as it constitutes a minimum standard to understand scientific results,
+to determine their reliability and generality, and eventually to build more scientific
+knowledge upon those results (\cite{king1995replication, peng2011reproducible, powers2019open}).
+Reproducibility rates in the natural sciences are low (\cite{ioannidis2005most, prinz2011believe}),
+prompting concerns about a reproducibility crisis in the field (\cite{baker2016reproducibility}).
+The scientific community has united to incentivize cultural changes that will improve
+reproducibility rates long term, such as transparency, availability, and workflow
+automatization, to name a few (\cite{peng2015reproducibility}).
+We argue that accesibility is a key aspect that must accompany availability (Box 1)
+in order to achieve full workflow automatization and reproducibility.
 
-In response, the scientific community has been developing new principles and standards to incentivize cultural changes that support a long term improvement of reproducibility rates in the natural sciences \citep{peng2015reproducibility, wilkinson2016fair, miyakawa2020no}.
-A standard for reproducibility that has received much attention is availability, which we define as a property denoting that a research product can be reached (acquired, copied, analyzed, processed and/or reused) at no financial, legal or technical cost \citep{arnold2019turing}, and without geographic, demographic, social or temporal barriers for the population \citep{fecher2014open}.
+\bigskip
+\bigskip
 
-In this paper, we argue that research products that are digitally available are often unapproachable in practice, because they are not friendly for learners and interested parties with different levels of experience in the field.
-Research products that are unapproachable counteract availability, and hinder reproducibility short and long term.
-To support long term adoption of reproducible practices in the natural sciences, research workflows need to be made approachable for learners, the researchers of the future \citep{roland2002think}.
+\noindent\fbox{%
+    \parbox{\textwidth}{%
+    \textbf{Box 1.}
+      Availability does not imply accessibility. One example we really like is the
+      marshmallows in an office story.
+      The marshmallow bags were there, availabe for anyone in the office to eat
+      the marshmallows inside. But the marshmallows stayed there for days, weeks even.
+      And it was not until someone opened the bag and put the marshmallows in a tray,
+      that people started actually eating them. They were gone in a matter of hours.
+      Code is not marshmallows. But the point is that by
+      sharing your code and documenting it might not be enough for the general researcher
+      to reproduce results. Researchers will certainly be able to eventually figure it out,
+      but the time needed
+      to do so might not be worthy for them, or might not be something they can invest
+      in, even if it would be useful for them long term, the short term investment
+      is too intense.
+    }%
+}
+\bigskip
 
-To elaborate on our thesis, we designed a case study within the research field of phylogenetics, a discipline within evolutionary biology.
-We use our case study to identify barriers that have made research workflows largely unapproachable to a general audience in the natural sciences.
-Then, we propose some principles for researchers to address these barriers, and create research workflows that are reproducible by a larger audience.
-% ** describe/introduce reproducibility curriculum here
-The principles proposed here can be generalized and integrated into the undergraduate and graduate school STEM curriculum, either for courses specialized in reproducibility or within other subject areas, as a necessary component of successful and impactful science.
+We focus on identifying factors that have specifically affected accessibility in
+the natural sciences, and ways researchers can address it to ensure reproducible
+and automatic workflows.
+We use an example in evolutionary biology. This example focuses on understanding
+the shared ancestry among organisms through evolutionary trees. These
+trees provide the basis to study and understand biological processes
+(\cite{dobzhansky1973nothing}). Hence, improving reproducibility and availability
+of phylogenetic research is relevant for many aspects of biological research.
+The Open Tree of Life project (OpenTree) has developed a platform that facilitates
+availability of results from phylogenetic research, by standardizing and storing
+phylogenetic data with the goal of synthesizing a single phylogenetic tree encompassing
+all life (\cite{opentreeoflife2019synth}).
+All data in OpenTree is open access and available programmatically through its many
+Application Programming Interface (API) services (\cite{opentreeAPIs}).
+Additionally, R packages (\cite{michonneau2016rotl}) and Python libraries
+(\cite{mctavish2021opentree}) have been developed as wrappers for OpenTree API services
+to make them available to a wider programming audience.
+The R and Python OpenTree wrappers have been utilized by computer-literate individuals,
+to seamlessly establish reproducible workflows to use and reuse expert phylogenetic
+knowledge for biological research (\cite{sanchez2019datelife}) and education
+(\cite{nguyen2020phylotastic, phylotasticedtools, galacticedtools}).
 
-%% \bigskip
-%% \bigskip
-%%
-%% \noindent\fbox{%
-%%     \parbox{\textwidth}{%
-%%     \textbf{Box 1. The role of approachable research workflows in teaching reproducibility.}
-%%       % Availability is defined as the fact or possibility that something can be bought,
-%%       % used, or reached by someone \citep{availability2022cambridge}.
-%%       % Accessibility is defined as "the fact of being able to be reached or obtained easily"
-%%       % \citep{accessibility2022cambridge}
-%%       % Yet, in practice, availability does not imply accessibility.
-%%       Availability and accessibility are conceptually accepted as synonyms for ``the fact
-%%       of being able to be reached" \citep{available-accessible2022cambridge}.
-%%       However, accessibility's definition goes further and describes it as the
-%%       ``fact of being able to be reached \textbf{easily}" \citep{accessibility2022cambridge}.
-%%       A secondary definition of accessibility elaborates more on this: ``the quality
-%%       of being \textbf{easy to understand}''.
-%%
-%%       What we realize from these definitions is that, in practice, something might
-%%       be available but not necessarily accessible,
-%%       hindering the realization of availability.
-%%       To illustrate the practical difference between availability and accessibility we
-%%       will use an allegory that goes like this:
-%%
-%%       ** Change marshmallow story to cake
-%%
-%%       ** Explain different types of barriers to accessibility within the example, then within science
-%%
-%%       ** Explain why we focus on psychological/social/pedagogical barriers
-%%
-%%       ** Gauging accessibility of a resource is tricky because it depends on an
-%%       individual experience of what is in effect \textbf{easy} for someone.
-%%       \textbf{Easy} is a subjective quality that fully depends on individual perceptions
-%%       resulting from individual contexts.
-%%       As a society we are moving towards validating and welcoming all human experiences.
-%%       To improve accessibility levels in the natural sciences, it
-%%       is important to account for the factors that are creating different experiences
-%%       of what is \textbf{easy}.
-%%
-%%       ** Now, explain different types of barriers to accessibility in science
-%%       (Jessica: Maybe say that lack of accessibility can be due to a number of
-%%       factors (physical, resources, economic), but one critical factor is the use
-%%       of impenetrable expert language.)
-%%     }%
-%% }
-%% \bigskip
+The OpenTree project demonstrates that efforts to increase reproducibility and availability have also increased the baseline required computational knowledge and skills in phylogenetic research.
+These computational skills requirements are likely increasing across all natural sciences.
+Computing is not traditionally a core skill taught to biologists and naturalists.
+However, many students are now being trained in R as a statistics and data analysis environment.
+In order for reproducible computational tools to be adopted for research, they need to be much more accessible to reseachers.
+Data and code availability is a core requirement for reproducibility research (\cite{peng2011reproducible, sandve2013ten, powers2019open}).
+However, the utility of data resources is limited by the technical challenges of accessing the data.
+In order to motivate reproducible research, that gap needs to be bridged.
+In this work we focus on improving accessibility of code examples and documentation.
+In particular, we identify the necessity for documentation that is written down using language that is common to the target audience to facilitate examination, application, and adoption of code by the wider audience.
 
-\section*{A case study from phylogenetics}
-\label{sec:case}
+We present a set of principles to generate documentation that improves accessibility of code and documentation.
+We applied these principles to a series of tutorials and vignettes for the OpenTree project.
 
-Phylogenetics is a key discipline within evolutionary biology \citep{dobzhansky1973nothing}.
-It focuses on investigating the history of shared ancestry of living and extinct organisms using biological data, and represents this evolutionary history with a diagram known as a phylogeny or phylogenetic tree (because it grows through time and appears to have branches; Figure \ref{fig:tree}).
-Phylogenies provide the basis to study and understand all biological processes in an evolutionary context \citep{dobzhansky1973nothing}.
-Hence, it follows that improving reproducibility rates in phylogenetics has the potential to positively impact research across the natural sciences.
+\section{Methods}
+\label{sec:meth}
+\subsection*{Identifying hurdles to accesibility}
+
+Good primary documentation for code describes general usage of individual functions, the components and variables a function can take, and it should be accompanied with function usage examples on how to apply it.
+As opposed to code, primary documentation is written in natural language (i.e., any known human language, e.g., English, Spanish, Chinese) and usually makes use of highly specialized computational jargon (computationally specific concepts, words, and phrases) as well as formal language, which often slows down or even obstructs examination, application, and adoption of code by external individuals.
+Because primary documentation is considered a professional document, acceptance of the research by the scientific community could be reduced if a more informal language is used.
+Secondary types of documentation, such as vignettes and tutorials, demonstrate additional cases of individual function usage, and describe analysis workflows in more detail, as well as function associations to generate a specific analysis and results.
+While secondary documentation has become more common practice, it is still often generated using highly specialized language.
+
+\subsection*{Adressing hurdles to accesibility: the principles}
+
+\subsubsection*{1. Demonstrate integration of function usage with motivating examples}
+
+We examined available primary documentation for the OpenTree API R wrapper (the package `rotl`),
+and designed a workflow that visits as many functions as possible, and demonstrate
+uses commonly requested by OpenTree users that are not demonstrated elsewhere.
+By framing the function workflow using highly requested uses, the documentation acquires a
+narrative arc that is easier to follow and remember by users. This facilitates the translation
+of functions to specfic use cases in biology.
+
+For the tutorial demonstrated here, we used the commonly requested use case of obtaining
+a phylogenetic tree for all lineages within a specific taxonomic rank.
+
+\subsubsection*{2. Demonstrate errors and warnings thoroughly}
+
+Primary documentation focuses on demonstrating usage function with examples that
+work seamlessly, without errors. We argue that the opposite is needed to support
+user adoption of reproducibile workflows: demonstrate examples that do not work
+as expected and exemplify ways to address them (Figure \ref{fig:first}). We identify inputs that would give
+a wide range of warnings and errors, focusing on demonstrating these cases. This
+helps users to not be afraid of errors and warnings, but instead to use them to
+their advantage.
+We also identify effects of warnings and errors downstream of the workflow.
+
+We identify ways to evaluate inputs to know if they will produce an error, and design
+alternatives on what to do when faced with an error or warning, and demonstrate
+these alternatives.
+One of the most essential skills in programming is interpreting and moving forward
+from errors.
+Many finely honed tutorials do not trigger errors, which precludes helping students
+to develop the tools to understand and address errors when they do encounter them,
+as they inevitably will.
 
@@
 \begin{center}
-\includegraphics[width=3in]{fig-tree.png}
+\includegraphics[width=3in]{fig1.png}
 \end{center}
-\caption{A phylogenetic tree from our tutorial.
-It was extracted using OpenTree of Life resources \citep{opentreeoflife2019synth} wrapped in the \texttt{rotl} R package \citep{michonneau2016rotl}.
-\label{fig:tree}}
+\caption{Snapshot of a section of the tutorial website, where we demonstrate a common error. \label{fig:first}}
 \end{figure}
 
-To explore barriers to approachable phylogenetics, we develop a case study that touches on three common problems within the field: standardizing organism names in phylogenies, obtaining current phylogenetic knowledge for a group of organisms, and summarizing this phylogenetic knowledge in a meaningful way.
-To address these problems, we propose a research workflow that relies on resources from the Open Tree of Life (OpenTree), an open source project that provides digital availability of phylogenetic results from published, peer-reviewed research, which is considered as vetted and state-of-the-art knowledge in the field.
-OpenTree phylogenies are stored in a public database, the Phylesystem \citep{mctavish2015phylesystem}, and are downloadable as various computer-readable file types, which is key for reusability and reproducible workflows \citep{wilson2017good}.
-OpenTree also provides access to a single standard for organism names (taxonomic standard) that is applied to the stored phylogenies \citep{rees2017automated}, which are then used to summarize a single phylogenetic tree encompassing all life \citep{opentreeoflife2019synth}.
+\subsubsection*{3. Avoid jargon and expert language}
 
-All of these resources 
-%% (the stored phylogenies, the taxonomic standard, the synthetic tree, and all
-%% tools developed to process and summarize stored phylogenies),
-are available for download and use from OpenTree free of cost to any user.
-One way to access OpenTree resources is through its Graphical User Interface (GUI; aka, a website or application that allows users to access and use functionalities with mouse or keyboard clicks).
-However, reducing as many manual steps as possible in research workflows is key for reproducibility, as manual data manipulation scales poorly and is prone to error \citep{bakken2019journey}.
-OpenTree's resources are also programmatically available through its Application Programming Interface services (APIs; aka, computer code that implements computer functionalities, in this case OpenTree resources, that can be used by programmers
-to build more functionalities), which provide scalability and reproducibility \citep{opentreeAPIv3}.
-However, this comes at a high cost for the user, which requires considerably more computer programming experience and literacy to be able to successfully use APIs.
-%% In recent years, OpenTree API services have been wrapped into software packages for R and Python
-%% \citep{michonneau2016rotl, mctavish2021opentree}, open source, free of cost
-%% programming languages that represent two of the most widely used programming languages
-%% in the sciences \citep{baker2017scientific}. These packages should contribute to making
-%% OpenTree resources more accessible to a wider user audience.
-The \texttt{rotl} R package \citep{michonneau2016rotl} and the \texttt{opentree} Python module \citep{mctavish2021opentree} have been developed to wrap OpenTree's API services
-and make them available through the R and Python programming languages.
-R and Python are open source and free of cost, and represent two of the most widely used programming languages in the sciences today \citep{eglen2009quick, baker2017scientific}.
-As such, \texttt{rotl} and \texttt{opentree} software packages should contribute to making OpenTree resources more accessible to a wider programming user audience.
+Besides avoiding formal language, and incorporating elements of pop culture, such as picture
+character icons known as "emojis", to make the language used more familiar to the
+target audience (see Figure \ref{fig:first}), we made an effort to specifically
+complement the primary documentation by identifying
+computational concepts that were assumed or were not explained in depth.
+We vetted the tutorials with an audience on workshops as well as individual user.
+We choose examples that are charismatic for the audience.
+For example, when we presented the tutorial for a team specialized in Amphibians,
+we tailored the examples using frogs and their allies.
 
-However, while learners in the natural sciences have been engaging independently with R and Python programming languages, computer programming is not traditionally a core skill formally taught to biologists and naturalists \citep{sayres2018bioinformatics, wright2019the, williams2019barriers}.
-As computers continue to play a larger role in most scientific disciplines \citep{piccolo2016tools}, higher baseline computational skills are required across all natural sciences not only to develop an original research workflow, but to be able to follow and reproduce research workflows from other researchers.
+\subsubsection*{4. Make it stable through time}
 
-Thus, efforts to increase reproducibility rates long term in the natural sciences would benefit from addressing specific barriers for learners in the field, to support them in acquiring the skills needed to reproduce research workflows that rely heavily on computer code \citep{peng2011reproducible, sandve2013ten, powers2019open}.
-
-In the next section, we describe (in no particular order) the barriers to approachable research workflows that we identified on our case study.
-Then we develop a set of principles to address these barriers, and apply the latter to a set of teaching materials that are available at https://mctavishlab.github.io/R\_OpenTree\_tutorials/.
-
-\section*{Identifying barriers to approachable research workflows}
-\label{sec:identifying}
-
-% Cite carpentries and Mosi
-
-The main goal of our case study is to obtain a single phylogeny summarizing data from a set of published phylogenies for the canids (the family of dogs, coyotes, wolves, etc.), our organism of study.
-All analysis for our case study can be completely accomplished using functions from the R package \texttt{rotl} or the Python module \texttt{opentree}. 
-If a researcher were to use the proposed analysis workflow in a publication, they would typically describe it in the methodology section as ``The canid summary phylogeny was obtained using functions from X package, details are available as supplementary files".
-This is usual practice, mainly because journals do not have space to publish all code used for an analysis in the methods section.
-Yet, supplementary files and data have the misfortune to not be peer-reviewed as thoroughly (or at all) as the main manuscript \citep{pop2015use}.
-They are also prone to the dreaded promise ``available upon request", which has very low rates of fulfillment \citep{krawczyk2012available}.
-Without the primary data and code that was used to perform an analysis it is impossible to reproduce said analysis \citep{miyakawa2020no}.
-When the code is available, other issues can complicate reproduction of the analysis, to the point of completely obstructing reproducibility.
-For example, some questions that are often unaddressed in research workflows are: what software do I need to open the code files?
-Can I run the code with the same software I used to open the files, or do I need a different one?
-Do I need additional software that the analysis depends on?
-What does the code even mean?
-
-These questions can usually be answered by referring to the software documentation, which are usually publicly available and can be accessed by any potential user.
-As opposed to code, software documentation is written in natural language (i.e., any known human language, e.g., English, Spanish, Chinese), and is considered a key element for successful adoption of software by target users \citep{karimzadeh2018top}.
-This might explain why documentation for software addressed to academic users is also usually written using highly specialized computational language or jargon (i.e., computationally specific concepts, words, and phrases) as well as formal scientific and academic language.
-We identify this as \textit{barrier 1} to approachable research workflows.-- \textit{Specialized language is intimidating}.
-While scientific jargon might have an important role for formal acceptance of software by the scientific and academic community, it can be perceived as cold and/or intimidating language, that often slows down or even obstructs examination, application, and adoption of code by a wider audience \citep{ball2017its}.
-In contrast, introducing information without the use of jargon, supports learners' conceptual understanding of new ideas and concepts \citep{mcdonnell2016concepts, pan2019online}.
-% jargon creates a hostile environment that does not foster learning \citep{CITATION}.
-
-Another element of good software documentation is that it has to be thorough \citep{karimzadeh2018top}.
-Meaning that it should describe general usage of individual functions, as well as arguments and variables that said function can take \citep{karimzadeh2018top}.
-Individual documentation for each function is usually presented in alphabetic order, and does not have a specific analysis goal.
-Moreover, most software has numerous functions, so documentation is usually very lengthy and it is hard to navigate.
-This can have the effect of increasing the amount of information that needs to be simultaneously processed by the users, which can lead to overload of the finite amount of working memory any one possesses, known as cognitive load \citep{sweller1988cognitive}.
-In this context, identifying connections across functions that are meant to work on the same analysis workflow can become a very difficult task.
-We recognize this as \textit{barrier 2.-- Lack of specific goals leading to high cognitive load}.
-High cognitive load is know to have a negative effect in learning software \citep{chandler1996cognitive, van2005research, lambert2009student}.
-
-A third important aspect of software documentation are examples that demonstrate usage of individual functions \citep{karimzadeh2018top}.
-Examples presented in software documentation are usually worked to perfection, as they are intended to showcase the ideal or minimal case in which a function works well.
-Perfectly worked examples ignore the user experience by maintaining focus on the software content, and fail to provide users with expert and clear advice on how to troubleshoot if needed.
-We identify this as \textit{barrier 3.-- Content-focused examples}.
-Error management training is an approach that focuses on framing mistakes as beneficial to learning complex tasks, to give learners the opportunity to
-actively explore a task with a positive mindset \citep{frese1995error}.
-Providing examples that showcase potential errors supports users performance \citep{steele2014error}, and can greatly improve learners ability to troubleshoot outside the classroom \citep{shannon2015live, nederbragt2020ten}.
-
-%% Last but not least, software documentation is usually explained verbally in a document, very
-%% rarely is it approached using diagrams or other allegories.
-%% We recognize this as \textit{barrier 4.-- Little to no diversity of representation of information}.
-
-In sum, best practices for good software documentation are not enough to promote reproducibility of published research workflows that rely heavily on code. 
-In the following section we describe some principles that can help to reduce or remove the identified barriers, to create research workflows that are more approachable and hence more reproducible by a larger audience.
-
-\section*{Principles for approachable research workflows}
-\label{sec:addressing}
-
-\subsubsection*{Principle 1. Use friendly, relatable and respectful language}
-
-% avoid specialized, intimidating and cold language
-
-Besides avoiding formal language, and incorporating elements of pop culture, such as picture character icons known as ``emojis``, to make the language more familiar to a broader target audience (Figure \ref{fig:error}), we made an effort to specifically complement the primary documentation by identifying computational concepts that were assumed or were not explained in depth.
-We vetted the tutorials through feedback from workshop participants as well as individual users to identify such specialized concepts.
+We published the tutorials on a public, free license, free of cost, and free for
+use and reuse repository and persistent website (\cite{RopentreeTutorials, RopentreeTutorialsWebsite}).
+The tutorial is available for the users to go back to it any time they need it,
+and to be passed on to other users (Figure \ref{fig:second}).
 
@@
 \begin{center}
-\includegraphics[width=3in]{fig-error.png}
+\includegraphics[width=3in]{fig2.png}
 \end{center}
-\caption{Snapshot of a section of the tutorial website, where we demonstrate a common error. \label{fig:error}}
+\caption{Snapshot of the home to our tutorial website, showing part of the schedule. \label{fig:second}}
 \end{figure}
 
-\subsubsection*{Principle 2. Reduce cognitive load by providing specific and clear goals with literate programming}
+Following the carpentries, we created a main version of the tutorial that
+is updated. Versions presented on workshops are a copy from the original repository,
+and represent a stable and temporal snapshot of the functions and workflows presented
+in the tutorial.
 
-Cognitive load can be greatly reduced for learners by applying an active learning strategy such as linking usage to a ``real world'' or ``human'' application \citep{felder2009active}.
-Programming computer languages are by themselves quite abstract and represent a learning subject with a potentially high cognitive load for most learners.
-Pedagogical research shows that active learning practices are one of the most effective ways to take on abstract subjects \citep{freeman2014active}.
-A story-like narrative that links code usage in an integrative example, invites learners to try the code, which can lead them to remember what they are doing and why they are doing it.
-This ``literate programming'' paradigm \citep{knuth1984literate, fritzson2002mathmodelica} makes code more approachable, as it integrates narratives with computer code in the same document, supporting learners in actively following the code usage, supporting memory and understanding \citep{piccolo2016tools}.
+%%% \subsubsection*{5. The (now) classics of computational reproducibility}
+%%%
+%%% Provide all information on package version and system capabilities.
+%%%
+%%% > [name=Emily Jane McTavish]link to where/and how you did that
 
-We propose that documents developed with ``literate programming'' can be made more accessible by choosing narratives that are relatable to a more general audience.
-An easy way to do this in biology is choosing a charismatic taxon as a model organism.
-For a research group this can be the group they are studying. For the general audience, a highly charismatic group such as dinosaurs should work well.
-For example, when we presented our tutorial to the Amphibia Web Organization \citep{van2002amphibiaweb} in January 2020, we tailored all examples to frogs and their allies.
-% \citep{cite-aw-workshop}
+\section{Results and Discussion}
+\label{sec:results}
 
-We examined available software documentation for the package \texttt{rotl}, and designed a narrative that requires the usage of as many functions as possible.
-We demonstrate code applications that are commonly requested by OpenTree users, but that are not demonstrated in the documentation of the package.
-By framing the function workflow using highly requested uses, the documentation acquires a narrative arc that is easier to follow and remember by users.
-This can also facilitate translating the code application to other use cases of interest for learners in biology.
+We explain the warnings and errors and design ways to avoid them, and detect them beforehand (i.e., before using an input that would give an error). We explain the consequences of warnings.
+We designed ways to access the different elements of the outputs.
+We have received emails from senior researchers thanking us for this materials, and students have been able to engage using the packages with less hep from the PIs.
 
-%% Maybe Figure!! In the tutorial showcased here, we used the commonly requested
-%% use case of obtaining
-%% a phylogenetic tree for all lineages within a specific taxonomic rank.
+The principles to create tutorials described here facilitate adoption of software and analysis wokflows among researchers at different academic levels, from undergrads to established researchers.
+It will also help closing the gap between students that had access to computational resources (and computational training) from an early age and students that did not.
+Late access to computational resources and training can occur due to lack of economic resources, often ocurring in households from underrepresented communities and minorites.
+It can also be due to gender-biased parental and community pressures, in which males are more often encouraged to perform activities related to computers, while females are discouraged.
+How to balance software acceptance VS. adoption?
+These principles can be used to aide not only reproducibility, but also software adoption in the natural sciences.
+Discuss: why address accessibility and not other apects of reproducibility?
 
-\subsubsection*{Principle 3. Provide examples that are user-focused by demonstrating errors and warnings}
+\section{Conclusion}
+\label{sec:conc}
 
-% approachable practice: user focused examples (instead of content focus)
-% demonstrate errors and warnings thoroughly
-
-An activity that has become increasingly widespread in programming-language education is live programming.
-During live programming, and instructor writes code and executes it in a way that is visible to all learners through a screen \citep{guzdial2013lure, selvaraj2021live}.
-One benefit of this practice is that typos and mistakes occur, normalizing them for learners. 
-Watching an instructor handling errors, demonstrates learners on how to solve them when they are outside theclassroom \citep{shannon2015live, nederbragt2020ten}.
-When co-author McTavish was a postdoc, teaching an introductory programming workshop as a volunteer with the Carpentries (a non-profit group that teaches foundational coding and data science skills to researchers worldwide) \citep{wilson2006swc, SWCwebsite},
-a senior faculty member taking the workshop complained that the typos were slowing things down and interfering with the pedagogy.
-McTavish replied "the typos ARE the pedagogy".
-This has become a slogan of sorts at the Carpentries, capturing the idea that embracing and discussing mistakes is essential to teaching programming \citep{wilson2019teaching}.
-Yet, working through mistakes is rarely done on written pedagogical materials.
-Software documentation focuses on demonstrating usage function with examples that work seamlessly, without errors. 
-We argue that the opposite is needed to support adoption of reproducible workflows and support long term independence in learners and users performance \citep{gaspar2007restoring, steele2014error}.
-
-In our tutorials, we apply this principle by demonstrating examples that do not work as expected, and exemplifying ways to address them (Figure \ref{fig:error}).
-For example, we identified inputs that would give a wide range of warnings and errors.
-We then focus on providing explanations for these errors and warning messages.
-We believe this supports users and learners to be less intimidated by the messages, and to practice taking useful information out of them.
-% This is probably part of the computer science curriculum, but it is not at all
-% part of the natural sciences curriculum.
-
-We also demonstrate ways to evaluate inputs to determine if they will trigger an error or warning, and design and demonstrate alternative analysis routes on what to do when faced with an error or warning.
-One of the most essential skills in programming is interpreting and moving forward from errors.
-This has two pedagogical benefits.
-First, it provides users and learners with the means to troubleshoot their own warnings and errors.
-Second, it allows them to understand with more depth what the function is doing.
-
-% \subsubsection*{Principle 4. Demonstrate one thing in different ways}
-
-% - and little to no diversity of representation of information (say the same thing with different images, metaphors, stories)
-
-% We designed ways to access the different elements of the outputs.
-
-\section*{Conclusion}
-\label{sec:conclusion}
-
-Response from the community has been invaluable in gauging success of our teaching materials.
-Senior researchers often comment on the usefulness of the tutorials for their research, as well as how they have supported students in using the demosntrated R packages more independently.
-
-When developing our tutorials, we not only applied the principles elaborated here,
-but we followed basic recommendations for successful reproducibility \citep{sandve2013ten}.
-For example, we published the tutorials on a persistent and public website \citep{RopentreeTutorials} that adheres to the four r's of openness, by being free license, free of cost, and thus free for use, reuse, redistribute, revise and remix \citep{hilton2010four}.
-To make the website persistent, any updates to the tutorial are published as new versions.
-Versions presented at workshops are a copy from the original repository, and constitute a temporally stable snapshot of functions and workflows presented during a live workshop \citep{wilson2006swc, SWCwebsite}.
-This ensures that the tutorials are available for the users to return to any time they need it, and to be shared with other users and learners (Figure \ref{fig:schedule}).
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=3in]{fig-schedule.png}
-\end{center}
-\caption{Snapshot of the home to our tutorial website, showing part of the schedule.
- Our tutorial website was constructed using The Software Carpentries workshop template \citep{swc-workshop-template}. \label{fig:schedule}}
-\end{figure}
-
-We note that making approachable research workflows has all the advantages of reproducible research workflows.
-It saves time during explanation and training, when analyses are run by new collaborators and students.
-It can also save research time for yourself, when analyses are run again with more data, a different dataset, a different organism or biological model.
-It contributes to scientific efforts that can build off of each other.
-
-% \citep{roland2002think}
-Some universities and academic groups have started to incorporate reproducibility as a subject into their curriculum
-\citep{uwlibraries2022, nigms2022, psyteachr2022}.
-The focus of these resources has been for students to acquire and practice skills to document their work.
-The principles identified and outlined here can be used to set learning goals and outcomes for new reproducibility syllabi.
-Ultimately, the long term improvement of reproducibility rates in science will depend on our ability to intentionally integrate best practices for achieving reproducibility into the educational framework of future data scientists \citep{nasem2018data}.
-Inclusion of reproducibility into the data acumen of undergraduate curriculum will provide college learners and future researchers with the tools to develop the fundamental skills needed to successfully create reproducible scientific workflows and research products.
-
-The principles to create tutorials described here facilitate adoption of software and analysis workflows among researchers at different academic levels, from undergrads to established researchers.
-It will also help close the gap between students that had access to computational resources (and computational training) from an early age and students that did not.
-Late access to computational resources and training can occur due to lack of economic resources, often occurring in households from underrepresented communities and minorities \citep{google2016diversity, warner2021quantifying}.
-It can also be due to gender-biased parental and community pressures, in which male identifying individuals are more often encouraged to perform activities related to computers, while female identifying individuals are discouraged \citep{warner2021quantifying}.
-These principles can be used to improve not only reproducibility practices, but also software adoption in the natural sciences.
+Making accessible reproducible workflows has several advantages:
+save explanation/training time when analyses are run again by students and collaborators.
+save research time for yourself when analyses are run again with more data, a different dataset, a different organism or biological model.
+scientific efforts can build off of each other
 
@@
 
+\bibliographystyle{agsm}
+
 \bibliography{Manuscript-bibliography}

### 2022-02-03 jsdse/Manuscript.tex
updating ms

@@
 Reproducibility is essential for scientific development. Efforts
-to increase computational reproducibility have focused on increasing availability
+to increase scientific reproducibility have focused on increasing availability
 of code and data. However, availability does not imply accessibility, and the latter
@@
  inaccessible for the average target user.
-We present a set of principles to generate documentation that improves accessibility
-of code and documentation and present a tutorial were we apply those principles.
+We identified a set of principles to improve accessibility of OpenTree scientific data,
+implement them in a tutorial, and elaborate on their general application to code and
+documentation of scientific workflows.
 
@@
 Reproducibility --the extent to which consistent results are obtained when a scientific
-experiment is repeated (\cite{repdef2021})-- is a key aspect for the advancement
-of science, as it constitutes a minimum standard to understand scientific results,
-to determine their reliability and generality, and eventually to build more scientific
-knowledge upon those results (\cite{king1995replication, peng2011reproducible, powers2019open}).
-Reproducibility rates in the natural sciences are low (\cite{ioannidis2005most, prinz2011believe}),
+research experiment is repeated (\cite{repdef2021})-- is a key aspect for the advancement
+of science, as it constitutes a minimum standard that allows understanding scientific products
+(methods, analysis, results),
+to determine their reliability and generality, and eventually build up scientific
+knowledge and applications based on those products
+(\cite{king1995replication, peng2011reproducible, powers2019open}).
+In the natural sciences, rates of reproducibility are low (\cite{ioannidis2005most, prinz2011believe}),
 prompting concerns about a reproducibility crisis in the field (\cite{baker2016reproducibility}).
-The scientific community has united to incentivize cultural changes that will improve
-reproducibility rates long term, such as transparency, availability, and workflow
-automatization, to name a few (\cite{peng2015reproducibility}).
-We argue that accesibility is a key aspect that must accompany availability (Box 1)
-in order to achieve full workflow automatization and reproducibility.
+%%% The scientific community has united to incentivize cultural changes that will improve
+%%% reproducibility rates long term, such as transparency, availability, and workflow
+%%% automatization, to name a few (\cite{peng2015reproducibility}).
+In response, the scientific community has been generating standards and incentivizing
+cultural changes in an effort to improve reproducibility rates long term
+(\cite{peng2015reproducibility, wilkinson2016fair}).
+One standard that has received much attention is scientific availability, which
+ we define as the property of research objects to be acquired, copied, analyzed,
+  processed and/or reused, at no financial, legal or technical cost (\cite{arnold2019turing}),
+   and with no geographic, demographic or temporal barriers (\cite{fecher2014open}).
+In this paper, we argue that availability can not be fully achieved without accessibility (Box 1)
+We identify factors that have particularly affected accessibility in
+the natural sciences, and ways researchers can address them to ensure reproducible
+and automatic scientific workflows.
+% full workflow automatization and reproducibility in the natural sciences
 
@@
     \textbf{Box 1.}
-      Availability does not imply accessibility. One example we really like is the
+      Availability does not imply accessibility. One example we find really useful is the
       marshmallows in an office story.
-      The marshmallow bags were there, availabe for anyone in the office to eat
-      the marshmallows inside. But the marshmallows stayed there for days, weeks even.
-      And it was not until someone opened the bag and put the marshmallows in a tray,
+      Brand new marshmallow bags remained in a common office area after a social gathering.
+      The marshmallows did not belong to anyone anymore, so they were freely
+      available for eating by anyone in the office.
+      Yet, the marshmallow bags stayed in the common area, unopened for days and even weeks.
+      It was not until someone opened the bags and placed the marshmallows on a tray,
       that people started actually eating them. They were gone in a matter of hours.
-      Code is not marshmallows. But the point is that by
-      sharing your code and documenting it might not be enough for the general researcher
-      to reproduce results. Researchers will certainly be able to eventually figure it out,
-      but the time needed
-      to do so might not be worthy for them, or might not be something they can invest
-      in, even if it would be useful for them long term, the short term investment
-      is too intense.
+      Scientific products are not sweet as marshmallows, but they are a coveted asset.
+      While sharing, describing and documenting makes scientific products available,
+      it does not necessarily makes them accessible. For example, documentation of an
+      available scientific workflow might be written using expert language that is too foreign
+      for a general audience to successfully reproduce the workflow. While some individuals might
+      be able to invest the time to learn the expert language to successfully reproduce
+      the workflow, this will not be the case for the majority, even if it is
+      something that would be useful long term, the short term time investment
+      is often too high for individuals to engage with it.
     }%
@@
 
-We focus on identifying factors that have specifically affected accessibility in
-the natural sciences, and ways researchers can address it to ensure reproducible
-and automatic workflows.
-We use an example in evolutionary biology. This example focuses on understanding
-the shared ancestry among organisms through evolutionary trees. These
-trees provide the basis to study and understand biological processes
-(\cite{dobzhansky1973nothing}). Hence, improving reproducibility and availability
+To elaborate on our thesis, we use an example in evolutionary biology that focuses
+on understanding the shared ancestry among organisms through phylogenetic trees, which provide the
+basis to study and understand biological processes in an evolutionary context (\cite{dobzhansky1973nothing}).
+As such, improving reproducibility and availability
 of phylogenetic research is relevant for many aspects of biological research.
@@
 
-The OpenTree project demonstrates that efforts to increase reproducibility and availability have also increased the baseline required computational knowledge and skills in phylogenetic research.
-These computational skills requirements are likely increasing across all natural sciences.
+The OpenTree project demonstrates that efforts to increase availability for reproducibility
+in phylogenetics have also increased baseline computational knowledge and skills
+required in the field.
+Baseline computational skill requirements are likely increasing across all natural sciences.
 Computing is not traditionally a core skill taught to biologists and naturalists.
 However, many students are now being trained in R as a statistics and data analysis environment.
-In order for reproducible computational tools to be adopted for research, they need to be much more accessible to reseachers.
-Data and code availability is a core requirement for reproducibility research (\cite{peng2011reproducible, sandve2013ten, powers2019open}).
+In order for reproducible computational tools to be adopted for research, they
+ need to be much more accessible for present and future researchers.
+Data and code availability is a core requirement for reproducibility research
+ (\cite{peng2011reproducible, sandve2013ten, powers2019open}).
 However, the utility of data resources is limited by the technical challenges of accessing the data.
@@
 In this work we focus on improving accessibility of code examples and documentation.
-In particular, we identify the necessity for documentation that is written down using language that is common to the target audience to facilitate examination, application, and adoption of code by the wider audience.
+In particular, we identify the necessity for documentation that is written down
+ using language that is common to the target audience to facilitate examination,
+  application, and adoption of code by the wider audience.
 
-We present a set of principles to generate documentation that improves accessibility of code and documentation.
-We applied these principles to a series of tutorials and vignettes for the OpenTree project.
+We present a set of principles to generate documentation that improves accessibility
+ of code and documentation. We applied these principles to a series of tutorials
+  and vignettes developed for the OpenTree project.
 
@@
 \label{sec:meth}
-\subsection*{Identifying hurdles to accesibility}
+\subsection*{Identifying hurdles to accessibility}
 
-Good primary documentation for code describes general usage of individual functions, the components and variables a function can take, and it should be accompanied with function usage examples on how to apply it.
-As opposed to code, primary documentation is written in natural language (i.e., any known human language, e.g., English, Spanish, Chinese) and usually makes use of highly specialized computational jargon (computationally specific concepts, words, and phrases) as well as formal language, which often slows down or even obstructs examination, application, and adoption of code by external individuals.
-Because primary documentation is considered a professional document, acceptance of the research by the scientific community could be reduced if a more informal language is used.
-Secondary types of documentation, such as vignettes and tutorials, demonstrate additional cases of individual function usage, and describe analysis workflows in more detail, as well as function associations to generate a specific analysis and results.
-While secondary documentation has become more common practice, it is still often generated using highly specialized language.
+Good primary documentation for code describes general usage of individual functions,
+ the components and variables a function can take, and it should be accompanied with
+  function usage examples on how to apply it.
+As opposed to code, primary documentation is written in natural language (i.e.,
+ any known human language, e.g., English, Spanish, Chinese) and usually makes use
+  of highly specialized computational jargon (computationally specific concepts,
+   words, and phrases) as well as formal language, which often slows down or even
+    obstructs examination, application, and adoption of code by external individuals.
+Because primary documentation is considered a professional document, acceptance
+ of the research by the scientific community could be reduced if a more informal
+  language is used.
+Vignettes and tutorials work as secondary pieces of documentation, that help to demonstrate
+ additional cases of individual function usage, and showcase function associations
+  that work for specific analysis workflows in more detail.
+   While secondary documentation has become more common practice,
+    it is still often generated using highly specialized language.
 
-\subsection*{Adressing hurdles to accesibility: the principles}
+\subsection*{Adressing hurdles to accessibility: the principles}
 
@@
 narrative arc that is easier to follow and remember by users. This facilitates the translation
-of functions to specfic use cases in biology.
+of functions to specific use cases in biology.
 
@@
 work seamlessly, without errors. We argue that the opposite is needed to support
-user adoption of reproducibile workflows: demonstrate examples that do not work
-as expected and exemplify ways to address them (Figure \ref{fig:first}). We identify inputs that would give
+user adoption of reproducible workflows: demonstrate examples that do not work
+as expected and exemplify ways to address them (Figure \ref{fig:first}).
+We identify inputs that would give
 a wide range of warnings and errors, focusing on demonstrating these cases. This
@@
 Besides avoiding formal language, and incorporating elements of pop culture, such as picture
-character icons known as "emojis", to make the language used more familiar to the
-target audience (see Figure \ref{fig:first}), we made an effort to specifically
+character icons known as "emojis", to make the language more familiar to a
+broader target audience (see Figure \ref{fig:first}), we made an effort to specifically
 complement the primary documentation by identifying
@@
 
-\section{Results and Discussion}
-\label{sec:results}
+\section{Conclusion}
+\label{sec:conclusion}
 
-We explain the warnings and errors and design ways to avoid them, and detect them beforehand (i.e., before using an input that would give an error). We explain the consequences of warnings.
+We explain the warnings and errors and design ways to avoid them, and detect them
+ beforehand (i.e., before using an input that would give an error).
+  We explain the consequences of warnings.
 We designed ways to access the different elements of the outputs.
-We have received emails from senior researchers thanking us for this materials, and students have been able to engage using the packages with less hep from the PIs.
+We have received emails from senior researchers thanking us for this materials,
+ and students have been able to engage using the packages with less hep from the PIs.
 
-The principles to create tutorials described here facilitate adoption of software and analysis wokflows among researchers at different academic levels, from undergrads to established researchers.
-It will also help closing the gap between students that had access to computational resources (and computational training) from an early age and students that did not.
-Late access to computational resources and training can occur due to lack of economic resources, often ocurring in households from underrepresented communities and minorites.
-It can also be due to gender-biased parental and community pressures, in which males are more often encouraged to perform activities related to computers, while females are discouraged.
-How to balance software acceptance VS. adoption?
-These principles can be used to aide not only reproducibility, but also software adoption in the natural sciences.
-Discuss: why address accessibility and not other apects of reproducibility?
-
-\section{Conclusion}
-\label{sec:conc}
+The principles to create tutorials described here facilitate adoption of software
+ and analysis workflows among researchers at different academic levels, from undergrads
+  to established researchers.
+It will also help closing the gap between students that had access to computational
+ resources (and computational training) from an early age and students that did not.
+  Late access to computational resources and training can occur due to lack of
+   economic resources, often occurring in households from underrepresented communities
+    and minorities.
+It can also be due to gender-biased parental and community pressures,
+ in which male individuals are more often encouraged to perform activities related to computers,
+  while female individuals are discouraged.
+% How to balance software acceptance VS. adoption?
+These principles can be used to aide not only in improving reproducibility practices,
+ but also software adoption in the natural sciences.
+%% Discuss: why address accessibility and not other aspects of reproducibility?
 
@@
 scientific efforts can build off of each other
+
+Intentional integration of the subject of scientific reproducibility into
+ the curriculum is a key aspect to promote reproducible practices long term.
+There are different elements that permit fully reproducible science.
 

### 2022-02-03 jsdse/Manuscript.tex
adding to conclusions

@@
 \spacingset{1.45} % DON'T change the spacing!
-\section{Introduction}
+\section*{Introduction}
 \label{sec:intro}
@@
     \parbox{\textwidth}{%
-    \textbf{Box 1.}
-      Availability does not imply accessibility. One example we find really useful is the
-      marshmallows in an office story.
-      Brand new marshmallow bags remained in a common office area after a social gathering.
-      The marshmallows did not belong to anyone anymore, so they were freely
-      available for eating by anyone in the office.
+    \textbf{Box 1. The role of accessibility to achieve full availability.}
+      Availability is generally defined as the ability or potential of something
+      to be used and accessed (\cite{availability2022cambridge}).
+      Following this definition, accessibility is generally considered a synonym of availability.
+      Yet, in practice, availability does not imply accessibility.
+      One example we find really useful to practically differentiate the concepts of availability and accessibility
+      is the "marshmallows in an office" story, which goes like this.
+      A couple of brand new marshmallow bags remained in a common office area after
+      a social gathering.
+      The marshmallows were not claimed by anybody, so they were left in the common are,
+      freely available to be taken or eaten by anyone in the office.
       Yet, the marshmallow bags stayed in the common area, unopened for days and even weeks.
-      It was not until someone opened the bags and placed the marshmallows on a tray,
+      It was not until someone decided to open the bags and placed the marshmallows on a tray,
       that people started actually eating them. They were gone in a matter of hours.
-      Scientific products are not sweet as marshmallows, but they are a coveted asset.
+      Scientific products are not sweet as marshmallows, but they are a coveted resource.
       While sharing, describing and documenting makes scientific products available,
@@
       something that would be useful long term, the short term time investment
-      is often too high for individuals to engage with it.
+      is often too high for individuals to engage with it, making the workflow and its results
+      effectively inaccessible and hence unavailable to the majority of the community.
     }%
@@
 To elaborate on our thesis, we use an example in evolutionary biology that focuses
-on understanding the shared ancestry among organisms through phylogenetic trees, which provide the
+on understanding the shared ancestry of organisms through phylogenetic trees, which provide the
 basis to study and understand biological processes in an evolutionary context (\cite{dobzhansky1973nothing}).
 As such, improving reproducibility and availability
-of phylogenetic research is relevant for many aspects of biological research.
+in phylogenetic research is relevant for many aspects of biological research.
 The Open Tree of Life project (OpenTree) has developed a platform that facilitates
@@
 
-\section{Methods}
-\label{sec:meth}
-\subsection*{Identifying hurdles to accessibility}
+\section*{Identifying hurdles to accessibility}
+\label{sec:identifying}
 
@@
 
-\subsection*{Adressing hurdles to accessibility: the principles}
+\section*{Addressing hurdles to accessibility: some principles}
+\label{sec:addressing}
 
-\subsubsection*{1. Demonstrate integration of function usage with motivating examples}
+\subsubsection*{a. Demonstrate integration of function usage with motivating examples}
 
@@
 
-\subsubsection*{2. Demonstrate errors and warnings thoroughly}
+\subsubsection*{b. Demonstrate errors and warnings thoroughly}
 
@@
 as they inevitably will.
+On our tutorial, we focus on explaining the meaning and downstream coof warnings and errors, and
+ showcase ways to detect them before they are triggered (i.e., before using an input
+  that would elicit a warning or error). This has two pedagogical benefits:
+1) it provides users/students with the means to troubleshoot their own warnings and erros, and
+2) it allows users/students to understand with more depth what the function is doing.
+
+We designed ways to access the different elements of the outputs.
 
@@
 
-\subsubsection*{3. Avoid jargon and expert language}
+\subsubsection*{c. Avoid jargon and expert language}
 
@@
 
-\subsubsection*{4. Make it stable through time}
+\subsubsection*{d. Make it stable through time}
 
@@
 
-\section{Conclusion}
+\section*{Conclusion}
 \label{sec:conclusion}
 
-We explain the warnings and errors and design ways to avoid them, and detect them
- beforehand (i.e., before using an input that would give an error).
-  We explain the consequences of warnings.
-We designed ways to access the different elements of the outputs.
+Ultimately, the improvement of reproducibility rates in science long term depends
+on our ability to intentionally integrate the subject of reproducibility in the
+the curriculum for college students and aspiring researchers to develop the skills
+necessary to create reproducible scientific workflows and materials.
+
+Some universities have been incorporating the subject in their classes (see University
+ of Washington and The National Institute for General Medical Sciences).
+The focus has been on students to develop skills to document their work.
+The principles identified and outlined here can be used to set learning goals and
+ outcomes on new reproducibility syllabus.
+
 We have received emails from senior researchers thanking us for this materials,
@@
 
-Intentional integration of the subject of scientific reproducibility into
- the curriculum is a key aspect to promote reproducible practices long term.
-There are different elements that permit fully reproducible science.
-
 \bigskip

### 2022-02-04 jsdse/Manuscript.tex
adding references

@@
 basis to study and understand biological processes in an evolutionary context (\cite{dobzhansky1973nothing}).
-As such, improving reproducibility and availability
+As such, improving reproducibility rates
 in phylogenetic research is relevant for many aspects of biological research.
-The Open Tree of Life project (OpenTree) has developed a platform that facilitates
-availability of results from phylogenetic research, by standardizing and storing
-phylogenetic data with the goal of synthesizing a single phylogenetic tree encompassing
+The Open Tree of Life project (OpenTree) has developed a computational platform that provides
+availability of results from phylogenetic research, by creating a standardized database of
+phylogenetic data that is used to synthesize a single phylogenetic tree encompassing
 all life (\cite{opentreeoflife2019synth}).
-All data in OpenTree is open access and available programmatically through its many
+The OpenTree database Phylesystem (\cite{}) is open access and available programmatically through its many
 Application Programming Interface (API) services (\cite{opentreeAPIs}).
@@
 required in the field.
-Baseline computational skill requirements are likely increasing across all natural sciences.
-Computing is not traditionally a core skill taught to biologists and naturalists.
-However, many students are now being trained in R as a statistics and data analysis environment.
+"Computers are playing an increasingly important role in many scientific disciplines" (\cite{piccolo2016tools}).
+Baseline computational skill requirements are increasing across all natural sciences.
+Yet, computing is not traditionally a core skill taught to biologists and naturalists
+(\cite{sayres2018bioinformatics, wright2019the, williams2019barriers}).
+Still, many students are now being trained in R programming language as a statistics and data analysis environment.
+% "Thus efforts to increase scientific reproducibility should consider the ubiquity of computers in research."
 In order for reproducible computational tools to be adopted for research, they
@@
 Because primary documentation is considered a professional document, acceptance
- of the research by the scientific community could be reduced if a more informal
+ of the research by the scientific community could diminish if a more informal
   language is used.
@@
   that work for specific analysis workflows in more detail.
-   While secondary documentation has become more common practice,
-    it is still often generated using highly specialized language.
+As secondary documentation has become more common practice, we identify and propose
+a series of principles to support improved accessibility for secondary documentation.
 
@@
 
-\subsubsection*{a. Demonstrate integration of function usage with motivating examples}
+\subsubsection*{a. Literate programming: Demonstrate code usage with integrative examples}
+
+%% By definition, all languages are abstract. We learn them because we grow up with them and they have a direct application for our lives.
+%% Programming languages are no different, however, we start learning those later in life and we do not use them as regularly, making it harder for us to learn them and use them.
+Programming languages are abstract and usually hard to remember without
+an obvious link to an application (\cite{knuth1984literate}).
+A story or narrative that links pieces of code together can greatly support users/students
+to remember the code and its usage.
+The "literate programming" paradigm (\cite(knuth1984literate))
+integrates narratives with computer code in the same document, to support users
+in following the code usage, improving reproducibility of scientific analyses (\cite(piccolo2016tools)).
 
@@
 as they inevitably will.
-On our tutorial, we focus on explaining the meaning and downstream coof warnings and errors, and
+On our tutorial, we focus on explaining the meaning and downstream of warnings and errors, and
  showcase ways to detect them before they are triggered (i.e., before using an input

### 2022-02-04 jsdse/Manuscript.tex
adding references

@@
 
+A practice that has become more and more widespread in programming languages pedagogical practices
+is the use of typos and mistakes to normalize them for learners and show them how
+to solve them when they are outside the classroom (\cite{shannon2015live}).
+Yet, this is rarely done for written pedagogical materials.
 Primary documentation focuses on demonstrating usage function with examples that
 work seamlessly, without errors. We argue that the opposite is needed to support
-user adoption of reproducible workflows: demonstrate examples that do not work
+adoption of reproducible workflows and support long term independence in learners
+(\cite{gaspar2007restoring}).
+We demonstrate examples that do not work
 as expected and exemplify ways to address them (Figure \ref{fig:first}).
+
 We identify inputs that would give

### 2022-02-04 jsdse/Manuscript.tex
beefing it up

--- 2022-02-04 jsdse/Manuscript.tex
@@
 % NOTE: To produce blinded version, replace "0" with "1" below.
-\newcommand{\blind}{1}
+\newcommand{\blind}{0}
 
@@
 %\bibliographystyle{natbib}
+%\bibliographystyle{bmc-mathphys}
+\bibliographystyle{agsm}
 
@@
   \author{Luna L. Sanchez Reyes\thanks{
-    The authors gratefully acknowledge "Sustaining the Open Tree of Life", NSF ABI No. 1759838, and ABI No. 1759846.}\hspace{.2cm}\\
+    The authors gratefully acknowledge ``Sustaining the Open Tree of Life``, NSF ABI No. 1759838, and ABI No. 1759846.}\hspace{.2cm}\\
     School of Natural Sciences, University of California, Merced\\
@@
 Reproducibility --the extent to which consistent results are obtained when a scientific
-research experiment is repeated (\cite{repdef2021})-- is a key aspect for the advancement
-of science, as it constitutes a minimum standard that allows understanding scientific products
-(methods, analysis, results),
+research experiment is repeated \citep{repdef2021}-- is a key aspect for the advancement
+of science, as it constitutes a minimum standard that allows understanding research products
+(e.g., methods, data, analysis, results, etc.),
 to determine their reliability and generality, and eventually build up scientific
 knowledge and applications based on those products
-(\cite{king1995replication, peng2011reproducible, powers2019open}).
-In the natural sciences, rates of reproducibility are low (\cite{ioannidis2005most, prinz2011believe}),
-prompting concerns about a reproducibility crisis in the field (\cite{baker2016reproducibility}).
+\citep{king1995replication, peng2011reproducible, powers2019open}.
+In the natural sciences, rates of reproducibility are low \citep{ioannidis2005most, prinz2011believe},
+which has elicited concerns about a reproducibility crisis in the field \citep{baker2016reproducibility}.
 %%% The scientific community has united to incentivize cultural changes that will improve
 %%% reproducibility rates long term, such as transparency, availability, and workflow
-%%% automatization, to name a few (\cite{peng2015reproducibility}).
-In response, the scientific community has been generating standards and incentivizing
-cultural changes in an effort to improve reproducibility rates long term
-(\cite{peng2015reproducibility, wilkinson2016fair}).
-One standard that has received much attention is scientific availability, which
- we define as the property of research objects to be acquired, copied, analyzed,
-  processed and/or reused, at no financial, legal or technical cost (\cite{arnold2019turing}),
-   and with no geographic, demographic or temporal barriers (\cite{fecher2014open}).
-In this paper, we argue that availability can not be fully achieved without accessibility (Box 1)
+%%% automatization, to name a few \citep{peng2015reproducibility}.
+In response, the scientific community has been developing new principles and standards to incentivize
+cultural changes in an effort to improve long term reproducibility rates in the natural sciences
+\citep{peng2015reproducibility, wilkinson2016fair}.
+A standard for reproducibility that has received much attention is scientific availability, which
+we define as the property of research products to be acquired, copied, analyzed,
+processed and/or reused, at no financial, legal or technical cost \citep{arnold2019turing},
+and with no geographic, demographic or temporal barriers for the population \citep{fecher2014open}.
+In this paper, we argue that availability can not be fully achieved without accessibility (Box 1).
 We identify factors that have particularly affected accessibility in
-the natural sciences, and ways researchers can address them to ensure reproducible
-and automatic scientific workflows.
+the natural sciences, and ways scientific developers can address them to ensure reproducible
+and automatic scientific workflows. This principles can be generalized and integrated
+into the reproducibility curriculum to hopefully provide present learners and future researchers with
+the tools for successful scientific reproducibility.
 % full workflow automatization and reproducibility in the natural sciences
@@
       Availability is generally defined as the ability or potential of something
-      to be used and accessed (\cite{availability2022cambridge}).
+      to be used and accessed \citep{availability2022cambridge}.
       Following this definition, accessibility is generally considered a synonym of availability.
@@
       One example we find really useful to practically differentiate the concepts of availability and accessibility
-      is the "marshmallows in an office" story, which goes like this.
+      is the ``marshmallows in an office`` story, which goes like this.
       A couple of brand new marshmallow bags remained in a common office area after
@@
 
-To elaborate on our thesis, we use an example in evolutionary biology that focuses
-on understanding the shared ancestry of organisms through phylogenetic trees, which provide the
-basis to study and understand biological processes in an evolutionary context (\cite{dobzhansky1973nothing}).
-As such, improving reproducibility rates
-in phylogenetic research is relevant for many aspects of biological research.
+To elaborate on our thesis, we use a widely used scientific workflow in phylogenetics,
+a research field within evolutionary biology that focuses
+on understanding the shared ancestry of living organisms through the reconstruction of
+phylogenetic trees from biological data. Phylogenies provide the
+basis to study and understand all biological processes in an evolutionary context
+\citep{dobzhansky1973nothing}.
+Given the importance of phylogenies, improving reproducibility rates
+in phylogenetic research is increasingly relevant for many aspects of the natural sciences.
 The Open Tree of Life project (OpenTree) has developed a computational platform that provides
@@
 phylogenetic data that is used to synthesize a single phylogenetic tree encompassing
-all life (\cite{opentreeoflife2019synth}).
-The OpenTree database Phylesystem (\cite{}) is open access and available programmatically through its many
-Application Programming Interface (API) services (\cite{opentreeAPIs}).
-Additionally, R packages (\cite{michonneau2016rotl}) and Python libraries
-(\cite{mctavish2021opentree}) have been developed as wrappers for OpenTree API services
-to make them available to a wider programming audience.
-The R and Python OpenTree wrappers have been utilized by computer-literate individuals,
+all life \citep{opentreeoflife2019synth}.
+The OpenTree database Phylesystem \citep{mctavish2015phylesystem} is programmatically
+available free of cost to users through OpenTree's Application Programming Interface
+(API) services \citep{opentreeAPIs}.
+Additionally, software packages written with open source, free of cost programming
+languages commonly used in scientific research such as R and Python
+\citep{baker2017scientific} have been developed as wrappers that have increased the availability
+of OpenTree's API services, by making the OpenTree API functionalities more accessible
+to a wider user audience \citep{michonneau2016rotl, mctavish2021opentree}.
+Yet, the R and Python OpenTree API wrapper software packages have been mainly used
+by computer-literate individuals
 to seamlessly establish reproducible workflows to use and reuse expert phylogenetic
-knowledge for biological research (\cite{sanchez2019datelife}) and education
-(\cite{nguyen2020phylotastic, phylotasticedtools, galacticedtools}).
+knowledge for biological research \citep{sanchez2019datelife, sanchez2021physcraper}
+and education \citep{nguyen2020phylotastic, phylotasticedtools, galacticedtools}.
+In the 5 years since its release, the R package wrapper for OpenTree \texttt{rotl}
+has been cited by more than 190 biological research papers addressing various
+topics \citep{googlerotl}.
+Comparatively, other R packages for phylogenetics, such as \texttt{GGTREE}
+and \texttt{mixOmics}, which were released one year after \texttt{rotl} have been
+cited over a thousand times, which might indicate that \texttt{rotl} has not been
+adopted as widely.
+%% We need a proper statistical distribution of citation numbers per year, to see if rotl is an outlier or not.
 
-The OpenTree project demonstrates that efforts to increase availability for reproducibility
-in phylogenetics have also increased baseline computational knowledge and skills
-required in the field.
-"Computers are playing an increasingly important role in many scientific disciplines" (\cite{piccolo2016tools}).
-Baseline computational skill requirements are increasing across all natural sciences.
-Yet, computing is not traditionally a core skill taught to biologists and naturalists
-(\cite{sayres2018bioinformatics, wright2019the, williams2019barriers}).
-Still, many students are now being trained in R programming language as a statistics and data analysis environment.
+While learners in the natural sciences have been engaging independently with R
+and Python programming language
+as they represent two of the most widely used programming languages in the sciences
+% for their fast and seamless statistics and data analysis environment
+\citep{baker2017scientific}, computer programming is not traditionally
+a core skill formally taught to biologists and naturalists
+\citep{sayres2018bioinformatics, wright2019the, williams2019barriers},
+However, the OpenTree project demonstrates that efforts to improve availability and reproducibility
+in phylogenetics have increased required baseline computational knowledge and skills
+in the field.
+As computers continue to play a larger role in most scientific disciplines \citep{piccolo2016tools},
+baseline computational skill requirements are also increasing across all natural sciences.
 % "Thus efforts to increase scientific reproducibility should consider the ubiquity of computers in research."
-In order for reproducible computational tools to be adopted for research, they
- need to be much more accessible for present and future researchers.
-Data and code availability is a core requirement for reproducibility research
- (\cite{peng2011reproducible, sandve2013ten, powers2019open}).
-However, the utility of data resources is limited by the technical challenges of accessing the data.
-In order to motivate reproducible research, that gap needs to be bridged.
-In this work we focus on improving accessibility of code examples and documentation.
-In particular, we identify the necessity for documentation that is written down
- using language that is common to the target audience to facilitate examination,
-  application, and adoption of code by the wider audience.
+% In order for reproducible computational tools to be adopted for research, they
+% need to be readily accessible for present and future researchers.
+Thus, efforts to increase reproducibility rates in the natural sciences must consider
+the particularities of scientific workflows that rely on usage of programming languages, such as
+availability and accessibility of data and code \citep{peng2011reproducible, sandve2013ten, powers2019open}.
+% However, the utility of data resources is limited by the technical challenges of accessing the data.
+% In order to motivate reproducible research, that gap needs to be bridged.
+% In this work we focus on improving accessibility of code examples and documentation.
 
-We present a set of principles to generate documentation that improves accessibility
- of code and documentation. We applied these principles to a series of tutorials
-  and vignettes developed for the OpenTree project.
+We chose a phylogenetics workflow that relies on data and code from OpenTree to
+identify the specific barriers to accessibility and reproducibility it presents.
+Then, we design ways to overcome or diminish these barriers and apply them to
+a series of tutorials and vignettes developed for the OpenTree project.
+Finally, we generalize our finding to suggest a set of
+principles that can be used as guide to develop code and documentation materials
+with increased accessibility that can improve availability and contribute to reproducibility.
+Notably, these principles can be incorporated as learning goals into any syllabus for a
+course or workshop on best practices for scientific reproducibility.
+% In particular, we identify the necessity for documentation that is written down
+% using language that is common to the target audience to facilitate examination,
+% application, and adoption of code by the wider audience.
 
@@
 
-Good primary documentation for code describes general usage of individual functions,
+Good primary documentation for code is thorough. It describes general usage of individual functions,
 the components and variables a function can take, and it should be accompanied with
-  function usage examples on how to apply it.
+function usage examples on how to apply it \citep{karimzadeh2018top}.
 As opposed to code, primary documentation is written in natural language (i.e.,
- any known human language, e.g., English, Spanish, Chinese) and usually makes use
-  of highly specialized computational jargon (computationally specific concepts,
-   words, and phrases) as well as formal language, which often slows down or even
-    obstructs examination, application, and adoption of code by external individuals.
-Because primary documentation is considered a professional document, acceptance
- of the research by the scientific community could diminish if a more informal
-  language is used.
+any known human language, e.g., English, Spanish, Chinese).
+Primary documentation is viewed as a key element for success of a piece of code
+\citep{karimzadeh2018top},
+which might be why it is also usually written using highly specialized computational jargon (i.e., computationally specific concepts,
+words, and phrases) as well as formal scientific language.
+While this might be important for formal acceptance of the code by the scientific
+and academic community, it often slows down or even
+obstructs examination, application, and adoption of code by the general audience \citep{ball2017its}.
+
 Vignettes and tutorials work as secondary pieces of documentation, that help to demonstrate
@@
 that work for specific analysis workflows in more detail.
-As secondary documentation has become more common practice, we identify and propose
-a series of principles to support improved accessibility for secondary documentation.
+As secondary documentation has become more common practice and is more flexible
+in its form and content, it constitutes an ideal canvas to develop, implement and
+test principles that can overcome current barriers to code accessibility.
+The tutorials are available at https://mctavishlab.github.io/R\_OpenTree\_tutorials/.
 
@@
 
+%% I think I have three different points here:
+%% internalizing knowledge is made easier with active learning practices: learn by doing!
+%% literate programming allows learning by doing AND at the same time provides a story that connects the acquired knowledge to a real worl problem.
 %% By definition, all languages are abstract. We learn them because we grow up with them and they have a direct application for our lives.
 %% Programming languages are no different, however, we start learning those later in life and we do not use them as regularly, making it harder for us to learn them and use them.
-Programming languages are abstract and usually hard to remember without
-an obvious link to an application (\cite{knuth1984literate}).
-A story or narrative that links pieces of code together can greatly support users/students
-to remember the code and its usage.
-The "literate programming" paradigm (\cite(knuth1984literate))
-integrates narratives with computer code in the same document, to support users
-in following the code usage, improving reproducibility of scientific analyses (\cite(piccolo2016tools)).
+Pedagogical research shows that active learning practices are one the most effective
+ways to take on abstract subjects \citep{freeman2014active}.
+Programming computer languages are quite abstract and learning them can be greatly
+enhanced by applying an active learning strategy such as a linking its usage to
+a ``real world'' or ``human'' application \citep{felder2009active}.
 
-We examined available primary documentation for the OpenTree API R wrapper (the package `rotl`),
-and designed a workflow that visits as many functions as possible, and demonstrate
-uses commonly requested by OpenTree users that are not demonstrated elsewhere.
+A story-like narrative that links pieces of code together and invites learners
+to try the code can greatly support learners to remember what they are doing and
+why they are doing it.
+This ``literate programming'' paradigm \citep{knuth1984literate, fritzson2002mathmodelica}
+makes code more approachable, as it integrates narratives with computer code in
+the same document, supporting learners in actively following, remembering and understanding
+the code usage \citep{piccolo2016tools}.
+Documents developed with ``literate programming'' can be made more accessible by
+choosing narratives that are relatable to a more general audience.
+
+We examined available primary documentation for the package \texttt{rotl},
+and designed a narrative that required the usage of as many functions as possible.
+We demonstrate code applications that are commonly requested by OpenTree users,
+but that are not demonstrated in the R package primary documentation.
 By framing the function workflow using highly requested uses, the documentation acquires a
-narrative arc that is easier to follow and remember by users. This facilitates the translation
-of functions to specific use cases in biology.
+narrative arc that is easier to follow and remember by users. This can also facilitate the application
+of code to other use cases in biology of interest for the learners.
 
-For the tutorial demonstrated here, we used the commonly requested use case of obtaining
-a phylogenetic tree for all lineages within a specific taxonomic rank.
+%% Maybe Figure!! For the tutorial demonstrated here, we used the commonly requested use case of obtaining
+%% a phylogenetic tree for all lineages within a specific taxonomic rank.
 
@@
 
+A practice that has become more and more widespread in programming languages pedagogical practices
+is the use of typos and mistakes to normalize them for learners and show them how
+to solve them when they are outside the classroom \citep{shannon2015live}.
+Yet, this is rarely done for written pedagogical materials.
 Primary documentation focuses on demonstrating usage function with examples that
 work seamlessly, without errors. We argue that the opposite is needed to support
-user adoption of reproducible workflows: demonstrate examples that do not work
+adoption of reproducible workflows and support long term independence in learners
+\citep{gaspar2007restoring}.
+We demonstrate examples that do not work
 as expected and exemplify ways to address them (Figure \ref{fig:first}).
+
 We identify inputs that would give
@@
   that would elicit a warning or error). This has two pedagogical benefits:
-1) it provides users/students with the means to troubleshoot their own warnings and erros, and
+1) it provides users/students with the means to troubleshoot their own warnings and errors, and
 2) it allows users/students to understand with more depth what the function is doing.
 
-We designed ways to access the different elements of the outputs.
+% We designed ways to access the different elements of the outputs.
 
@@
 Besides avoiding formal language, and incorporating elements of pop culture, such as picture
-character icons known as "emojis", to make the language more familiar to a
+character icons known as ``emojis``, to make the language more familiar to a
 broader target audience (see Figure \ref{fig:first}), we made an effort to specifically
@@
 We published the tutorials on a public, free license, free of cost, and free for
-use and reuse repository and persistent website (\cite{RopentreeTutorials, RopentreeTutorialsWebsite}).
+use and reuse repository and persistent website \citep{RopentreeTutorials, RopentreeTutorialsWebsite}.
 The tutorial is available for the users to go back to it any time they need it,
@@
 
-Ultimately, the improvement of reproducibility rates in science long term depends
-on our ability to intentionally integrate the subject of reproducibility in the
-the curriculum for college students and aspiring researchers to develop the skills
-necessary to create reproducible scientific workflows and materials.
+Ultimately, the long term improvement of reproducibility rates in science will depend
+on our ability to intentionally integrate the subject of reproducibility into the
+undergraduate curriculum, so college learners and future researchers have the
+basis to develop the fundamental skills needed to successfully create reproducible
+scientific workflows and materials.
 
-Some universities have been incorporating the subject in their classes (see University
- of Washington and The National Institute for General Medical Sciences).
-The focus has been on students to develop skills to document their work.
+Some universities have been incorporating the subject in their classes (see
+\cite{uwlibraries2022, nigms2022}).
+The focus of this resources has been for students to develop skills to document their work.
 The principles identified and outlined here can be used to set learning goals and
@@
 
-\bibliographystyle{agsm}
-
 \bibliography{Manuscript-bibliography}

### 2022-02-09 jsdse/Manuscript.tex
reviews: incorporating mega lab meeting comments

@@
 {
-  \title{\bf Promoting analysis reproducibility with accessibility: An example in evolutionary biology}
+  \title{\bf Teaching analysis reproducibility with accessibility: Enriching secondary documentation to support skill level diversity of users.}
   \author{Luna L. Sanchez Reyes\thanks{
@@
   \begin{center}
-    {\LARGE\bf Promoting analysis reproducibility with accessibility: An example in evolutionary biology}
+    {\LARGE\bf Teaching analysis reproducibility with accessibility: Enriching secondary documentation to support skill level diversity of users.}
 \end{center}
@@
 \begin{abstract}
-Reproducibility is essential for scientific development. Efforts
-to increase scientific reproducibility have focused on increasing availability
-of code and data. However, availability does not imply accessibility, and the latter
-is infrequently addressed, even if it is key to achieve full workflow automatization
-and reproducibility. Using an example in evolutionary biology, we identify factors
-that have specifically affected accessibility in
-the natural sciences, and ways researchers can address this to ensure reproducible
-and automatic workflows.
-The Open Tree of Life project (OpenTree) has developed a platform that facilitates
-availability of results from evolutionary biology research. However, baseline
-computational knowledge and skills required to
-access scientific results are often not found among target users. While documentation
-is available, it is often written using highly specialized language that is also
- inaccessible for the average target user.
-We identified a set of principles to improve accessibility of OpenTree scientific data,
-implement them in a tutorial, and elaborate on their general application to code and
-documentation of scientific workflows.
+Research reproducibility is essential for scientific development. Yet, rates of reproducibility
+ are low, especially in the natural sciences. In the current scientific landscape,
+most research relies on computing tools and software. Hence, efforts for increasing
+reproducibility rates have focused on providing availability of scientific workflows
+as computer code, as well as raw and processed data in a digital form.
+However, making these research products widely available, usually online, does not
+necessarily ensures that
+they can be easily reached by the majority of the population,
+making them effectively inaccessible, counteracting availability and hindering reproducibility.
+
+In this paper, we argue that accessibility of resources is a key aspect of research reproducibility,
+yet, to our knowledge, it has been widely overlooked.
+We recognize that accessibility has many angles (social, psychological,
+education level), but we focus on lack of accessibility at the skill and pedagogical level.
+Using an example within evolutionary biology, we identify barriers to
+accessibility in the natural sciences:
+%% list the barriers
+highly specialized jargon,
+Then, we propose a set of principles to improve accessibility of code and data,
+and illustrate their application in a set of teaching materials for evolutionary biology that we put
+to work in the field.
+Finally, we elaborate on the general application of these principles for
+documenting scientific workflows, to provide present learners
+and future researchers with tools for successful scientific reproducibility.
+%> ** add a sentence on implications for teaching reproducibility
+
+%% Projects that address availability of code and data in the natural sciences
+%% have focused on producing documentation that describes the nature and usage of the
+%% resources.
+%% However, it is often written using highly specialized language that is largely
+%% inaccessible for the average target user. Additionally, baseline
+%% computational knowledge and skills required to access scientific results are often
+%% not found among potential users, and keep on increasing.
+
+%% **Introduce OpenTree in abstract??
 
@@
 \noindent%
-{\it Keywords:}  open science, education, R, phylogenetics, tutorials
+{\it Keywords:}  open science, R, phylogenetics, Open Tree of Life
 \vfill
@@
 Reproducibility --the extent to which consistent results are obtained when a scientific
-research experiment is repeated \citep{repdef2021}-- is a key aspect for the advancement
+research experiment or workflow is repeated \citep{repdef2021}--
+is a key aspect of the advancement
 of science, as it constitutes a minimum standard that allows understanding research products
-(e.g., methods, data, analysis, results, etc.),
+(e.g., methods, data, analysis, results, etc.)
 to determine their reliability and generality, and eventually build up scientific
@@
 A standard for reproducibility that has received much attention is scientific availability, which
-we define as the property of research products to be acquired, copied, analyzed,
+we define as a property denoting that a research product can be acquired, copied, analyzed,
 processed and/or reused, at no financial, legal or technical cost \citep{arnold2019turing},
@@
 We identify factors that have particularly affected accessibility in
-the natural sciences, and ways scientific developers can address them to ensure reproducible
-and automatic scientific workflows. This principles can be generalized and integrated
-into the reproducibility curriculum to hopefully provide present learners and future researchers with
-the tools for successful scientific reproducibility.
-% full workflow automatization and reproducibility in the natural sciences
+the natural sciences, and ways the scientific community can address them to create reproducible
+scientific workflows. These principles can be generalized and integrated
+into the undergraduate and graduate school STEM curriculum, either for courses specialized
+in reproducibility or within other subject areas as a necessary component of
+successful and impactful science.
+
+% ** describe/introduce reproducibility curriculum here
 
@@
     \parbox{\textwidth}{%
-    \textbf{Box 1. The role of accessibility to achieve full availability.}
-      Availability is generally defined as the ability or potential of something
-      to be used and accessed \citep{availability2022cambridge}.
-      Following this definition, accessibility is generally considered a synonym of availability.
-      Yet, in practice, availability does not imply accessibility.
-      One example we find really useful to practically differentiate the concepts of availability and accessibility
-      is the ``marshmallows in an office`` story, which goes like this.
+    \textbf{Box 1. The role of accessibility in improving reproducibility.}
+      % Availability is defined as the fact or possibility that something can be bought,
+      % used, or reached by someone \citep{availability2022cambridge}.
+      % Accessibility is defined as "the fact of being able to be reached or obtained easily"
+      % \citep{accessibility2022cambridge}
+      Availability and accessibility are conceptually accepted as synonyms for ``the fact
+      of being able to be reached" \citep{available-accessible2022cambridge}.
+      However, accessibility's definition goes further and describes it as the
+      ``fact of being able to be reached \textbf{easily}" \citep{accessibility2022cambridge}.
+      A secondary definition of accessibility elaborates more on this: ``the quality
+      of being \textbf{easy to understand}''.
+      % Yet, in practice, availability does not imply accessibility.
+
+      What we realize from these definitions is that in practice, something might
+      be available but not necessarily accessible,
+      hindering the realization of availability.
+      To illustrate the practical difference between availability and accessibility we
+      will use an allegory that goes like this:
+
+      ** Change marshmallow story to cake
+
+      ** Explain different types of barriers to accessibility within the example, then within science
+
+      ** Explain why we focus on psychological/social/pedagogical barriers
+
       A couple of brand new marshmallow bags remained in a common office area after
       a social gathering.
-      The marshmallows were not claimed by anybody, so they were left in the common are,
+      The marshmallows were not claimed by anybody, so they were left in the common area,
       freely available to be taken or eaten by anyone in the office.
@@
       While sharing, describing and documenting makes scientific products available,
-      it does not necessarily makes them accessible. For example, documentation of an
+      it does not necessarily makes them accessible.
+
+      ** Now, explain different types of barriers to accessibility in science
+      (Jessica: Maybe say that lack of accessibility can be due to a number of
+      factors (physical, resources, economic), but one critical factor is the use
+      of impenetrable expert language.)
+
+      For example, documentation of an
       available scientific workflow might be written using expert language that is too foreign
-      for a general audience to successfully reproduce the workflow. While some individuals might
+      for a general audience to successfully reproduce the workflow.
+
+      While some individuals might
       be able to invest the time to learn the expert language to successfully reproduce
-      the workflow, this will not be the case for the majority, even if it is
+      the workflow, this will not be the case for the majority; even if it is
       something that would be useful long term, the short term time investment
@@
 
-To elaborate on our thesis, we use a widely used scientific workflow in phylogenetics,
-a research field within evolutionary biology that focuses
-on understanding the shared ancestry of living organisms through the reconstruction of
-phylogenetic trees from biological data. Phylogenies provide the
+To elaborate on our thesis,
+%% we use a common scientific workflow in phylogenetics,
+%% ** Jessica mentions that workflow is confusing as she thought it referred to the OpenTree
+%% workflow, but we actually mean the more general workflow of reconstructing a phylogenetic tree.
+%% She suggests something more general:
+we choose an example from the research field of phylogenetics,
+a discipline within evolutionary biology. Phylogenetics focuses
+on investigating the history of shared ancestry of living organisms from biological data,
+and represents
+this evolutionary history in a diagram known as phylogeny (a figure from the tutorial).
+Phylogenies provide the
 basis to study and understand all biological processes in an evolutionary context
 \citep{dobzhansky1973nothing}.
-Given the importance of phylogenies, improving reproducibility rates
-in phylogenetic research is increasingly relevant for many aspects of the natural sciences.
+Given the importance of phylogenies for biological research, improving reproducibility rates
+in phylogenetics is increasingly relevant for many aspects of the natural sciences.
+
+***What other platforms provide resources for reproducibility in phylogenetics???
+
 The Open Tree of Life project (OpenTree) has developed a computational platform that provides
-availability of results from phylogenetic research, by creating a standardized database of
-phylogenetic data that is used to synthesize a single phylogenetic tree encompassing
-all life \citep{opentreeoflife2019synth}.
-The OpenTree database Phylesystem \citep{mctavish2015phylesystem} is programmatically
+availability of results from vetted, state-of-the-art research in phylogenetics
+\citep{opentreeoflife2019synth}.
+%% by creating a standardized database of phylogenetic data that is used to synthesize
+%% a single phylogenetic tree encompassing all life
+OpenTree's database, the Phylesystem \citep{mctavish2015phylesystem}, provides
+accessibility of phylogenetic data by standardizing organism names across biological databases,
+and supporting digital data obtention in a computer readable format.
+While this is accessible through their Graphical User Interface,
+It does not support reproducibility
+Accessibility does not imply reproducibility
+In order for something to be highly reproducible, it has to be largely accessible.
+
+is programmatically
 available free of cost to users through OpenTree's Application Programming Interface
@@
 and \texttt{mixOmics}, which were released one year after \texttt{rotl} have been
-cited over a thousand times, which might indicate that \texttt{rotl} has not been
-adopted as widely.
-%% We need a proper statistical distribution of citation numbers per year, to see if rotl is an outlier or not.
+cited over a thousand times, which might indicate a delay \texttt{rotl}
+adoption. Of course one potential explanation is that OpenTree services do not have
+as wide an application as services provided by the afore mentioned R packages.
+However, the
+expressed enthusiasm (interest to use/inability to use the resource) by the target
+ audience and larger community might point to a different issue.
+%% We need a proper statistical distribution of citation numbers per year, to see
+%% if rotl is an outlier or not.
 
 While learners in the natural sciences have been engaging independently with R
-and Python programming language
+and Python programming languages
 as they represent two of the most widely used programming languages in the sciences
@@
 a core skill formally taught to biologists and naturalists
-\citep{sayres2018bioinformatics, wright2019the, williams2019barriers},
-However, the OpenTree project demonstrates that efforts to improve availability and reproducibility
-in phylogenetics have increased required baseline computational knowledge and skills
+\citep{sayres2018bioinformatics, wright2019the, williams2019barriers}.
+Moreover, the main usage of the OpenTree API services hints that tools intended to
+improve availability and reproducibility come at a cost for the potential users:
+ an increased required baseline computational knowledge and skillset
 in the field.
 As computers continue to play a larger role in most scientific disciplines \citep{piccolo2016tools},
-baseline computational skill requirements are also increasing across all natural sciences.
+higher baseline computational skills are required across all natural sciences.
 % "Thus efforts to increase scientific reproducibility should consider the ubiquity of computers in research."
@@
 Thus, efforts to increase reproducibility rates in the natural sciences must consider
-the particularities of scientific workflows that rely on usage of programming languages, such as
+the specifics of scientific workflows that rely on usage of programming languages, such as
 availability and accessibility of data and code \citep{peng2011reproducible, sandve2013ten, powers2019open}.
@@
 We chose a phylogenetics workflow that relies on data and code from OpenTree to
-identify the specific barriers to accessibility and reproducibility it presents.
+identify the specific barriers to accessibility it presents.
 Then, we design ways to overcome or diminish these barriers and apply them to
 a series of tutorials and vignettes developed for the OpenTree project.
-Finally, we generalize our finding to suggest a set of
-principles that can be used as guide to develop code and documentation materials
-with increased accessibility that can improve availability and contribute to reproducibility.
+The tutorials are available at https://mctavishlab.github.io/R\_OpenTree\_tutorials/.
+
+Finally, we generalize our findings to suggest a set of
+principles that can be used as a guide to develop code and documentation materials
+with increased accessibility, to improve availability and contribute to reproducibility.
 Notably, these principles can be incorporated as learning goals into any syllabus for a
@@
 test principles that can overcome current barriers to code accessibility.
-The tutorials are available at https://mctavishlab.github.io/R\_OpenTree\_tutorials/.
 

### 2022-02-09 jsdse/Manuscript.tex
reviews: more mega lab meeting comments

@@
 Programming computer languages are quite abstract and learning them can be greatly
-enhanced by applying an active learning strategy such as a linking its usage to
+enhanced by applying an active learning strategy such as linking its usage to
 a ``real world'' or ``human'' application \citep{felder2009active}.
@@
 A story-like narrative that links pieces of code together and invites learners
-to try the code can greatly support learners to remember what they are doing and
+to try the code can lead learners to remember what they are doing and
 why they are doing it.
@@
 as they inevitably will.
-On our tutorial, we focus on explaining the meaning and downstream of warnings and errors, and
+In our tutorial, we focus on explaining the meaning and downstream of warnings and errors, and
  showcase ways to detect them before they are triggered (i.e., before using an input
@@
 computational concepts that were assumed or were not explained in depth.
-We vetted the tutorials with an audience on workshops as well as individual user.
+We vetted the tutorials through feedback from workshop participants as well as individual users.
 We choose examples that are charismatic for the audience.
@@
 use and reuse repository and persistent website \citep{RopentreeTutorials, RopentreeTutorialsWebsite}.
-The tutorial is available for the users to go back to it any time they need it,
+The tutorial is available for the users to go back to any time they need it,
 and to be passed on to other users (Figure \ref{fig:second}).
@@
 
-Following the carpentries, we created a main version of the tutorial that
-is updated. Versions presented on workshops are a copy from the original repository,
-and represent a stable and temporal snapshot of the functions and workflows presented
-in the tutorial.
+We created a main version of the tutorial that is stable. Any updates to the tutorial are
+published as new versions, or tutorials for new workshops \citep{CARPENTRIES}
+ Versions presented at workshops are a copy from the original repository.
+They represent a temporally stable snapshot of functions and workflows presented
+during a workshop.
 
@@
 
+Response form the community has been invaluable in gauging success of the tutorials.
+Senior researchers often comment on the usefulness
+of the tutorials for their research, as well as supporting students in using the
+R packages with less help from them as PIs.
+
+Making accessible reproducible workflows has several advantages:
+
+- save explanation/training time when analyses are run again by students and collaborators;
+- save research time for yourself when analyses are run again with more data, a
+  different dataset, a different organism or biological model;
+- scientific efforts can build off of each other.
+
 Ultimately, the long term improvement of reproducibility rates in science will depend
@@
 \cite{uwlibraries2022, nigms2022}).
-The focus of this resources has been for students to develop skills to document their work.
+The focus of these resources has been for students to develop skills to document their work.
 The principles identified and outlined here can be used to set learning goals and
-outcomes on new reproducibility syllabus.
-
-We have received emails from senior researchers thanking us for this materials,
- and students have been able to engage using the packages with less hep from the PIs.
+outcomes on new reproducibility syllabi.
 
@@
   to established researchers.
-It will also help closing the gap between students that had access to computational
+It will also help close the gap between students that had access to computational
 resources (and computational training) from an early age and students that did not.
@@
 economic resources, often occurring in households from underrepresented communities
-    and minorities.
+and minorities \citep{REFERENCE-NEEDED}.
 It can also be due to gender-biased parental and community pressures,
 in which male individuals are more often encouraged to perform activities related to computers,
-  while female individuals are discouraged.
+while female individuals are discouraged \citep{REFERENCE-NEEDED}.
 % How to balance software acceptance VS. adoption?
-These principles can be used to aide not only in improving reproducibility practices,
+These principles can be used to improve not only reproducibility practices,
  but also software adoption in the natural sciences.
 %% Discuss: why address accessibility and not other aspects of reproducibility?
-
-Making accessible reproducible workflows has several advantages:
-save explanation/training time when analyses are run again by students and collaborators.
-save research time for yourself when analyses are run again with more data, a different dataset, a different organism or biological model.
-scientific efforts can build off of each other
 

### 2022-02-10 jsdse/Manuscript.tex
figure: tree from tutorial

@@
 Research reproducibility is essential for scientific development. Yet, rates of reproducibility
- are low, especially in the natural sciences. In the current scientific landscape,
-most research relies on computing tools and software. Hence, efforts for increasing
-reproducibility rates have focused on providing availability of scientific workflows
+are low, especially in the natural sciences. As increasingly more research is relying on
+computing tools and software, efforts for improving
+reproducibility rates have focused on providing availability of research workflows
 as computer code, as well as raw and processed data in a digital form.
-However, making these research products widely available, usually online, does not
+However, making research products digitally available, does not
 necessarily ensures that
-they can be easily reached by the majority of the population,
-making them effectively inaccessible, counteracting availability and hindering reproducibility.
+they can be easily reached by the majority of the target user base, let alone the general population,
+making them effectively inaccessible, counteracting availability, and hindering reproducibility.
 
-In this paper, we argue that accessibility of resources is a key aspect of research reproducibility,
-yet, to our knowledge, it has been widely overlooked.
-We recognize that accessibility has many angles (social, psychological,
-education level), but we focus on lack of accessibility at the skill and pedagogical level.
+In this paper, we argue that accessibility of resources is a key aspect for
+successful research reproducibility.
+While accessibility is a broad term that encompasses various aspects
+of the human experience (visual, motor, auditory, cognitive, social, psychological,
+educational, etc.), we particularly address barriers to accessibility at the
+technological, social and pedagogical level.
 Using an example within evolutionary biology, we identify barriers to
@@
 %% list the barriers
-highly specialized jargon,
+specialized jargon, user programming experience, cognitive load,
+
 Then, we propose a set of principles to improve accessibility of code and data,
@@
 
-Reproducibility --the extent to which consistent results are obtained when a scientific
-research experiment or workflow is repeated \citep{repdef2021}--
+Research reproducibility --the extent to which consistent results are obtained when
+a scientific experiment or research workflow is repeated \citep{repdef2021}--
 is a key aspect of the advancement
@@
 In the natural sciences, rates of reproducibility are low \citep{ioannidis2005most, prinz2011believe},
-which has elicited concerns about a reproducibility crisis in the field \citep{baker2016reproducibility}.
+which has elicited concerns about a crisis in the field \citep{baker2016reproducibility}.
 %%% The scientific community has united to incentivize cultural changes that will improve
@@
 A standard for reproducibility that has received much attention is scientific availability, which
-we define as a property denoting that a research product can be acquired, copied, analyzed,
-processed and/or reused, at no financial, legal or technical cost \citep{arnold2019turing},
-and with no geographic, demographic or temporal barriers for the population \citep{fecher2014open}.
+we define as a property denoting that a research product can be reached (acquired, copied, analyzed,
+processed and/or reused) at no financial, legal or technical cost \citep{arnold2019turing},
+and without geographic, demographic, social or temporal barriers for the population \citep{fecher2014open}.
 In this paper, we argue that availability can not be fully achieved without accessibility (Box 1).
+
+We recognize that accessibility is a term that encompasses various aspects
+of the human experience (visual, motor, auditory, cognitive, social, psychological,
+educational, etc.).
+However, some forms of accessibility do not support reproducibility. For example,
+Graphical User Interfaces
+and
+we focus on X types of accessibility
+
+% not all types of accessibility support reproducibility
+% while accessibility is key to support availability, and availability is ket for reproducibility,
+% some forms of accessibility do not support reproducibility.
+
 We identify factors that have particularly affected accessibility in
 the natural sciences, and ways the scientific community can address them to create reproducible
-scientific workflows. These principles can be generalized and integrated
+research workflows. These principles can be generalized and integrated
 into the undergraduate and graduate school STEM curriculum, either for courses specialized
-in reproducibility or within other subject areas as a necessary component of
+in reproducibility or within other subject areas, as a necessary component of
 successful and impactful science.
@@
 
+\section*{An example from phylogenetics}
+\label{sec:example}
+
 To elaborate on our thesis,
@@
 and represents
-this evolutionary history in a diagram known as phylogeny (a figure from the tutorial).
+this evolutionary history in a diagram known as phylogeny (Figure \ref{fig:tree}).
 Phylogenies provide the
 basis to study and understand all biological processes in an evolutionary context
-\citep{dobzhansky1973nothing}.
-Given the importance of phylogenies for biological research, improving reproducibility rates
-in phylogenetics is increasingly relevant for many aspects of the natural sciences.
+\citep{dobzhansky1973nothing}. Hence,
+% Given the importance of phylogenies for biological research,
+it appears that improving reproducibility rates
+in phylogenetics has the potential to positively impact research across the natural sciences.
 
@@
 and supporting digital data obtention in a computer readable format.
+
 While this is accessible through their Graphical User Interface,
 It does not support reproducibility
+
 Accessibility does not imply reproducibility
+
 In order for something to be highly reproducible, it has to be largely accessible.
@@
 
+\begin{figure}
+\begin{center}
+\includegraphics[width=3in]{fig-tree.png}
+\end{center}
+\caption{A phylogenetic tree from our tutorial website, extracted using OpenTree of Life tools and the rotl R package. \label{fig:second}}
+\end{figure}
+
 \section*{Identifying hurdles to accessibility}
@@
 A story-like narrative that links pieces of code together and invites learners
-to try the code can lead learners to remember what they are doing and
+to try the code, can lead learners to remember what they are doing and
 why they are doing it.
@@
 makes code more approachable, as it integrates narratives with computer code in
-the same document, supporting learners in actively following, remembering and understanding
-the code usage \citep{piccolo2016tools}.
-Documents developed with ``literate programming'' can be made more accessible by
-choosing narratives that are relatable to a more general audience.
+the same document, supporting learners in actively following
+the code usage, supporting memory and understanding \citep{piccolo2016tools}.
+
+We propose that documents developed with ``literate programming'' can be made more
+accessible by choosing narratives that are relatable to a more general audience.
+An easy way to do this in biology is choosing as model organism a taxon that a
+research group is more interested in studying. For the non-specialized user base,
+a highly charismatic group such as dinosaurs will do the trick.
 
@@
 We demonstrate code applications that are commonly requested by OpenTree users,
-but that are not demonstrated in the R package primary documentation.
+but that are not demonstrated in the primary documentation of the R package.
 By framing the function workflow using highly requested uses, the documentation acquires a
-narrative arc that is easier to follow and remember by users. This can also facilitate the application
-of code to other use cases in biology of interest for the learners.
+narrative arc that is easier to follow and remember by users. This can also facilitate
+the application of code to other use cases in biology of interest for the users.
 
-%% Maybe Figure!! For the tutorial demonstrated here, we used the commonly requested use case of obtaining
+%% Maybe Figure!! For the tutorial demonstrated here, we used the commonly requested
+%% use case of obtaining
 %% a phylogenetic tree for all lineages within a specific taxonomic rank.
@@
 
-A practice that has become more and more widespread in programming languages pedagogical practices
-is the use of typos and mistakes to normalize them for learners and show them how
+A practice that has become more and more widespread in programming-language pedagogical practices
+is the use of typos and mistakes to normalize them for learners, and show them how
 to solve them when they are outside the classroom \citep{shannon2015live}.
-Yet, this is rarely done for written pedagogical materials.
+Yet, this is rarely done on written pedagogical materials.
 Primary documentation focuses on demonstrating usage function with examples that

### 2022-02-10 jsdse/Manuscript.tex
reorganizing figure

@@
 \end{center}
-\caption{A phylogenetic tree from our tutorial website, extracted using OpenTree of Life tools and the rotl R package. \label{fig:second}}
+\caption{A phylogenetic tree from our tutorial website, extracted using OpenTree of Life tools and the rotl R package. \label{fig:tree}}
 \end{figure}
@@
 We demonstrate examples that do not work
-as expected and exemplify ways to address them (Figure \ref{fig:first}).
+as expected and exemplify ways to address them (Figure \ref{fig:error
+}).
 
@@
 \begin{center}
-\includegraphics[width=3in]{fig1.png}
+\includegraphics[width=3in]{fig-error.png}
 \end{center}
-\caption{Snapshot of a section of the tutorial website, where we demonstrate a common error. \label{fig:first}}
+\caption{Snapshot of a section of the tutorial website, where we demonstrate a common error. \label{fig:error}}
 \end{figure}
@@
 character icons known as ``emojis``, to make the language more familiar to a
-broader target audience (see Figure \ref{fig:first}), we made an effort to specifically
+broader target audience (see Figure \ref{fig:error}), we made an effort to specifically
 complement the primary documentation by identifying
@@
 The tutorial is available for the users to go back to any time they need it,
-and to be passed on to other users (Figure \ref{fig:second}).
+and to be passed on to other users (Figure \ref{fig:schedule}).
 
@@
 \begin{center}
-\includegraphics[width=3in]{fig2.png}
+\includegraphics[width=3in]{fig-schedule.png}
 \end{center}
-\caption{Snapshot of the home to our tutorial website, showing part of the schedule. \label{fig:second}}
+\caption{Snapshot of the home to our tutorial website, showing part of the schedule. \label{fig:schedule}}
 \end{figure}

### 2022-02-11 jsdse/Manuscript.tex
elaborating on accessibility types

@@
 computing tools and software, efforts for improving
-reproducibility rates have focused on providing availability of research workflows
-as computer code, as well as raw and processed data in a digital form.
-However, making research products digitally available, does not
+reproducibility rates have focused on digital availability of research workflows
+as computer code, as well as raw and processed data in computer readable form.
+However, making research products digitally available does not
 necessarily ensures that
-they can be easily reached by the majority of the target user base, let alone the general population,
+they can be easily reached by the majority of a target user base, let alone the general population,
 making them effectively inaccessible, counteracting availability, and hindering reproducibility.
 
-In this paper, we argue that accessibility of resources is a key aspect for
-successful research reproducibility.
-While accessibility is a broad term that encompasses various aspects
-of the human experience (visual, motor, auditory, cognitive, social, psychological,
-educational, etc.), we particularly address barriers to accessibility at the
+In this paper, we argue that accessibility of research products is a key aspect for
+improving reproducibility rates.
+While accessibility typically addresses differences in abilities (visual, hearing, cognitive).
+
+across individuals, we particularly address barriers to accessibility at the
 technological, social and pedagogical level.
+
 Using an example within evolutionary biology, we identify barriers to
@@
 
-We recognize that accessibility is a term that encompasses various aspects
-of the human experience (visual, motor, auditory, cognitive, social, psychological,
-educational, etc.).
-However, some forms of accessibility do not support reproducibility. For example,
-Graphical User Interfaces
-and
+We recognize that the term accessibility is used in relation to various aspects of
+the human experience. In relation to digital resources, these aspects have been broadly
+categorized in four axis: perception, operability, understanding and robustness \citep{chisholm2001web, caldwell2008web}
+% (visual, motor, auditory, cognitive, social, psychological, educational, etc.).
+Some forms of accessibility do not support automatization nor scalability.
+For example, Graphical User Interfaces (GUIs) are accessible, but not scalable and rarely automatizable.
+
 we focus on X types of accessibility
@@
 
-      What we realize from these definitions is that in practice, something might
+      What we realize from these definitions is that, in practice, something might
       be available but not necessarily accessible,
@@
       ** Explain why we focus on psychological/social/pedagogical barriers
+
+      ** Gauging accessibility of a resource is tricky because it depends on an
+      individual experience of what is in effect \textbf{easy} for someone.
+      \textbf{Easy} is a subjective quality that fully depends on individual perceptions
+      resulting from individual contexts.
+      As a society we are moving towards validating and welcoming all human experiences.
+      To improve accessibility levels in the natural sciences, it
+      is important to account for the factors that are creating different experiences
+      of what is \textbf{easy}.
 
@@
 and represents
-this evolutionary history in a diagram known as phylogeny (Figure \ref{fig:tree}).
+this evolutionary history in a diagram known as a phylogeny (Figure \ref{fig:tree}).
 Phylogenies provide the
@@
 The Open Tree of Life project (OpenTree) has developed a computational platform that provides
-availability of results from vetted, state-of-the-art research in phylogenetics
-\citep{opentreeoflife2019synth}.
+digital availability of phylogenetic results from vetted, state-of-the-art research
+in the field \citep{opentreeoflife2019synth}.
 %% by creating a standardized database of phylogenetic data that is used to synthesize
@@
 OpenTree's database, the Phylesystem \citep{mctavish2015phylesystem}, provides
-accessibility of phylogenetic data by standardizing organism names across biological databases,
+usability of phylogenetic data by standardizing organism names across biological databases,
 and supporting digital data obtention in a computer readable format.
@@
 available free of cost to users through OpenTree's Application Programming Interface
-(API) services \citep{opentreeAPIs}.
+(API) services \citep{opentreeAPIv3}.
 Additionally, software packages written with open source, free of cost programming
@@
 We demonstrate examples that do not work
-as expected and exemplify ways to address them (Figure \ref{fig:error
-}).
+as expected and exemplify ways to address them (Figure \ref{fig:error}).
 
@@
 We published the tutorials on a public, free license, free of cost, and free for
-use and reuse repository and persistent website \citep{RopentreeTutorials, RopentreeTutorialsWebsite}.
+use and reuse repository and persistent website \citep{RopentreeTutorials}.
 The tutorial is available for the users to go back to any time they need it,
@@
 We created a main version of the tutorial that is stable. Any updates to the tutorial are
-published as new versions, or tutorials for new workshops \citep{CARPENTRIES}
+published as new versions, or tutorials for new workshops \citep{wilson2006swc, SWCwebsite}
  Versions presented at workshops are a copy from the original repository.

### 2022-02-11 jsdse/Manuscript.tex
elaborating on inclusivity

@@
 making them effectively inaccessible, counteracting availability, and hindering reproducibility.
-
-In this paper, we argue that accessibility of research products is a key aspect for
-improving reproducibility rates.
-While accessibility typically addresses differences in abilities (visual, hearing, cognitive).
-
-across individuals, we particularly address barriers to accessibility at the
-technological, social and pedagogical level.
-
-Using an example within evolutionary biology, we identify barriers to
-accessibility in the natural sciences:
+Thus, accessibility of research products is a key aspect for the long term improvement
+of reproducibility rates.
+Using an example within evolutionary biology, we identify inclusivity barriers to accessibility,
+% accessibility typically refers to addressing differences in abilities (visual, hearing, cognitive).
+research products in the natural sciences:
 %% list the barriers
-specialized jargon, user programming experience, cognitive load,
+use of highly specialized computing jargon and programming techniques; high cognitive load,
+unspecified, unclear or lengthy goals; content-focused descriptions instead of user-focused;
+inflexible learning environment; unapproachable (cold or intimidating) language;
+and little to no diversity of representation of information.
 
@@
 
-      A couple of brand new marshmallow bags remained in a common office area after
-      a social gathering.
-      The marshmallows were not claimed by anybody, so they were left in the common area,
-      freely available to be taken or eaten by anyone in the office.
-      Yet, the marshmallow bags stayed in the common area, unopened for days and even weeks.
-      It was not until someone decided to open the bags and placed the marshmallows on a tray,
-      that people started actually eating them. They were gone in a matter of hours.
-      Scientific products are not sweet as marshmallows, but they are a coveted resource.
-      While sharing, describing and documenting makes scientific products available,
-      it does not necessarily makes them accessible.
-
       ** Now, explain different types of barriers to accessibility in science
@@
       of impenetrable expert language.)
-
-      For example, documentation of an
-      available scientific workflow might be written using expert language that is too foreign
-      for a general audience to successfully reproduce the workflow.
-
-      While some individuals might
-      be able to invest the time to learn the expert language to successfully reproduce
-      the workflow, this will not be the case for the majority; even if it is
-      something that would be useful long term, the short term time investment
-      is often too high for individuals to engage with it, making the workflow and its results
-      effectively inaccessible and hence unavailable to the majority of the community.
     }%

### 2022-02-12 jsdse/Manuscript.tex
rewriting to approachable

@@
 {
-  \title{\bf Teaching analysis reproducibility with accessibility: Enriching secondary documentation to support skill level diversity of users.}
+  \title{\bf Approachable case studies support learning and reproducibility in data science: An example from evolutionary biology}
   \author{Luna L. Sanchez Reyes\thanks{
@@
   \begin{center}
-    {\LARGE\bf Teaching analysis reproducibility with accessibility: Enriching secondary documentation to support skill level diversity of users.}
+    {\LARGE\bf Approachable case studies support learning and reproducibility in data science: An example from evolutionary biology}
 \end{center}
@@
 computing tools and software, efforts for improving
-reproducibility rates have focused on digital availability of research workflows
+reproducibility rates have focused on making available research workflows
 as computer code, as well as raw and processed data in computer readable form.
-However, making research products digitally available does not
-necessarily ensures that
-they can be easily reached by the majority of a target user base, let alone the general population,
-making them effectively inaccessible, counteracting availability, and hindering reproducibility.
-Thus, accessibility of research products is a key aspect for the long term improvement
-of reproducibility rates.
-Using an example within evolutionary biology, we identify inclusivity barriers to accessibility,
-% accessibility typically refers to addressing differences in abilities (visual, hearing, cognitive).
-research products in the natural sciences:
-%% list the barriers
-use of highly specialized computing jargon and programming techniques; high cognitive load,
-unspecified, unclear or lengthy goals; content-focused descriptions instead of user-focused;
-inflexible learning environment; unapproachable (cold or intimidating) language;
-and little to no diversity of representation of information.
+However, research products that are digitally available are not necessarily
+friendly for learners and interested parties with little to no %%% different levels of
+experience in the field. This renders research products unapproachable, which
+counteracts availability, and hinders reproducibility short and long term.
+To improve long term adoption of reproducible workflows in research, they
+need to be made approachable for learners, the researchers of the future.
 
-Then, we propose a set of principles to improve accessibility of code and data,
-and illustrate their application in a set of teaching materials for evolutionary biology that we put
-to work in the field.
+Using an example within evolutionary biology, we identify aspects of research workflows
+that make them unapproachable to the general audience:
+use of highly specialized computing jargon and programming techniques;
+high cognitive load;
+unspecified, unclear or lengthy goals;
+content-focused descriptions instead of user-focused;
+inflexible learning environment;
+unapproachable (cold or intimidating) language; and
+little to no diversity of representation of information.
+Then, we propose a set of principles to improve the unapproachable aspects of
+research workflows, and illustrate their application in a case study from
+evolutionary biology that we used as teaching material.
 Finally, we elaborate on the general application of these principles for
-documenting scientific workflows, to provide present learners
+documenting research workflows and products, to provide present learners
 and future researchers with tools for successful scientific reproducibility.
@@
 %% not found among potential users, and keep on increasing.
-
-%% **Introduce OpenTree in abstract??
 
@@
 which has elicited concerns about a crisis in the field \citep{baker2016reproducibility}.
-%%% The scientific community has united to incentivize cultural changes that will improve
-%%% reproducibility rates long term, such as transparency, availability, and workflow
-%%% automatization, to name a few \citep{peng2015reproducibility}.
+
 In response, the scientific community has been developing new principles and standards to incentivize
@@
 \citep{peng2015reproducibility, wilkinson2016fair}.
-A standard for reproducibility that has received much attention is scientific availability, which
+A standard for reproducibility that has received much attention is availability, which
 we define as a property denoting that a research product can be reached (acquired, copied, analyzed,
@@
 and without geographic, demographic, social or temporal barriers for the population \citep{fecher2014open}.
-In this paper, we argue that availability can not be fully achieved without accessibility (Box 1).
 
-We recognize that the term accessibility is used in relation to various aspects of
-the human experience. In relation to digital resources, these aspects have been broadly
-categorized in four axis: perception, operability, understanding and robustness \citep{chisholm2001web, caldwell2008web}
-% (visual, motor, auditory, cognitive, social, psychological, educational, etc.).
-Some forms of accessibility do not support automatization nor scalability.
-For example, Graphical User Interfaces (GUIs) are accessible, but not scalable and rarely automatizable.
+%% In this paper, we argue that availability can not be fully achieved
+%% without approachability (Box 1).
+However, research products that are digitally available are not necessarily
+friendly for learners and interested parties with different levels of
+experience in the field. This renders research products unapproachable, which
+counteracts availability, and hinders reproducibility short and long term.
+To support long term adoption of reproducible practices in research, research
+workflows need to be made approachable for learners, the researchers of the future.
 
-we focus on X types of accessibility
-
-% not all types of accessibility support reproducibility
-% while accessibility is key to support availability, and availability is ket for reproducibility,
-% some forms of accessibility do not support reproducibility.
-
-We identify factors that have particularly affected accessibility in
-the natural sciences, and ways the scientific community can address them to create reproducible
-research workflows. These principles can be generalized and integrated
-into the undergraduate and graduate school STEM curriculum, either for courses specialized
-in reproducibility or within other subject areas, as a necessary component of
-successful and impactful science.
+To elaborate on our thesis, we develop a case study from the research field of
+phylogenetics, a discipline within evolutionary biology.
+This case study allowed us to identify barriers that have made research workflows
+largely unapproachable to a general audience in the natural sciences.
+Then, we propose some principles for researchers to address these barriers, to
+create research workflows that are reproducible by a larger audience.
 
@@
 
-\bigskip
-\bigskip
+The principles proposed here can be generalized and integrated into the undergraduate
+and graduate school STEM curriculum, either for courses specialized in reproducibility
+or within other subject areas, as a necessary component of successful and impactful science.
 
-\noindent\fbox{%
-    \parbox{\textwidth}{%
-    \textbf{Box 1. The role of accessibility in improving reproducibility.}
-      % Availability is defined as the fact or possibility that something can be bought,
-      % used, or reached by someone \citep{availability2022cambridge}.
-      % Accessibility is defined as "the fact of being able to be reached or obtained easily"
-      % \citep{accessibility2022cambridge}
-      Availability and accessibility are conceptually accepted as synonyms for ``the fact
-      of being able to be reached" \citep{available-accessible2022cambridge}.
-      However, accessibility's definition goes further and describes it as the
-      ``fact of being able to be reached \textbf{easily}" \citep{accessibility2022cambridge}.
-      A secondary definition of accessibility elaborates more on this: ``the quality
-      of being \textbf{easy to understand}''.
-      % Yet, in practice, availability does not imply accessibility.
+%% \bigskip
+%% \bigskip
+%%
+%% \noindent\fbox{%
+%%     \parbox{\textwidth}{%
+%%     \textbf{Box 1. The role of accessibility in improving reproducibility.}
+%%       % Availability is defined as the fact or possibility that something can be bought,
+%%       % used, or reached by someone \citep{availability2022cambridge}.
+%%       % Accessibility is defined as "the fact of being able to be reached or obtained easily"
+%%       % \citep{accessibility2022cambridge}
+%%       Availability and accessibility are conceptually accepted as synonyms for ``the fact
+%%       of being able to be reached" \citep{available-accessible2022cambridge}.
+%%       However, accessibility's definition goes further and describes it as the
+%%       ``fact of being able to be reached \textbf{easily}" \citep{accessibility2022cambridge}.
+%%       A secondary definition of accessibility elaborates more on this: ``the quality
+%%       of being \textbf{easy to understand}''.
+%%       % Yet, in practice, availability does not imply accessibility.
+%%
+%%       What we realize from these definitions is that, in practice, something might
+%%       be available but not necessarily accessible,
+%%       hindering the realization of availability.
+%%       To illustrate the practical difference between availability and accessibility we
+%%       will use an allegory that goes like this:
+%%
+%%       ** Change marshmallow story to cake
+%%
+%%       ** Explain different types of barriers to accessibility within the example, then within science
+%%
+%%       ** Explain why we focus on psychological/social/pedagogical barriers
+%%
+%%       ** Gauging accessibility of a resource is tricky because it depends on an
+%%       individual experience of what is in effect \textbf{easy} for someone.
+%%       \textbf{Easy} is a subjective quality that fully depends on individual perceptions
+%%       resulting from individual contexts.
+%%       As a society we are moving towards validating and welcoming all human experiences.
+%%       To improve accessibility levels in the natural sciences, it
+%%       is important to account for the factors that are creating different experiences
+%%       of what is \textbf{easy}.
+%%
+%%       ** Now, explain different types of barriers to accessibility in science
+%%       (Jessica: Maybe say that lack of accessibility can be due to a number of
+%%       factors (physical, resources, economic), but one critical factor is the use
+%%       of impenetrable expert language.)
+%%     }%
+%% }
+%% \bigskip
 
-      What we realize from these definitions is that, in practice, something might
-      be available but not necessarily accessible,
-      hindering the realization of availability.
-      To illustrate the practical difference between availability and accessibility we
-      will use an allegory that goes like this:
+\section*{A case study from phylogenetics}
+\label{sec:case}
 
-      ** Change marshmallow story to cake
-
-      ** Explain different types of barriers to accessibility within the example, then within science
-
-      ** Explain why we focus on psychological/social/pedagogical barriers
-
-      ** Gauging accessibility of a resource is tricky because it depends on an
-      individual experience of what is in effect \textbf{easy} for someone.
-      \textbf{Easy} is a subjective quality that fully depends on individual perceptions
-      resulting from individual contexts.
-      As a society we are moving towards validating and welcoming all human experiences.
-      To improve accessibility levels in the natural sciences, it
-      is important to account for the factors that are creating different experiences
-      of what is \textbf{easy}.
-
-      ** Now, explain different types of barriers to accessibility in science
-      (Jessica: Maybe say that lack of accessibility can be due to a number of
-      factors (physical, resources, economic), but one critical factor is the use
-      of impenetrable expert language.)
-    }%
-}
-\bigskip
-
-\section*{An example from phylogenetics}
-\label{sec:example}
-
-To elaborate on our thesis,
-%% we use a common scientific workflow in phylogenetics,
-%% ** Jessica mentions that workflow is confusing as she thought it referred to the OpenTree
-%% workflow, but we actually mean the more general workflow of reconstructing a phylogenetic tree.
-%% She suggests something more general:
-we choose an example from the research field of phylogenetics,
-a discipline within evolutionary biology. Phylogenetics focuses
-on investigating the history of shared ancestry of living organisms from biological data,
-and represents
-this evolutionary history in a diagram known as a phylogeny (Figure \ref{fig:tree}).
-Phylogenies provide the
-basis to study and understand all biological processes in an evolutionary context
-\citep{dobzhansky1973nothing}. Hence,
-% Given the importance of phylogenies for biological research,
-it appears that improving reproducibility rates
-in phylogenetics has the potential to positively impact research across the natural sciences.
+Phylogenetics is a key discipline within evolutionary biology \citep{dobzhansky1973nothing}.
+It focuses on investigating the history of shared ancestry of living and extinct
+organisms using biological data,
+and represents this evolutionary history with a diagram known as a phylogeny
+or phylogenetic tree (because it grows through time and appears to have branches; Figure \ref{fig:tree}).
+Phylogenies provide the basis to study and understand all biological processes
+in an evolutionary context \citep{dobzhansky1973nothing}
+Hence, it appears that improving reproducibility rates in phylogenetics has the
+potential to positively impact research across the natural sciences.
 
@@
 %% a single phylogenetic tree encompassing all life
-OpenTree's database, the Phylesystem \citep{mctavish2015phylesystem}, provides
-usability of phylogenetic data by standardizing organism names across biological databases,
-and supporting digital data obtention in a computer readable format.
+OpenTree's database, the Phylesystem \citep{mctavish2015phylesystem}, stores
+phylogenetic data in a reusable form by applying a single standard to name organisms
+in the stored phylogenies \citep{rees2017automated}.
+%% It also supports digital data obtention in a computer readable format.
 
-While this is accessible through their Graphical User Interface,
+While this is available through their Graphical User Interface,
 It does not support reproducibility
@@
 % "Thus efforts to increase scientific reproducibility should consider the ubiquity of computers in research."
-% In order for reproducible computational tools to be adopted for research, they
-% need to be readily accessible for present and future researchers.
+
 Thus, efforts to increase reproducibility rates in the natural sciences must consider

### 2022-02-14 jsdse/Manuscript.tex
case study

@@
 
-%% In this paper, we argue that availability can not be fully achieved
-%% without approachability (Box 1).
-However, research products that are digitally available are not necessarily
+In this paper, we argue that research products that are digitally available are
+often unapproachable in practice (Box 1), because they are not
 friendly for learners and interested parties with different levels of
-experience in the field. This renders research products unapproachable, which
-counteracts availability, and hinders reproducibility short and long term.
-To support long term adoption of reproducible practices in research, research
+experience in the field. Research products that are unapproachable
+counteract availability, and hinder reproducibility short and long term.
+To support long term adoption of reproducible practices in the natural sciences, research
 workflows need to be made approachable for learners, the researchers of the future.
 
-To elaborate on our thesis, we develop a case study from the research field of
+To elaborate on our thesis, we designed a case study from the research field of
 phylogenetics, a discipline within evolutionary biology.
-This case study allowed us to identify barriers that have made research workflows
+We use our case study to identify barriers that have made research workflows
 largely unapproachable to a general audience in the natural sciences.
-Then, we propose some principles for researchers to address these barriers, to
+Then, we propose some principles for researchers to address these barriers, and
 create research workflows that are reproducible by a larger audience.
-
 % ** describe/introduce reproducibility curriculum here
-
 The principles proposed here can be generalized and integrated into the undergraduate
@@
 %%     \parbox{\textwidth}{%
-%%     \textbf{Box 1. The role of accessibility in improving reproducibility.}
+%%     \textbf{Box 1. The role of approachable research workflows in teacing reproducibility.}
 %%       % Availability is defined as the fact or possibility that something can be bought,
@@
 %%       % \citep{accessibility2022cambridge}
+%%       % Yet, in practice, availability does not imply accessibility.
 %%       Availability and accessibility are conceptually accepted as synonyms for ``the fact
@@
 %%       of being \textbf{easy to understand}''.
-%%       % Yet, in practice, availability does not imply accessibility.
 %%
@@
 and represents this evolutionary history with a diagram known as a phylogeny
-or phylogenetic tree (because it grows through time and appears to have branches; Figure \ref{fig:tree}).
+or phylogenetic tree (because it grows through time and appears to have branches;
+Figure \ref{fig:tree}).
 Phylogenies provide the basis to study and understand all biological processes
@@
 
-***What other platforms provide resources for reproducibility in phylogenetics???
+To explore barriers to approachability in phylogenetics, we develop a study case
+that touches on three common problems within the field: standardizing
+organism names in phylogenies, obtaining current phylogenetic knowledge for a group of organisms,
+and summarizing this phylogenetic knowledge in a meaningful way.
+To address these problems, we propose a research workflow that relies on resources from the Open
+Tree of Life (OpenTree), an open source project that provides
+digital availability of phylogenetic results from published, peer-reviewed research, which
+is considered as vetted and state-of-the-art knowledge in the field.
+OpenTree phylogenies are stored in a public database, the Phylesystem \citep{mctavish2015phylesystem},
+and are downloadable as various computer-readable file types, which is key for reusability
+and reproducible workflows \citep{CITATION}.
+OpenTree also provides access to a single naming standard for organisms (taxonomic standard)
+that is applied to the stored phylogenies \citep{rees2017automated}, which are
+then used to summarize a single phylogenetic tree encompassing all life \citep{opentreeoflife2019synth}.
 
-The Open Tree of Life project (OpenTree) has developed a computational platform that provides
-digital availability of phylogenetic results from vetted, state-of-the-art research
-in the field \citep{opentreeoflife2019synth}.
-%% by creating a standardized database of phylogenetic data that is used to synthesize
-%% a single phylogenetic tree encompassing all life
-OpenTree's database, the Phylesystem \citep{mctavish2015phylesystem}, stores
-phylogenetic data in a reusable form by applying a single standard to name organisms
-in the stored phylogenies \citep{rees2017automated}.
-%% It also supports digital data obtention in a computer readable format.
+All of OpenTree resources
+%% (the stored phylogenies, the taxonomic standard, the synthetic tree, and all
+%% tools developed to process and summarize stored phylogenies),
+are free of cost to any user, and are available for download and use through its Graphical User
+Interface (aka, a website or application that allows users to access computer functionalities, in this case OpenTree resources, with mouse or keyboard clicks).
+However, reducing as many manual steps as possible in research workflows is key
+for reproducibility, as manual data manipulation scales poorly and is prone to
+error \citep{bakken2019journey}.
+OpenTree's resources are also programmatically available through its Application
+Programming Interface services (APIs; aka computer code that implements computer functionalities,
+in this case OpenTree resources, that can be used by programmers
+to build more functionalities),
+which provide scalability and reproducibility \citep{opentreeAPIv3}.
+However, this comes at a high cost for the user, which requires considerable more
+experience and literacy with computer programming to be able to successfully use APIs.
+%% In recent years, OpenTree API services have been wrapped into software packages for R and Python
+%% \citep{michonneau2016rotl, mctavish2021opentree}, open source, free of cost
+%% programming languages that represent two of the most widely used programming languages
+%% in the sciences \citep{baker2017scientific}. These packages should contribute to making
+%% OpenTree resources more accessible to a wider user audience.
+The R and Python programming languages are open source and free of cost, and represent
+two of the most widely used programming languages in the sciences \citep{baker2017scientific}.
+The rotl R package \citep{michonneau2016rotl} and the OpenTree Python module
+\citep{mctavish2021opentree} are now available as wrappers for OpenTree's API services.
+As such, these packages should contribute to making OpenTree resources more accessible
+to a wider programming user audience.
 
-While this is available through their Graphical User Interface,
-It does not support reproducibility
-
-Accessibility does not imply reproducibility
-
-In order for something to be highly reproducible, it has to be largely accessible.
-
-is programmatically
-available free of cost to users through OpenTree's Application Programming Interface
-(API) services \citep{opentreeAPIv3}.
-Additionally, software packages written with open source, free of cost programming
-languages commonly used in scientific research such as R and Python
-\citep{baker2017scientific} have been developed as wrappers that have increased the availability
-of OpenTree's API services, by making the OpenTree API functionalities more accessible
-to a wider user audience \citep{michonneau2016rotl, mctavish2021opentree}.
-Yet, the R and Python OpenTree API wrapper software packages have been mainly used
-by computer-literate individuals
-to seamlessly establish reproducible workflows to use and reuse expert phylogenetic
-knowledge for biological research \citep{sanchez2019datelife, sanchez2021physcraper}
-and education \citep{nguyen2020phylotastic, phylotasticedtools, galacticedtools}.
-In the 5 years since its release, the R package wrapper for OpenTree \texttt{rotl}
-has been cited by more than 190 biological research papers addressing various
-topics \citep{googlerotl}.
-Comparatively, other R packages for phylogenetics, such as \texttt{GGTREE}
-and \texttt{mixOmics}, which were released one year after \texttt{rotl} have been
-cited over a thousand times, which might indicate a delay \texttt{rotl}
-adoption. Of course one potential explanation is that OpenTree services do not have
-as wide an application as services provided by the afore mentioned R packages.
-However, the
-expressed enthusiasm (interest to use/inability to use the resource) by the target
- audience and larger community might point to a different issue.
-%% We need a proper statistical distribution of citation numbers per year, to see
-%% if rotl is an outlier or not.
-
-While learners in the natural sciences have been engaging independently with R
-and Python programming languages
-as they represent two of the most widely used programming languages in the sciences
-% for their fast and seamless statistics and data analysis environment
-\citep{baker2017scientific}, computer programming is not traditionally
+However, while learners in the natural sciences have been engaging independently
+with R and Python programming languages, computer programming is not traditionally
 a core skill formally taught to biologists and naturalists
 \citep{sayres2018bioinformatics, wright2019the, williams2019barriers}.
-Moreover, the main usage of the OpenTree API services hints that tools intended to
-improve availability and reproducibility come at a cost for the potential users:
- an increased required baseline computational knowledge and skillset
-in the field.
 As computers continue to play a larger role in most scientific disciplines \citep{piccolo2016tools},
 higher baseline computational skills are required across all natural sciences.
-% "Thus efforts to increase scientific reproducibility should consider the ubiquity of computers in research."
+Thus, efforts to increase reproducibility rates in the natural sciences must consider
+the specifics of scientific workflows that rely on usage of computer tools and
+programming languages \citep{peng2011reproducible, sandve2013ten, powers2019open}.
 
-Thus, efforts to increase reproducibility rates in the natural sciences must consider
-the specifics of scientific workflows that rely on usage of programming languages, such as
-availability and accessibility of data and code \citep{peng2011reproducible, sandve2013ten, powers2019open}.
-% However, the utility of data resources is limited by the technical challenges of accessing the data.
-% In order to motivate reproducible research, that gap needs to be bridged.
-% In this work we focus on improving accessibility of code examples and documentation.
-
-We chose a phylogenetics workflow that relies on data and code from OpenTree to
-identify the specific barriers to accessibility it presents.
-Then, we design ways to overcome or diminish these barriers and apply them to
-a series of tutorials and vignettes developed for the OpenTree project.
-The tutorials are available at https://mctavishlab.github.io/R\_OpenTree\_tutorials/.
-
-Finally, we generalize our findings to suggest a set of
-principles that can be used as a guide to develop code and documentation materials
-with increased accessibility, to improve availability and contribute to reproducibility.
-Notably, these principles can be incorporated as learning goals into any syllabus for a
-course or workshop on best practices for scientific reproducibility.
-% In particular, we identify the necessity for documentation that is written down
-% using language that is common to the target audience to facilitate examination,
-% application, and adoption of code by the wider audience.
+In the next section que describe the barriers to approachability that we identified
+on our case study. We addressed these barriers on a teaching material that is
+available at
+https://mctavishlab.github.io/R\_OpenTree\_tutorials/.
 
@@
 \end{center}
-\caption{A phylogenetic tree from our tutorial website, extracted using OpenTree of Life tools and the rotl R package. \label{fig:tree}}
+\caption{A phylogenetic tree from our tutorial.
+It was extracted using OpenTree of Life resources \citep{opentreeoflife2019synth} wrapped in the \texttt{rotl} R package \citep{michonneau2016rotl}.
+\label{fig:tree}}
 \end{figure}
@@
 
-Vignettes and tutorials work as secondary pieces of documentation, that help to demonstrate
-additional cases of individual function usage, and showcase function associations
-that work for specific analysis workflows in more detail.
-As secondary documentation has become more common practice and is more flexible
-in its form and content, it constitutes an ideal canvas to develop, implement and
-test principles that can overcome current barriers to code accessibility.
+%% Vignettes and tutorials work as secondary pieces of documentation, that help to demonstrate
+%% additional cases of individual function usage, and showcase function associations
+%% that work for specific analysis workflows in more detail.
+%% As secondary documentation has become more common practice and is more flexible
+%% in its form and content, it constitutes an ideal canvas to develop, implement and
+%% test principles that can overcome current barriers to code accessibility.
+
+- programming experience/ skills
+- specialized jargon knowledge
+- use of highly specialized computing jargon and programming techniques;
+
+Some principles of an inclusive syllabus:
+- high cognitive load,
+- unspecified, unclear or lengthy goals;
+- content-focused descriptions instead of user-focused;
+- inflexible learning environment;
+- unapproachable (cold or intimidating) language;
+- and little to no diversity of representation of information
 
@@
 
-%% I think I have three different points here:
-%% internalizing knowledge is made easier with active learning practices: learn by doing!
-%% literate programming allows learning by doing AND at the same time provides a story that connects the acquired knowledge to a real worl problem.
-%% By definition, all languages are abstract. We learn them because we grow up with them and they have a direct application for our lives.
-%% Programming languages are no different, however, we start learning those later in life and we do not use them as regularly, making it harder for us to learn them and use them.
+% approachable practice: reduced cognitive load and specific and clear goals
+
 Pedagogical research shows that active learning practices are one the most effective
 ways to take on abstract subjects \citep{freeman2014active}.
-Programming computer languages are quite abstract and learning them can be greatly
-enhanced by applying an active learning strategy such as linking its usage to
+Programming computer languages are quite abstract and cognitive load can be greatly
+reduced for learners by applying an active learning strategy such as linking its usage to
 a ``real world'' or ``human'' application \citep{felder2009active}.
@@
 accessible by choosing narratives that are relatable to a more general audience.
-An easy way to do this in biology is choosing as model organism a taxon that a
+An easy way to do this in biology is choosing as model organism a charismatic taxon that a
 research group is more interested in studying. For the non-specialized user base,
@@
 \subsubsection*{b. Demonstrate errors and warnings thoroughly}
+
+% approachable practice: user focused examples (instead of content focus)
 
@@
 
+% approachable best practice: friendly and relatable language
+
 Besides avoiding formal language, and incorporating elements of pop culture, such as picture
@@
 
-Response form the community has been invaluable in gauging success of the tutorials.
+Response form the community has been invaluable in gauging success of our teaching
+materials.
 Senior researchers often comment on the usefulness
-of the tutorials for their research, as well as supporting students in using the
+of the tutorials for their research, as well as how they have supported students in using the
 R packages with less help from them as PIs.
@@
 
-- save explanation/training time when analyses are run again by students and collaborators;
-- save research time for yourself when analyses are run again with more data, a
-  different dataset, a different organism or biological model;
-- scientific efforts can build off of each other.
+* save explanation/training time when analyses are run again by students and collaborators;
+* save research time for yourself when analyses are run again with more data, a* different dataset, a different organism or biological model;
+* scientific efforts can build off of each other.
 
@@
 economic resources, often occurring in households from underrepresented communities
-and minorities \citep{REFERENCE-NEEDED}.
+and minorities \citep{google2016diversity, warner2021quantifying}.
 It can also be due to gender-biased parental and community pressures,
 in which male individuals are more often encouraged to perform activities related to computers,
-while female individuals are discouraged \citep{REFERENCE-NEEDED}.
+while female individuals are discouraged \citep{warner2021quantifying}.
 % How to balance software acceptance VS. adoption?

### 2022-02-14 jsdse/Manuscript.tex
barriers

@@
 In the next section que describe the barriers to approachability that we identified
-on our case study. We addressed these barriers on a teaching material that is
+on our case study. We addressed these barriers on teaching materials that are
 available at
@@
 
-\section*{Identifying hurdles to accessibility}
+\section*{Identifying barriers to approchable research workflows}
 \label{sec:identifying}
+
+Cite carpentries and Mosi
 
@@
 
-- programming experience/ skills
-- specialized jargon knowledge
-- use of highly specialized computing jargon and programming techniques;
+Some principles of an inclusive syllabus:
 
-Some principles of an inclusive syllabus:
-- high cognitive load,
-- unspecified, unclear or lengthy goals;
-- content-focused descriptions instead of user-focused;
-- inflexible learning environment;
-- unapproachable (cold or intimidating) language;
-- and little to no diversity of representation of information
+- high cognitive load/ skill level; (literate programming)
 
-\section*{Addressing hurdles to accessibility: some principles}
+- unspecified, unclear or lengthy goals;(literate programming)
+
+- content-focused descriptions instead of user-focused; (demonstrate errors)
+
+- inflexible learning environment; (make it stable so users can learn whenever is best for them)
+
+- unapproachable (cold or intimidating) language, use of highly specialized computing jargon; (use friendly and informal but respectful language)
+
+- and little to no diversity of representation of information (say the same thing with different images, metaphors, stories)
+
+\section*{Best practices for approachable research workflows}
 \label{sec:addressing}

### 2022-02-14 jsdse/Manuscript.tex
adding citations

@@
 of science, as it constitutes a minimum standard that allows understanding research products
-(e.g., methods, data, analysis, results, etc.)
+(e.g., methods, data, analysis, results, etc.; \cite{piwowar2013value})
 to determine their reliability and generality, and eventually build up scientific
@@
 In response, the scientific community has been developing new principles and standards to incentivize
-cultural changes in an effort to improve long term reproducibility rates in the natural sciences
-\citep{peng2015reproducibility, wilkinson2016fair}.
+cultural changes that support a long term improvement of reproducibility rates in the natural sciences
+\citep{peng2015reproducibility, wilkinson2016fair, miyakawa2020no}.
 A standard for reproducibility that has received much attention is availability, which
@@
 
-To elaborate on our thesis, we designed a case study from the research field of
+To elaborate on our thesis, we designed a case study within the research field of
 phylogenetics, a discipline within evolutionary biology.
@@
 
-To explore barriers to approachability in phylogenetics, we develop a study case
+To explore barriers to approachability in phylogenetics, we develop a case study
 that touches on three common problems within the field: standardizing
@@
 and are downloadable as various computer-readable file types, which is key for reusability
-and reproducible workflows \citep{CITATION}.
+and reproducible workflows \citep{wilson2017good}.
 OpenTree also provides access to a single naming standard for organisms (taxonomic standard)
@@
 However, this comes at a high cost for the user, which requires considerable more
-experience and literacy with computer programming to be able to successfully use APIs.
+computer programming experience and literacy to be able to successfully use APIs.
 %% In recent years, OpenTree API services have been wrapped into software packages for R and Python
@@
 The R and Python programming languages are open source and free of cost, and represent
-two of the most widely used programming languages in the sciences \citep{baker2017scientific}.
+two of the most widely used programming languages in the sciences today \citep{baker2017scientific}.
 The rotl R package \citep{michonneau2016rotl} and the OpenTree Python module
@@
 In the next section que describe the barriers to approachability that we identified
-on our case study. We addressed these barriers on teaching materials that are
+on our case study. We address these barriers by developing teaching materials that are
 available at
@@
 
-\section*{Identifying barriers to approchable research workflows}
+\section*{Identifying barriers to approachable research workflows}
 \label{sec:identifying}

### 2022-02-15 jsdse/Manuscript.tex
more on barriers

@@
 that make them unapproachable to the general audience:
-use of highly specialized computing jargon and programming techniques;
+use of specialized computing language and programming techniques;
 high cognitive load;
@@
 is a key aspect of the advancement
-of science, as it constitutes a minimum standard that allows understanding research products
-(e.g., methods, data, analysis, results, etc.; \cite{piwowar2013value})
+of science, as it constitutes a minimum standard that allows understanding research products,
+i.e., methods, data, analysis, results, etc. \citep{piwowar2013value},
 to determine their reliability and generality, and eventually build up scientific
@@
 
+\bigskip
+\bigskip
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=3in]{fig-tree.png}
+\end{center}
+\caption{A phylogenetic tree from our tutorial.
+It was extracted using OpenTree of Life resources \citep{opentreeoflife2019synth} wrapped in the \texttt{rotl} R package \citep{michonneau2016rotl}.
+\label{fig:tree}}
+\end{figure}
+
 \section*{A case study from phylogenetics}
@@
 are free of cost to any user, and are available for download and use through its Graphical User
-Interface (aka, a website or application that allows users to access computer functionalities, in this case OpenTree resources, with mouse or keyboard clicks).
+Interface (GUI; aka, a website or application that allows users to access computer functionalities, in this case OpenTree resources, with mouse or keyboard clicks).
 However, reducing as many manual steps as possible in research workflows is key
@@
 OpenTree's resources are also programmatically available through its Application
-Programming Interface services (APIs; aka computer code that implements computer functionalities,
+Programming Interface services (APIs; aka, computer code that implements computer functionalities,
 in this case OpenTree resources, that can be used by programmers
@@
 %% OpenTree resources more accessible to a wider user audience.
-The R and Python programming languages are open source and free of cost, and represent
-two of the most widely used programming languages in the sciences today \citep{baker2017scientific}.
-The rotl R package \citep{michonneau2016rotl} and the OpenTree Python module
-\citep{mctavish2021opentree} are now available as wrappers for OpenTree's API services.
-As such, these packages should contribute to making OpenTree resources more accessible
-to a wider programming user audience.
+The \texttt{rotl} R package \citep{michonneau2016rotl} and the \texttt{opentree} Python module
+\citep{mctavish2021opentree} have been developed as wrappers for OpenTree's API services.
+R and Python are open source and free of cost programming languages that represent
+two of the most widely used programming languages in the sciences today \citep{eglen2009quick, baker2017scientific}.
+As such, \texttt{rotl} and \texttt{opentree} software packages should contribute
+to making OpenTree resources more accessible to a wider programming user audience.
 
@@
 As computers continue to play a larger role in most scientific disciplines \citep{piccolo2016tools},
-higher baseline computational skills are required across all natural sciences.
-Thus, efforts to increase reproducibility rates in the natural sciences must consider
-the specifics of scientific workflows that rely on usage of computer tools and
-programming languages \citep{peng2011reproducible, sandve2013ten, powers2019open}.
+higher baseline computational skills are required across all natural sciences not only
+to develop an original research workflow, but to be able to follow and reproduce
+research workflows from other researchers.
 
-In the next section que describe the barriers to approachability that we identified
-on our case study. We address these barriers by developing teaching materials that are
+Thus, efforts to increase reproducibility rates long term in the natural sciences
+would benefit from addressing specific barriers for learners in the field, to
+support them in acquiring the skills needed to reproduce research workflows
+that rely heavily on computer code \citep{peng2011reproducible, sandve2013ten, powers2019open}.
+
+In the next section, we describe the barriers to approachable research workflows that we identified
+on our case study. We address these barriers in a set of teaching materials that are
 available at
 https://mctavishlab.github.io/R\_OpenTree\_tutorials/.
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=3in]{fig-tree.png}
-\end{center}
-\caption{A phylogenetic tree from our tutorial.
-It was extracted using OpenTree of Life resources \citep{opentreeoflife2019synth} wrapped in the \texttt{rotl} R package \citep{michonneau2016rotl}.
-\label{fig:tree}}
-\end{figure}
 
@@
 
-Cite carpentries and Mosi
+% Cite carpentries and Mosi
 
-Good primary documentation for code is thorough. It describes general usage of individual functions,
-the components and variables a function can take, and it should be accompanied with
-function usage examples on how to apply it \citep{karimzadeh2018top}.
-As opposed to code, primary documentation is written in natural language (i.e.,
-any known human language, e.g., English, Spanish, Chinese).
-Primary documentation is viewed as a key element for success of a piece of code
-\citep{karimzadeh2018top},
-which might be why it is also usually written using highly specialized computational jargon (i.e., computationally specific concepts,
-words, and phrases) as well as formal scientific language.
-While this might be important for formal acceptance of the code by the scientific
-and academic community, it often slows down or even
-obstructs examination, application, and adoption of code by the general audience \citep{ball2017its}.
+The main goal of our research workflow is to obtain a single phylogeny summarizing
+data from a set of published phylogenies for the canids (the family of dogs, coyotes, wolves, etc.),
+our organism of study.
+Our analysis can be completely accomplished using functions
+from the R package \texttt{rotl} or the Python module \texttt{opentree}. If
+a researcher were to use this workflow in a publication, they would typically
+describe it in the methodology section as ``The canid summary phylogeny was obtained using
+functions from X package, details are available as supplementary files".
+This is usual practice, mainly because journals do not have space to publish
+all code used in an analysis in the methods section. Yet, supplementary files have
+the misfortune to not be peer-reviewed as thoroughly (or at all) as the main manuscript
+\citep{CITATION}.
+They are also prone to the dreaded promise ``available upon request", which has
+very low rates of fulfillment \citep{CITATION}.
+Without the primary code that was used to perform an analysis it is impossible
+to reproduce said analysis. When the code is available, other issues can
+complicate reproduction of the analysis to the point of completely obstructing
+reproducibility.
+For example, what software do I need to open the code files? Can I run the code within the same
+software, or do I need a different one?
+Do I need additional software that the analysis depends on? What does the code even mean?
 
-%% Vignettes and tutorials work as secondary pieces of documentation, that help to demonstrate
-%% additional cases of individual function usage, and showcase function associations
-%% that work for specific analysis workflows in more detail.
-%% As secondary documentation has become more common practice and is more flexible
-%% in its form and content, it constitutes an ideal canvas to develop, implement and
-%% test principles that can overcome current barriers to code accessibility.
+These questions are usually addressed in the software documentation.
+As opposed to code, documentation is written in natural language (i.e.,
+any known human language, e.g., English, Spanish, Chinese), and is considered
+a key element for successful adoption of software
+by the target users \citep{karimzadeh2018top}. This might explain
+why documentation for software addressed to academic users is also
+usually written using highly specialized computational
+language or jargon (i.e., computationally specific concepts,
+words, and phrases) as well as formal scientific and academic language.
+\textit{Barrier 1.--} While scientific jargon might have an important role for formal acceptance
+of software by the scientific and academic community, it can be perceived as
+cold and/or intimidating language, that often slows down or even
+obstructs examination, application, and adoption of code by a wider audience \citep{ball2017its},
+and discourages learners by creating a hostile environment that does not foster learning \citep{CITATION}.
 
-Some principles of an inclusive syllabus:
+Good software documentation for code has to be thorough \citep{karimzadeh2018top}.
+It should describe general usage of individual functions, as well as
+the argument and variables said function can take, and it should be accompanied with
+usage examples on how to apply each function \citep{karimzadeh2018top}.
+Individual documentation for each function is usually presented
+in alphabetic order, and does not have a specific goal, besides being as thorough as possible.
+In this context, identifying connections across functions that are meant to work on the same
+analysis workflow is difficult. Moreover, most software has many functions, so documentation
+is usually very lengthy. \textit{Barrier 2.--}  All of this increases the cognitive
+load for the target users and learners, which discourages learners \citep{CITATION}.
 
-- high cognitive load/ skill level; (literate programming)
+Another important aspect of software documentation is that examples are usually
+worked and showcase the ideal or minimal case in which a function should work well.
+\textit{Barrier 3.--} Focusing importance on the content of the software instead
+of considering the user experience, ignoring what can go wrong and avoiding clear
+advice on how to troubleshoot.
 
-- unspecified, unclear or lengthy goals;(literate programming)
+Software documentation is usually explained with words, very rarely using diagrams or
+other allegories.
+\textit{Barrier 4.--} Little to no diversity of representation of information
 
-- content-focused descriptions instead of user-focused; (demonstrate errors)
+Documentation should always be available for the users and learners
+\textit{Barrier 5.--} Inflexible learning environment, is this one really an issue???
 
-- inflexible learning environment; (make it stable so users can learn whenever is best for them)
+% Some principles of an inclusive syllabus:
 
-- unapproachable (cold or intimidating) language, use of highly specialized computing jargon; (use friendly and informal but respectful language)
+% - unspecified, unclear or lengthy goals; high cognitive load; (literate programming)
 
-- and little to no diversity of representation of information (say the same thing with different images, metaphors, stories)
+% - content-focused descriptions/examples instead of user-focused; (demonstrate errors)
+
+% - inflexible learning environment; (make it stable so users can learn whenever is best for them)
+
+% - unapproachable (cold or intimidating) language, use of highly specialized computing jargon, skill level; (use friendly and informal but respectful language)
+
+% - and little to no diversity of representation of information (say the same thing with different images, metaphors, stories)
+
+In sum, best practices for good primary documentation are not enough to ensure
+reproducibility of research workflows that rely heavily on code.
 
@@
 
-\subsubsection*{a. Literate programming: Demonstrate code usage with integrative examples}
-
-% approachable practice: reduced cognitive load and specific and clear goals
+\subsubsection*{a. Reduce cognitive load and provide specific and clear goals
+with literate programming}
 
@@
 
-A story-like narrative that links pieces of code together and invites learners
-to try the code, can lead learners to remember what they are doing and
+A story-like narrative that links code usage in an integrative example, invites learners
+to try the code, which can lead them to remember what they are doing and
 why they are doing it.
@@
 accessible by choosing narratives that are relatable to a more general audience.
-An easy way to do this in biology is choosing as model organism a charismatic taxon that a
-research group is more interested in studying. For the non-specialized user base,
+An easy way to do this in biology is choosing a charismatic taxon as model organism.
+For a research group this can be the group they are studying. For the general audience,
 a highly charismatic group such as dinosaurs will do the trick.
@@
 
-\subsubsection*{b. Demonstrate errors and warnings thoroughly}
+\subsubsection*{b. Provide examples that are user-focused by demonstrating errors and warnings}
 
 % approachable practice: user focused examples (instead of content focus)
+% demonstrate errors and warnings thoroughly
 
@@
 
-\subsubsection*{c. Avoid jargon and expert language}
+\subsubsection*{c. Use friendly, relatable and respectful language}
 
-% approachable best practice: friendly and relatable language
+% avoid specialized, intimidating and cold language
 
@@
 computational concepts that were assumed or were not explained in depth.
-We vetted the tutorials through feedback from workshop participants as well as individual users.
-We choose examples that are charismatic for the audience.
-For example, when we presented the tutorial for a team specialized in Amphibians,
-we tailored the examples using frogs and their allies.
+We vetted the tutorials through feedback from workshop participants as well as
+individual users to identify such specialized concepts.
 
-\subsubsection*{d. Make it stable through time}
+\subsubsection*{d. Make it accessible geographically and through time}
+
+% - inflexible learning environment; (make it stable so users can learn whenever is best for them)
 

### 2022-02-16 jsdse/Manuscript.tex
almost there

@@
   \author{Luna L. Sanchez Reyes\thanks{
-    The authors gratefully acknowledge ``Sustaining the Open Tree of Life``, NSF ABI No. 1759838, and ABI No. 1759846.}\hspace{.2cm}\\
+    The authors gratefully acknowledge ``Sustaining the Open Tree of Life'',
+    NSF ABI No. 1759838, and ABI No. 1759846. They also acknowledge ``The Carpentries''
+    project, without it this project could not have been possible}\hspace{.2cm}\\
     School of Natural Sciences, University of California, Merced\\
@@
 
-Using an example within evolutionary biology, we identify aspects of research workflows
+Using a case study within evolutionary biology, we identify aspects of research workflows
 that make them unapproachable to the general audience:
-use of specialized computing language and programming techniques;
-high cognitive load;
-unspecified, unclear or lengthy goals;
-content-focused descriptions instead of user-focused;
-inflexible learning environment;
-unapproachable (cold or intimidating) language; and
+use of highly specialized language is intimidating;
+unspecified, unclear or lengthy goals lead to high cognitive load;
+content-focused descriptions instead of user-focused; and
 little to no diversity of representation of information.
@@
 \noindent%
-{\it Keywords:}  open science, R, phylogenetics, Open Tree of Life
+{\it Keywords:}  open science, R, phylogenetics, Open Tree of Life, pedagogy
 \vfill
@@
 In this paper, we argue that research products that are digitally available are
-often unapproachable in practice (Box 1), because they are not
+often unapproachable in practice, because they are not
 friendly for learners and interested parties with different levels of
@@
 To support long term adoption of reproducible practices in the natural sciences, research
-workflows need to be made approachable for learners, the researchers of the future.
+workflows need to be made approachable for learners, the researchers of the future \citep{roland2002think}.
 
@@
 
-\bigskip
-\bigskip
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=3in]{fig-tree.png}
-\end{center}
-\caption{A phylogenetic tree from our tutorial.
-It was extracted using OpenTree of Life resources \citep{opentreeoflife2019synth} wrapped in the \texttt{rotl} R package \citep{michonneau2016rotl}.
-\label{fig:tree}}
-\end{figure}
-
 \section*{A case study from phylogenetics}
@@
 Phylogenies provide the basis to study and understand all biological processes
-in an evolutionary context \citep{dobzhansky1973nothing}
+in an evolutionary context \citep{dobzhansky1973nothing}.
 Hence, it appears that improving reproducibility rates in phylogenetics has the
@@
 
-To explore barriers to approachability in phylogenetics, we develop a case study
+To explore barriers to approachable phylogenetics, we develop a case study
 that touches on three common problems within the field: standardizing
@@
 and reproducible workflows \citep{wilson2017good}.
-OpenTree also provides access to a single naming standard for organisms (taxonomic standard)
+OpenTree also provides access to a single standard for organism names (taxonomic standard)
 that is applied to the stored phylogenies \citep{rees2017automated}, which are
@@
 
-All of OpenTree resources
+All of these resources
 %% (the stored phylogenies, the taxonomic standard, the synthetic tree, and all
 %% tools developed to process and summarize stored phylogenies),
-are free of cost to any user, and are available for download and use through its Graphical User
-Interface (GUI; aka, a website or application that allows users to access computer functionalities, in this case OpenTree resources, with mouse or keyboard clicks).
+are available for download and use from OpenTree free of cost to any user.
+One way to access OpenTree resources is through it Graphical User
+Interface (GUI; aka, a website or application that allows users to access computer
+functionalities, in this case OpenTree resources, with mouse or keyboard clicks).
 However, reducing as many manual steps as possible in research workflows is key
@@
 The \texttt{rotl} R package \citep{michonneau2016rotl} and the \texttt{opentree} Python module
-\citep{mctavish2021opentree} have been developed as wrappers for OpenTree's API services.
-R and Python are open source and free of cost programming languages that represent
-two of the most widely used programming languages in the sciences today \citep{eglen2009quick, baker2017scientific}.
+\citep{mctavish2021opentree} have been developed to wrap OpenTree's API services
+and make them available through the R and Python programming languages.
+R and Python are open source and free of cost, and represent
+two of the most widely used programming languages in the sciences today \citep{eglen2009quick,
+baker2017scientific}.
 As such, \texttt{rotl} and \texttt{opentree} software packages should contribute
@@
 
-In the next section, we describe the barriers to approachable research workflows that we identified
-on our case study. We address these barriers in a set of teaching materials that are
-available at
+In the next section, we describe (in no particular order) the barriers to approachable research
+workflows that we identified on our case study.
+Then we develop a set of principles to address these barriers, and apply the latter
+to a set of teaching materials that are available at
 https://mctavishlab.github.io/R\_OpenTree\_tutorials/.
@@
 
-The main goal of our research workflow is to obtain a single phylogeny summarizing
+The main goal of our case study is to obtain a single phylogeny summarizing
 data from a set of published phylogenies for the canids (the family of dogs, coyotes, wolves, etc.),
 our organism of study.
-Our analysis can be completely accomplished using functions
+All analysis for our case study can be completely accomplished using functions
 from the R package \texttt{rotl} or the Python module \texttt{opentree}. If
-a researcher were to use this workflow in a publication, they would typically
+a researcher were to use the proposed analysis workflow in a publication, they would typically
 describe it in the methodology section as ``The canid summary phylogeny was obtained using
@@
 This is usual practice, mainly because journals do not have space to publish
-all code used in an analysis in the methods section. Yet, supplementary files have
+all code used for an analysis in the methods section. Yet, supplementary files and data have
 the misfortune to not be peer-reviewed as thoroughly (or at all) as the main manuscript
-\citep{CITATION}.
+\citep{pop2015use}.
 They are also prone to the dreaded promise ``available upon request", which has
-very low rates of fulfillment \citep{CITATION}.
-Without the primary code that was used to perform an analysis it is impossible
-to reproduce said analysis. When the code is available, other issues can
-complicate reproduction of the analysis to the point of completely obstructing
+very low rates of fulfillment \citep{krawczyk2012available}.
+Without the primary data and code that was used to perform an analysis it is impossible
+to reproduce said analysis \citep{miyakawa2020no}. When the code is available, other issues can
+complicate reproduction of the analysis, to the point of completely obstructing
 reproducibility.
-For example, what software do I need to open the code files? Can I run the code within the same
-software, or do I need a different one?
-Do I need additional software that the analysis depends on? What does the code even mean?
+For example, some questions that are often unaddressed in research workflows are:
+what software do I need to open the code files?
+Can I run the code with the same software I used to open the files, or do I need a different one?
+Do I need additional software that the analysis depends on?
+What does the code even mean?
 
-These questions are usually addressed in the software documentation.
-As opposed to code, documentation is written in natural language (i.e.,
+These questions can usually be answered by referring to the software documentation,
+which are usually publicly available and can be accessed by any potential user.
+As opposed to code, software documentation is written in natural language (i.e.,
 any known human language, e.g., English, Spanish, Chinese), and is considered
 a key element for successful adoption of software
-by the target users \citep{karimzadeh2018top}. This might explain
+by target users \citep{karimzadeh2018top}. This might explain
 why documentation for software addressed to academic users is also
@@
 words, and phrases) as well as formal scientific and academic language.
-\textit{Barrier 1.--} While scientific jargon might have an important role for formal acceptance
+We identify this as \textit{barrier 1} to approachable research workflows.--
+\textit{Specialized language is intimidating}.
+While scientific jargon might have an important role for formal acceptance
 of software by the scientific and academic community, it can be perceived as
 cold and/or intimidating language, that often slows down or even
-obstructs examination, application, and adoption of code by a wider audience \citep{ball2017its},
-and discourages learners by creating a hostile environment that does not foster learning \citep{CITATION}.
+obstructs examination, application, and adoption of code by a wider audience
+\citep{ball2017its}.
+In contrast, introducing information without the use of jargon, supports learners'
+conceptual understanding of new ideas and concepts \citep{mcdonnell2016concepts, pan2019online}.
+% jargon creates a hostile environment that does not foster learning \citep{CITATION}.
 
-Good software documentation for code has to be thorough \citep{karimzadeh2018top}.
-It should describe general usage of individual functions, as well as
-the argument and variables said function can take, and it should be accompanied with
-usage examples on how to apply each function \citep{karimzadeh2018top}.
-Individual documentation for each function is usually presented
-in alphabetic order, and does not have a specific goal, besides being as thorough as possible.
-In this context, identifying connections across functions that are meant to work on the same
-analysis workflow is difficult. Moreover, most software has many functions, so documentation
-is usually very lengthy. \textit{Barrier 2.--}  All of this increases the cognitive
-load for the target users and learners, which discourages learners \citep{CITATION}.
+Another element of good software documentation is that it has to be thorough \citep{karimzadeh2018top}.
+Meaning that it should describe general usage of individual functions, as well as
+arguments and variables that said function can take \citep{karimzadeh2018top}.
+Individual documentation for each function is usually presented in alphabetic order,
+and does not have a specific analysis goal.
+Moreover, most software has numerous functions, so documentation is usually very
+lengthy and it is hard to navigate.
+This can have the effect of increasing the amount of information that needs to be
+simultaneously processed by the users, which can lead to overload of the finite amount
+of working memory any one possesses, known as cognitive load \citep{sweller1988cognitive}.
+In this context, identifying connections across functions that are meant to work
+on the same analysis workflow can become a very difficult task.
+We recognize this as \textit{barrier 2.-- Lack of specific goals leading to high cognitive load}.
+High cognitive load is know to have a negative effect in learning software
+\citep{chandler1996cognitive, van2005research, lambert2009student}.
 
-Another important aspect of software documentation is that examples are usually
-worked and showcase the ideal or minimal case in which a function should work well.
-\textit{Barrier 3.--} Focusing importance on the content of the software instead
-of considering the user experience, ignoring what can go wrong and avoiding clear
-advice on how to troubleshoot.
+A third important aspect of software documentation are examples that demonstrate
+usage of individual functions \citep{karimzadeh2018top}.
+Examples presented in software documentation are usually
+worked to perfection, as they are intended to showcase the ideal or minimal case
+in which a function works well.
+Perfectly worked examples ignore the user experience by
+maintaining focus on the software content, and fail to provide users with expert
+and clear advice on how to troubleshoot if needed.
+We identify this as \textit{barrier 3.-- Content-focused examples}.
+Error management training is an approach that focuses on framing mistakes as beneficial
+to learning complex tasks, to give learners the opportunity to
+actively explore a task \citep{frese1995error}.
+Providing examples that showcase potential errors supports users performance \citep{steele2014error},
+and can greatly improve learners ability to troubleshoot outside the classroom
+\citep{shannon2015live, nederbragt2020ten}.
 
-Software documentation is usually explained with words, very rarely using diagrams or
-other allegories.
-\textit{Barrier 4.--} Little to no diversity of representation of information
+Last but not least, software documentation is usually explained verbally in a document, very
+rarely is it approached using diagrams or other allegories.
+We recognize this as \textit{barrier 4.-- Little to no diversity of representation of information}.
 
-Documentation should always be available for the users and learners
-\textit{Barrier 5.--} Inflexible learning environment, is this one really an issue???
+In sum, best practices for good software documentation are not enough to promote
+reproducibility of research workflows that rely heavily on code. In the following
+section we describe some principles that can help to reduce or remove the
+identified barriers, to create
+research workflows that are more approachable and hence more reproducible by a
+larger audience.
 
-% Some principles of an inclusive syllabus:
-
-% - unspecified, unclear or lengthy goals; high cognitive load; (literate programming)
-
-% - content-focused descriptions/examples instead of user-focused; (demonstrate errors)
-
-% - inflexible learning environment; (make it stable so users can learn whenever is best for them)
-
-% - unapproachable (cold or intimidating) language, use of highly specialized computing jargon, skill level; (use friendly and informal but respectful language)
-
-% - and little to no diversity of representation of information (say the same thing with different images, metaphors, stories)
-
-In sum, best practices for good primary documentation are not enough to ensure
-reproducibility of research workflows that rely heavily on code.
-
-\section*{Best practices for approachable research workflows}
+\section*{Principles for approachable research workflows}
 \label{sec:addressing}
 
-\subsubsection*{a. Reduce cognitive load and provide specific and clear goals
+\subsubsection*{Principle 1. Use friendly, relatable and respectful language}
+
+% avoid specialized, intimidating and cold language
+
+Besides avoiding formal language, and incorporating elements of pop culture, such as picture
+character icons known as ``emojis``, to make the language more familiar to a
+broader target audience (see Figure \ref{fig:error}), we made an effort to specifically
+complement the primary documentation by identifying
+computational concepts that were assumed or were not explained in depth.
+We vetted the tutorials through feedback from workshop participants as well as
+individual users to identify such specialized concepts.
+
+\subsubsection*{Principle 2. Reduce cognitive load by providing specific and clear goals
 with literate programming}
 
-Pedagogical research shows that active learning practices are one the most effective
+Cognitive load can be greatly reduced for learners by applying an active learning
+strategy such as linking usage to a ``real world'' or ``human'' application
+\citep{felder2009active}.
+Programming computer languages are by themselves quite abstract and represent a
+learning subject with a potentially high cognitive load for most learners.
+Pedagogical research shows that active learning practices are one of the most effective
 ways to take on abstract subjects \citep{freeman2014active}.
-Programming computer languages are quite abstract and cognitive load can be greatly
-reduced for learners by applying an active learning strategy such as linking its usage to
-a ``real world'' or ``human'' application \citep{felder2009active}.
 
@@
 
-\subsubsection*{b. Provide examples that are user-focused by demonstrating errors and warnings}
+\subsubsection*{Principle 3. Provide examples that are user-focused by demonstrating errors and warnings}
 
@@
 A practice that has become more and more widespread in programming-language pedagogical practices
-is the use of typos and mistakes to normalize them for learners, and show them how
-to solve them when they are outside the classroom \citep{shannon2015live}.
-Yet, this is rarely done on written pedagogical materials.
-Primary documentation focuses on demonstrating usage function with examples that
+is live programming.
+One benefit of live programming is that typos and mistakes occur -- which normalizes
+them for learners, and shows them how to solve them when they are outside the
+classroom \citep{shannon2015live, nederbragt2020ten}.
+When co-author McTavish was a postdoc, teaching an introductory programming workshop as a volunteer
+with the Carpentries (a non-profit group that teaches foundational coding and
+ data science skills to researchers worldwide) \citep{wilson2006swc, SWCwebsite},
+a senior faculty member taking the workshop
+complained that the typos were slowing things down and interfering with the pedagogy.
+McTavish replied "the typos ARE the pedagogy".
+This has become a slogan of sorts at the Carpentries, capturing the idea that embracing
+and discussing mistakes is essential to teaching programming \citep{wilson2019teaching}.
+Yet, working through mistakes is rarely done on written pedagogical materials.
+Software documentation focuses on demonstrating usage function with examples that
 work seamlessly, without errors. We argue that the opposite is needed to support
 adoption of reproducible workflows and support long term independence in learners
-\citep{gaspar2007restoring}.
-We demonstrate examples that do not work
-as expected and exemplify ways to address them (Figure \ref{fig:error}).
+and users performance
+\citep{gaspar2007restoring, steele2014error}.
+In our tutorials, we apply this principle by demonstrating examples that do not work
+as expected, and exemplifying ways to address them (Figure \ref{fig:error}).
 
-We identify inputs that would give
-a wide range of warnings and errors, focusing on demonstrating these cases. This
-helps users to not be afraid of errors and warnings, but instead to use them to
-their advantage.
+We identified inputs that would give
+a wide range of warnings and errors, and support learners on reading the information
+provided by the message. This
+helps users to not be afraid of errors and warnings, but instead use them as
+an informative measure.
+% This is probably part of the computer science curriculum, but it is not at all part of the natural sciences curriculum.
 We also identify effects of warnings and errors downstream of the workflow.
@@
 from errors.
-Many finely honed tutorials do not trigger errors, which precludes helping students
-to develop the tools to understand and address errors when they do encounter them,
-as they inevitably will.
-In our tutorial, we focus on explaining the meaning and downstream of warnings and errors, and
+
+In our tutorial, we focus on explaining the meaning of warnings and errors, and
  showcase ways to detect them before they are triggered (i.e., before using an input
@@
 
-% We designed ways to access the different elements of the outputs.
+\subsubsection*{Principle 4. Demonstrate one thing in different ways}
 
-\begin{figure}
-\begin{center}
-\includegraphics[width=3in]{fig-error.png}
-\end{center}
-\caption{Snapshot of a section of the tutorial website, where we demonstrate a common error. \label{fig:error}}
-\end{figure}
+% - and little to no diversity of representation of information (say the same thing with different images, metaphors, stories)
 
-\subsubsection*{c. Use friendly, relatable and respectful language}
-
-% avoid specialized, intimidating and cold language
-
-Besides avoiding formal language, and incorporating elements of pop culture, such as picture
-character icons known as ``emojis``, to make the language more familiar to a
-broader target audience (see Figure \ref{fig:error}), we made an effort to specifically
-complement the primary documentation by identifying
-computational concepts that were assumed or were not explained in depth.
-We vetted the tutorials through feedback from workshop participants as well as
-individual users to identify such specialized concepts.
-
-\subsubsection*{d. Make it accessible geographically and through time}
-
-% - inflexible learning environment; (make it stable so users can learn whenever is best for them)
-
-We published the tutorials on a public, free license, free of cost, and free for
-use and reuse repository and persistent website \citep{RopentreeTutorials}.
-The tutorial is available for the users to go back to any time they need it,
-and to be passed on to other users (Figure \ref{fig:schedule}).
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=3in]{fig-schedule.png}
-\end{center}
-\caption{Snapshot of the home to our tutorial website, showing part of the schedule. \label{fig:schedule}}
-\end{figure}
-
-We created a main version of the tutorial that is stable. Any updates to the tutorial are
-published as new versions, or tutorials for new workshops \citep{wilson2006swc, SWCwebsite}
- Versions presented at workshops are a copy from the original repository.
-They represent a temporally stable snapshot of functions and workflows presented
-during a workshop.
-
-%%% \subsubsection*{5. The (now) classics of computational reproducibility}
-%%%
-%%% Provide all information on package version and system capabilities.
-%%%
-%%% > [name=Emily Jane McTavish]link to where/and how you did that
+%We designed ways to access the different elements of the outputs.
 
@@
 
-Response form the community has been invaluable in gauging success of our teaching
+Response from the community has been invaluable in gauging success of our teaching
 materials.
@@
 
-Making accessible reproducible workflows has several advantages:
+One key aspect for success of our tutorials is that we applied all rules for
+successful reproducibility, in particular making our tutorials available
+through time in a public website.
+
+Making approachable research workflows has several advantages:
 
@@
 
+% \citep{roland2002think}
 Ultimately, the long term improvement of reproducibility rates in science will depend
-on our ability to intentionally integrate the subject of reproducibility into the
+on our ability to intentionally integrate reproducibility into the
 undergraduate curriculum, so college learners and future researchers have the
@@
 
+\bigskip
+\bigskip
+\bigskip
+
+\begin{center}
+{\large\bf FIGURES}
+\end{center}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=3in]{fig-tree.png}
+\end{center}
+\caption{A phylogenetic tree from our tutorial.
+It was extracted using OpenTree of Life resources \citep{opentreeoflife2019synth} wrapped in the \texttt{rotl} R package \citep{michonneau2016rotl}.
+\label{fig:tree}}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=3in]{fig-error.png}
+\end{center}
+\caption{Snapshot of a section of the tutorial website, where we demonstrate a common error. \label{fig:error}}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=3in]{fig-schedule.png}
+\end{center}
+\caption{Snapshot of the home to our tutorial website, showing part of the schedule.
+ Our tutorial website was constructed using The Software Carpentries workshop template \citep{swc-workshop-template}. \label{fig:schedule}}
+\end{figure}
+
 \bibliography{Manuscript-bibliography}

### 2022-02-16 jsdse/Manuscript.tex
almost there

@@
     The authors gratefully acknowledge ``Sustaining the Open Tree of Life'',
-    NSF ABI No. 1759838, and ABI No. 1759846. They also acknowledge ``The Carpentries''
-    project, without it this project could not have been possible}\hspace{.2cm}\\
+    NSF ABI No. 1759838, and ABI No. 1759846. They are also also deeply grateful towards
+    ``The Carpentries'' organization; without its invaluable volunteers and resources
+    this project could not have been possible}\hspace{.2cm}\\
     School of Natural Sciences, University of California, Merced\\
@@
 are available for download and use from OpenTree free of cost to any user.
-One way to access OpenTree resources is through it Graphical User
+One way to access OpenTree resources is through its Graphical User
 Interface (GUI; aka, a website or application that allows users to access computer
@@
 to learning complex tasks, to give learners the opportunity to
-actively explore a task \citep{frese1995error}.
+actively explore a task with a positive mindset \citep{frese1995error}.
 Providing examples that showcase potential errors supports users performance \citep{steele2014error},
@@
 character icons known as ``emojis``, to make the language more familiar to a
-broader target audience (see Figure \ref{fig:error}), we made an effort to specifically
+broader target audience (Figure \ref{fig:error}), we made an effort to specifically
 complement the primary documentation by identifying
@@
 ways to take on abstract subjects \citep{freeman2014active}.
-
 A story-like narrative that links code usage in an integrative example, invites learners
@@
 a highly charismatic group such as dinosaurs will do the trick.
+For example, when we presented our tutorial to the Amphibia Web Organization \citep{van2002amphibiaweb}
+in January 2020, we tailored all examples to frogs and their allies.
+% \citep{cite-aw-workshop}
 
-We examined available primary documentation for the package \texttt{rotl},
-and designed a narrative that required the usage of as many functions as possible.
+We examined available software documentation for the package \texttt{rotl},
+and designed a narrative that requires the usage of as many functions as possible.
 We demonstrate code applications that are commonly requested by OpenTree users,
-but that are not demonstrated in the primary documentation of the R package.
+but that are not demonstrated in the documentation of the package.
 By framing the function workflow using highly requested uses, the documentation acquires a
 narrative arc that is easier to follow and remember by users. This can also facilitate
-the application of code to other use cases in biology of interest for the users.
+translating the code application to other use cases of interest for learners in biology.
 
-%% Maybe Figure!! For the tutorial demonstrated here, we used the commonly requested
+%% Maybe Figure!! In the tutorial showcased here, we used the commonly requested
 %% use case of obtaining
@@
 adoption of reproducible workflows and support long term independence in learners
-and users performance
-\citep{gaspar2007restoring, steele2014error}.
+and users performance \citep{gaspar2007restoring, steele2014error}.
+
 In our tutorials, we apply this principle by demonstrating examples that do not work
 as expected, and exemplifying ways to address them (Figure \ref{fig:error}).
+For example, we identified inputs that would give
+a wide range of warnings and errors. We then focus on providing explanations for
+error and warning messages. We believe this supports users and learners to be less
+intimidated by the messages and practice taking useful information out of it.
+% This is probably part of the computer science curriculum, but it is not at all
+% part of the natural sciences curriculum.
 
-We identified inputs that would give
-a wide range of warnings and errors, and support learners on reading the information
-provided by the message. This
-helps users to not be afraid of errors and warnings, but instead use them as
-an informative measure.
-% This is probably part of the computer science curriculum, but it is not at all part of the natural sciences curriculum.
-We also identify effects of warnings and errors downstream of the workflow.
-
-We identify ways to evaluate inputs to know if they will produce an error, and design
-alternatives on what to do when faced with an error or warning, and demonstrate
-these alternatives.
+We also demonstrate ways to evaluate inputs to determine if they will trigger an
+error or warning, and design and demonstrate alternative analysis routes on what
+to do when faced with an error or warning.
 One of the most essential skills in programming is interpreting and moving forward
 from errors.
-
-In our tutorial, we focus on explaining the meaning of warnings and errors, and
- showcase ways to detect them before they are triggered (i.e., before using an input
-  that would elicit a warning or error). This has two pedagogical benefits:
-1) it provides users/students with the means to troubleshoot their own warnings and errors, and
-2) it allows users/students to understand with more depth what the function is doing.
+This has two pedagogical benefits. First, it provides users and learners with the
+means to troubleshoot their own warnings and errors. Second, it allows them to
+understand with more depth what the function is doing.
 
@@
 
-Making approachable research workflows has several advantages:
-
-* save explanation/training time when analyses are run again by students and collaborators;
-* save research time for yourself when analyses are run again with more data, a* different dataset, a different organism or biological model;
-* scientific efforts can build off of each other.
+Making approachable research workflows has all the advantages of reproducible research
+workflows.
+It saves explanation/training time when analyses are run again by students and
+collaborators.
+It can also save research time for yourself when analyses are run again with more
+data, a different dataset, a different organism or biological model;
+It contributes to scientific efforts that can build off of each other.
 
@@
 while female individuals are discouraged \citep{warner2021quantifying}.
-% How to balance software acceptance VS. adoption?
 These principles can be used to improve not only reproducibility practices,
 but also software adoption in the natural sciences.
-%% Discuss: why address accessibility and not other aspects of reproducibility?
 
@@
 
-\bigskip
-\bigskip
-\bigskip
+\pagebreak
 
@@
 \end{center}
+
+\newline
+\newline
+\newline
 
@@
 
+\pagebreak
+
 \bibliography{Manuscript-bibliography}

### 2022-02-16 jsdse/reviews/response_2022_02_15.Rmd
done!

---
output: rmarkdown::pdf_document
header-includes:
    - \usepackage{setspace}\doublespacing
    - \usepackage{float}
geometry: margin=1.2in
fontsize: 12pt
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

February 15, 2022

University of California, Merced

&nbsp;  &nbsp;

Dear Dr. Horton, associate editor, and reviewers,

&nbsp;

We appreciate the helpful and thoughtful comments, and thank you for the opportunity to revise and re-submit our manuscript, formerly titled *“Promoting analysis reproducibility with accessibility: An example in evolutionary biology”*, and now titled **“Approachable case studies support learning and reproducibility in data science: An example from evolutionary biology”** (please see response to reviewer 1).

We have thoroughly revised and restructured the manuscript. The comments of the editors and reviewers have greatly improved the manuscript. We focused on the contributions of our work related to teaching reproducibility, and connected our strategies to the pedagogical literature. We hope that you find it suitable for publication in the special issue about teaching reproducibility and responsible workflow.

We address individual comments line by line below.

&nbsp;  &nbsp;

Please receive our respectful regards,

&nbsp;  &nbsp;

The authors

\pagebreak

\singlespacing

*22-Dec-2021*

*Dear Dr Luna Luisa Sanchez Reyes:*

*Your manuscript entitled "Promoting analysis reproducibility with accessibility: An example in evolutionary biology", which you submitted to Journal of Statistics and Data Science Education, has been reviewed by myself, an associate editor, and two reviewers. Overall the reviews were quite positive about the submission, though some issues were noted that dampen enthusiasm. The reviewer comments are included at the bottom of this letter or as attachments.*

*I regret to inform you that the reviewers have raised serious concerns, and therefore your paper cannot be accepted for publication in Journal of Statistics and Data Science Education. However since the reviewers do find considerable merit in the paper, I would be willing to reconsider if you wish to undertake major revisions and re-submit, addressing the reviewers' concerns.*
*Please note that resubmitting your manuscript does not guarantee eventual acceptance, and that your resubmission will be subject to re-review before a decision is rendered.*

*Sincerely,*

*Dr Nicholas Horton*

*Editor, Journal of Statistics and Data Science Education*

*nhorton@amherst.edu, nicholasjhorton@gmail.com*

&nbsp;  &nbsp;

*Comments from the Editors and Reviewers:*

*Editor*

*Comments to the Author:*

*It would be helpful to ground this research in broader initiatives (e.g., NASEM 2019 reproducibility report, NASEM 2018 data science for undergraduates definition of "data acumen").*

These references were really helpful, thank you. We address the NASEM initiatives and its relation to the principles we propose in the third paragraph of the **Conclusions** section.

*In addition, a close edit would be indicated as there were a number of typographical errors throughout the initial submission.*

Thanks. We performed a close edit to eliminate as many typographical errors as possible. We apologize in advance for any errors that might have passed unnoticed and hence uncorrected.

*********************
\pagebreak

*Associate Editor*

*Comments to the Author:*

*While the teaching materials developed by you have a potential to be useful for other lecturers and students of statistics and evolutionary biology, there are some deficiencies in the paper related to how you present your work. Especially sections 3 and 4 need some extensive revisions and improvement (see below). Your writing style could be improved in places too, as many sentences are short enumerations of steps you have undertaken (i.e., ‘We identified…’, We did this and We did that), without unpacking your rationale or any reflections on the process. So some sentences read like notes, not arguments in a scientific paper.*

Thank you for your comment. You accurately point out that the last sections of the ms were quite raw. We revised our ms extensively and worked especially on the sections that appeared as enumerations. 
**Section 3 - Results and Discussion** and **Section 4 - Conclusions** are now one single section **Conclusions**. We decided to eliminate the discussion section based on a suggestion of Reviewer 2: "This isn’t a study that requires the traditional Intro/Methods/Results/Discussion format. This is more of a description or commentary. I believe that this inevitably leads to the Results/Discussion section not making much sense, at least to me. I suggest replacing the section headers with something more appropriate and then after explaining your main thesis/criticism you then wrap up the article with a concluding paragraph [...]"
We elaborate more on this below.

&nbsp;  &nbsp;

*Section 3 - Results and Discussion - please revise and expand the section. It is not clear how it discusses your results and results of what activity or a test are presented here. The section reads in many places like notes which need further explanation and evidence.*

We expanded and elaborated the ideas that were originally stated in the Section 3 of the original ms, and connected them to the barriers and principles identified in previous sections.

&nbsp;  &nbsp;

*Section 4 - Conclusions - this section reads as an unfinished draft. Please polish it in the light of the original contribution of your work related to teaching reproducibility. You should also link your contribution to the current debates on reproducibility, open scholarship and pedagogical approaches.*

We added a final paragraph addressing what we believe to be the main contribution of our work to teaching reproducibility.

&nbsp;  &nbsp;

*When revising the paper please remember that your manuscript was submitted to a special issue on 'Teaching Reproducibility and Responsible Workflow' (more details can be found here: https://magazine.amstat.org/blog/2021/07/01/jsdse-call-for-papers/).*

Thanks, we framed our analysis around the particular challenges related to teaching reproducible workflows. As suggested by the editor, we ground our principles around the "data acumen" concept defined in the "Data Science for Undergraduates" report (NASEM, 2018 doi:10.17226/25104.)

********************

&nbsp;  &nbsp;  &nbsp;

***Reviewer: 1***

*Comments to the Author*

*This article emphasizes the importance of accessibility of computational tools and datasets, as opposed to their mere availability. Accessibility is discussed using the example of a tutorial given as part of a workshop on using the "Open Tree of Life", an online database front-end meant to facilitate access to public phylogenetic data. The tutorial focuses on the use of the R language for accessing the Open Tree of Life, assuming some prior familiarity with this language. The support material for the tutorial is publicly available, both for students who wish to re-visit it, and for educators as a source of inspiration or reuse.*

*Accessibility is an important and often overlooked aspect in the ongoing transition to more open, transparent, and reproducible research practices. Following the lead of the Open Source movement for computer software, Open Science activists tend to focus on publication and citation of digital assets, claiming that "if it's available, people can inspect and reuse it". The emphasis on accessibility is an important complement contributed by this work.  Accessibility can be improved in essentially two ways: writing code and documentation in such a way to be usable with less technical knowledge, and educating scientists in the techniques required to work with published digital assets. The article discusses both aspects, focusing on the role of documentation. Another aspect that the article emphasizes is the need to teach not only the correct use of software, but also how to deal with the inevitable errors.*

Thank you for your comments!

That is exactly what we were trying to convey: the fact that something is available does not ensure that people can actually reach that something and interact with it in the way it is intended.

While our main thesis remains unchanged, we decided to replace the term **accesibility** for **approachability**. After sharing a reworked version of the ms with colleagues for feedback, several of them pointed out that the term accessibility is historically and practically tied to addressing barriers faced by people with different physical abilities (visual, cognitive, motor, etc.). As we were not referring to that type of  barriers in our ms, they suggested that we should elaborate more on the way we were applying the term **accessibility** for our thesis. This inevitably lead to somewhat of a dark rabbit-hole of jargon. Trying to be consistent with our own principles (avoid use of specialized language), we decided to use a word that conveyed the idea without becoming a coined concept of its own. Approachable appeared as the most friendly candidate, so we decided to reframe the whole ms in terms of **approachability** and move away from accessibility.
Nevertheless, the problem that we identify and try to tackle with this work remains the same.

&nbsp;  &nbsp;

*There are two minor points that need to be addressed in a revision:*

*Page 6: "following the carpentries" This comment deserves a reference, and maybe a few words about the carpentries' approach, unless this can be considered generally known in the intended audience.*

We introduced "the Carpentries" and added references to the materials and research papers that support the theoretical and practical backbone of our paper. We also added "the Carpentries" to acknowledgements, as the resources released by the organization were truly essential for the existence of this work.

&nbsp;  &nbsp;

*Page 7: "Discuss: why address accessibility and not other aspects of reproducibility?" This looks like an unhandled reminder to the authors themselves.*

Indeed, thanks for pointing this out. We removed this sentence. 
As we no longer address the issue of reproducibility from the viewpoint of accessibility
in particular, but from a broader concept of approachability, it is no longer needed to elaborate on this reminder.

********************

&nbsp;  &nbsp;  &nbsp;

***Reviewer: 2***

*Comments to the Author*

*General comment*

*This manuscript touches on some really important facets of the replication/reproducibility problem, namely the link between producing and publishing code and the end user’s ability to use the code. The problem is often one of communication or what Steven Pinker calls “the curse of knowledge”, the speaker knows what they are talking about so well that they leave out important details that the naive listener needs to make sense of the statement.*

Thanks!

&nbsp;  &nbsp;

*My main comment here is that I do not think the format of the manuscript is appropriate. This isn’t a study that requires the traditional Intro/Methods/Results/Discussion format. This is more of a description or commentary. I believe that this inevitably leads to the Results/Discussion section not making much sense, at least to me. I would suggest replacing the section headers with something more appropriate and then after explaining your main thesis/criticism you then wrap up the article with a concluding paragraph highlighting what authors of future vignettes and tutorials need to provide for everyone to be able to use the code.*

We agree. We reorganized the ms and removed the results/discussion section. We elaborated on the sections where we identify the barriers and describe the principles and its application. 
We wrap up the ms with a **conclusion** section.

&nbsp;  &nbsp;

*Specific comments*

*Page 3, Line 31: Replace “in” with “on”* - Done

*Page 3, Line 45: Accessibility of what specifically?* 
- Accessibility turned out to be a somewhat confusing term (see response to reviewer 1). What we were trying to convey is that most software documentation is difficult to be approached by a wider audience, which renders analysis workflows that rely on that software harder to reproduce by the majority of the potential user base. We replaced the term accessibility for approachability across the ms.

*Page 3, Line 52: Remove a closing bracket.* - Done

*Page 4, Line 15: Replace “utilized” with “used”* - Done

*Page 4, Line 24: required baseline, not baseline required* - Done

*Page 4, Line 26: The required computational skills are likely…* - Done

*Page 4, Line 35: Replace “for reproducibility research…” with “…of reproducing published research”* - Done

*Page 4, Line 36: What does “the utility of data resources…” mean? Please clarify.* We removed this sentence, it was no longer necessary. But it referred to the usability or usefulness of data resources.

*Page 4, Line 38: What does “that gap” refer to? Please clarify.* - We removed that statement, it was no longer necessary.

*Page 4, Line 42-43: Replace “is written using” with “communicates in a”.* - Done

*Page 5, Line 19: By “external” do you mean “naive”?* - We mean "the non-specialized users" that we now refer to as the "wider audience".

*Page 7, Line 35-36: Should be: “the individual user” or “individual users”.* - Done

*Page 7, Line 49-50: Should be: “…users to return to to at any time, and to be…”* - Done

*Page 7, Line 52: What are carpentries? This article aims to make data and code clearer for users yet you drop lingo like this into it; make sure you follow your own advice.* - You are right. We define the carpentries and reference it when relevant.

*Page 7, Line 54: Should be: “…presented in workshops…”*  - Done

*Page 8: Results and Discussion, and Conclusions. Perhaps transform these into single summary paragraph*  - Agreed. All those are now a single section named **"Conclusions"**.

### 2022-02-16 jsdse/Manuscript.tex
done!

@@
 Research reproducibility is essential for scientific development. Yet, rates of reproducibility
-are low, especially in the natural sciences. As increasingly more research is relying on
-computing tools and software, efforts for improving
+are low, especially in the natural sciences. As increasingly more research relies on
+computers and software, efforts for improving
 reproducibility rates have focused on making available research workflows
@@
 However, research products that are digitally available are not necessarily
-friendly for learners and interested parties with little to no %%% different levels of
-experience in the field. This renders research products unapproachable, which
+friendly for learners and interested parties with little to no
+experience in the field. This renders research products unapproachable,
 counteracts availability, and hinders reproducibility short and long term.
@@
 that make them unapproachable to the general audience:
-use of highly specialized language is intimidating;
-unspecified, unclear or lengthy goals lead to high cognitive load;
-content-focused descriptions instead of user-focused; and
-little to no diversity of representation of information.
-Then, we propose a set of principles to improve the unapproachable aspects of
-research workflows, and illustrate their application in a case study from
-evolutionary biology that we used as teaching material.
-Finally, we elaborate on the general application of these principles for
+use of highly specialized language;
+unclear goals and high cognitive load; and
+content-focused descriptions.
+We propose a set of principles to improve the unapproachable aspects of
+research workflows, and illustrate their application in a tutorial.
+We elaborate on the general application of these principles for
 documenting research workflows and products, to provide present learners
@@
 %% resources.
-%% However, it is often written using highly specialized language that is largely
+%% However, it often communicates in a highly specialized language that is largely
 %% inaccessible for the average target user. Additionally, baseline
-%% computational knowledge and skills required to access scientific results are often
+%% computational knowledge and required skills to access scientific results are often
 %% not found among potential users, and keep on increasing.
@@
 \noindent%
-{\it Keywords:}  open science, R, phylogenetics, Open Tree of Life, pedagogy
+{\it Keywords:}  open science, R programming language, phylogenetics, Open Tree of Life, pedagogy
 \vfill
@@
 potential to positively impact research across the natural sciences.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=3in]{fig-tree.png}
+\end{center}
+\caption{A phylogenetic tree from our tutorial.
+It was extracted using OpenTree of Life resources \citep{opentreeoflife2019synth} wrapped in the \texttt{rotl} R package \citep{michonneau2016rotl}.
+\label{fig:tree}}
+\end{figure}
 
@@
 
-Last but not least, software documentation is usually explained verbally in a document, very
-rarely is it approached using diagrams or other allegories.
-We recognize this as \textit{barrier 4.-- Little to no diversity of representation of information}.
+%% Last but not least, software documentation is usually explained verbally in a document, very
+%% rarely is it approached using diagrams or other allegories.
+%% We recognize this as \textit{barrier 4.-- Little to no diversity of representation of information}.
 
 In sum, best practices for good software documentation are not enough to promote
-reproducibility of research workflows that rely heavily on code. In the following
+reproducibility of published research workflows that rely heavily on code. In the following
 section we describe some principles that can help to reduce or remove the
@@
 individual users to identify such specialized concepts.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=3in]{fig-error.png}
+\end{center}
+\caption{Snapshot of a section of the tutorial website, where we demonstrate a common error. \label{fig:error}}
+\end{figure}
 
@@
 
-\subsubsection*{Principle 4. Demonstrate one thing in different ways}
+% \subsubsection*{Principle 4. Demonstrate one thing in different ways}
 
@@
 
-%We designed ways to access the different elements of the outputs.
+% We designed ways to access the different elements of the outputs.
 
@@
 of the tutorials for their research, as well as how they have supported students in using the
-R packages with less help from them as PIs.
+R packages more independently.
 
-One key aspect for success of our tutorials is that we applied all rules for
-successful reproducibility, in particular making our tutorials available
-through time in a public website.
+When developing our tutorials, we not only applied the principles elaborated here,
+but we followed basic recommendations for successful reproducibility \citep{sandve2013ten}.
+For example, we published the tutorials on a public, free license, free of cost, and free for
+use and reuse repository and persistent website \citep{RopentreeTutorials}.
+We also created a main version of the tutorial that is stable. Any updates to the tutorial are
+then published as new versions, or tutorials for new workshops \citep{wilson2006swc, SWCwebsite}
+Versions presented at workshops are a copy from the original repository, and
+constitute a temporally stable snapshot of functions and workflows presented
+during a workshop.
+This ensures that the tutorials are available for the users to return to any time
+they need it, and to be shared with other users and learners (Figure \ref{fig:schedule}).
 
-Making approachable research workflows has all the advantages of reproducible research
-workflows.
+\begin{figure}
+\begin{center}
+\includegraphics[width=3in]{fig-schedule.png}
+\end{center}
+\caption{Snapshot of the home to our tutorial website, showing part of the schedule.
+ Our tutorial website was constructed using The Software Carpentries workshop template \citep{swc-workshop-template}. \label{fig:schedule}}
+\end{figure}
+
+We note that making approachable research workflows has all the advantages of
+reproducible research workflows.
 It saves explanation/training time when analyses are run again by students and
@@
 % \citep{roland2002think}
+Some universities have been incorporating reproducibility in their curriculum
+\citep{uwlibraries2022, nigms2022}.
+The focus of these resources has been for students to acquire and practice skills
+to document their work.
+The principles identified and outlined here can be used to set learning goals and
+outcomes for new reproducibility syllabi.
 Ultimately, the long term improvement of reproducibility rates in science will depend
-on our ability to intentionally integrate reproducibility into the
-undergraduate curriculum, so college learners and future researchers have the
-basis to develop the fundamental skills needed to successfully create reproducible
-scientific workflows and materials.
-
-Some universities have been incorporating the subject in their classes (see
-\cite{uwlibraries2022, nigms2022}).
-The focus of these resources has been for students to develop skills to document their work.
-The principles identified and outlined here can be used to set learning goals and
-outcomes on new reproducibility syllabi.
+on our ability to intentionally integrate best practices for achieving reproducibility
+into the educational framework of future data scientists \citep{nasem2018data}.
+Inclusion of reproducibility into the data acumen of undergraduate curriculum will
+provide college learners and future researchers with the tools to develop
+the fundamental skills needed to successfully create reproducible scientific workflows
+and research products.
 
@@
 
-\pagebreak
-
-\begin{center}
-{\large\bf FIGURES}
-\end{center}
-
-\newline
-\newline
-\newline
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=3in]{fig-tree.png}
-\end{center}
-\caption{A phylogenetic tree from our tutorial.
-It was extracted using OpenTree of Life resources \citep{opentreeoflife2019synth} wrapped in the \texttt{rotl} R package \citep{michonneau2016rotl}.
-\label{fig:tree}}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=3in]{fig-error.png}
-\end{center}
-\caption{Snapshot of a section of the tutorial website, where we demonstrate a common error. \label{fig:error}}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=3in]{fig-schedule.png}
-\end{center}
-\caption{Snapshot of the home to our tutorial website, showing part of the schedule.
- Our tutorial website was constructed using The Software Carpentries workshop template \citep{swc-workshop-template}. \label{fig:schedule}}
-\end{figure}
-
-\pagebreak
-
 \bibliography{Manuscript-bibliography}

### 2022-06-08 jsdse/Manuscript.tex

@@
 \begin{abstract}
-Research reproducibility is essential for scientific development. Yet, rates of reproducibility
-are low, especially in the natural sciences. As increasingly more research relies on
-computers and software, efforts for improving
-reproducibility rates have focused on making available research workflows
-as computer code, as well as raw and processed data in computer readable form.
-However, research products that are digitally available are not necessarily
-friendly for learners and interested parties with little to no
-experience in the field. This renders research products unapproachable,
-counteracts availability, and hinders reproducibility short and long term.
-To improve long term adoption of reproducible workflows in research, they
-need to be made approachable for learners, the researchers of the future.
+Research reproducibility is essential for scientific development. Yet, rates of reproducibility are low, especially in the natural sciences. 
+As increasingly more research relies on computers and software, efforts for improving reproducibility rates have focused on making available research workflows as computer code, as well as raw and processed data in computer readable form.
+However, research products that are digitally available are not necessarily friendly for learners and interested parties with little to no experience in the field. 
+This renders research products unapproachable, counteracts availability, and hinders reproducibility short and long term. 
+To improve long term adoption of reproducible workflows in research, they need to be made approachable for learners, the researchers of the future.
 
-Using a case study within evolutionary biology, we identify aspects of research workflows
-that make them unapproachable to the general audience:
+Using a case study within evolutionary biology, we identify aspects of research workflows that make them unapproachable to the general audience: 
 use of highly specialized language; 
@@
 content-focused descriptions. 
-We propose a set of principles to improve the unapproachable aspects of
-research workflows, and illustrate their application in a tutorial.
-We elaborate on the general application of these principles for
-documenting research workflows and products, to provide present learners
-and future researchers with tools for successful scientific reproducibility.
+We propose a set of principles to improve the unapproachable aspects of research workflows, and illustrate their application in a tutorial. 
+We elaborate on the general application of these principles for documenting research workflows and products, to provide present learners and future researchers with tools for successful scientific reproducibility.
 %> ** add a sentence on implications for teaching reproducibility
@@
 \noindent%
-{\it Keywords:}  open science, R programming language, phylogenetics, Open Tree of Life, pedagogy
+{\it Keywords:}  open science, R programming language, phylogenetics, Open Tree of Life, pedagogy, phylogenetics, datelife
 \vfill
@@
 Research reproducibility --the extent to which consistent results are obtained when
-a scientific experiment or research workflow is repeated \citep{repdef2021}--
-is a key aspect of the advancement
-of science, as it constitutes a minimum standard that allows understanding research products,
-i.e., methods, data, analysis, results, etc. \citep{piwowar2013value},
-to determine their reliability and generality, and eventually build up scientific
-knowledge and applications based on those products
-\citep{king1995replication, peng2011reproducible, powers2019open}.
-In the natural sciences, rates of reproducibility are low \citep{ioannidis2005most, prinz2011believe},
-which has elicited concerns about a crisis in the field \citep{baker2016reproducibility}.
+a scientific experiment or research workflow is repeated \citep{repdef2021}-- is a key aspect of the advancement of science, as it constitutes a minimum standard that allows understanding research products, i.e., methods, data, analysis, results, etc.
+\citep{piwowar2013value}, to determine their reliability and generality, and eventually build up scientific knowledge and applications based on those products \citep{king1995replication, peng2011reproducible, powers2019open}.
+In the natural sciences, rates of reproducibility are low \citep{ioannidis2005most, prinz2011believe}, which has elicited concerns about a crisis in the field \citep{baker2016reproducibility}.
 
-In response, the scientific community has been developing new principles and standards to incentivize
-cultural changes that support a long term improvement of reproducibility rates in the natural sciences
-\citep{peng2015reproducibility, wilkinson2016fair, miyakawa2020no}.
-A standard for reproducibility that has received much attention is availability, which
-we define as a property denoting that a research product can be reached (acquired, copied, analyzed,
-processed and/or reused) at no financial, legal or technical cost \citep{arnold2019turing},
-and without geographic, demographic, social or temporal barriers for the population \citep{fecher2014open}.
+In response, the scientific community has been developing new principles and standards to incentivize cultural changes that support a long term improvement of reproducibility rates in the natural sciences \citep{peng2015reproducibility, wilkinson2016fair, miyakawa2020no}.
+A standard for reproducibility that has received much attention is availability, which we define as a property denoting that a research product can be reached (acquired, copied, analyzed, processed and/or reused) at no financial, legal or technical cost \citep{arnold2019turing}, and without geographic, demographic, social or temporal barriers for the population \citep{fecher2014open}.
 
-In this paper, we argue that research products that are digitally available are
-often unapproachable in practice, because they are not
-friendly for learners and interested parties with different levels of
-experience in the field. Research products that are unapproachable
-counteract availability, and hinder reproducibility short and long term.
-To support long term adoption of reproducible practices in the natural sciences, research
-workflows need to be made approachable for learners, the researchers of the future \citep{roland2002think}.
+In this paper, we argue that research products that are digitally available are often unapproachable in practice, because they are not friendly for learners and interested parties with different levels of experience in the field.
+Research products that are unapproachable counteract availability, and hinder reproducibility short and long term.
+To support long term adoption of reproducible practices in the natural sciences, research workflows need to be made approachable for learners, the researchers of the future \citep{roland2002think}.
 
-To elaborate on our thesis, we designed a case study within the research field of
-phylogenetics, a discipline within evolutionary biology.
-We use our case study to identify barriers that have made research workflows
-largely unapproachable to a general audience in the natural sciences.
-Then, we propose some principles for researchers to address these barriers, and
-create research workflows that are reproducible by a larger audience.
+To elaborate on our thesis, we designed a case study within the research field of phylogenetics, a discipline within evolutionary biology.
+We use our case study to identify barriers that have made research workflows largely unapproachable to a general audience in the natural sciences.
+Then, we propose some principles for researchers to address these barriers, and create research workflows that are reproducible by a larger audience.
 % ** describe/introduce reproducibility curriculum here
-The principles proposed here can be generalized and integrated into the undergraduate
-and graduate school STEM curriculum, either for courses specialized in reproducibility
-or within other subject areas, as a necessary component of successful and impactful science.
+The principles proposed here can be generalized and integrated into the undergraduate and graduate school STEM curriculum, either for courses specialized in reproducibility or within other subject areas, as a necessary component of successful and impactful science.
 
@@
 %%     \parbox{\textwidth}{%
-%%     \textbf{Box 1. The role of approachable research workflows in teacing reproducibility.}
+%%     \textbf{Box 1. The role of approachable research workflows in teaching reproducibility.}
 %%       % Availability is defined as the fact or possibility that something can be bought,
@@
 Phylogenetics is a key discipline within evolutionary biology \citep{dobzhansky1973nothing}.
-It focuses on investigating the history of shared ancestry of living and extinct
-organisms using biological data,
-and represents this evolutionary history with a diagram known as a phylogeny
-or phylogenetic tree (because it grows through time and appears to have branches;
-Figure \ref{fig:tree}).
-Phylogenies provide the basis to study and understand all biological processes
-in an evolutionary context \citep{dobzhansky1973nothing}.
-Hence, it appears that improving reproducibility rates in phylogenetics has the
-potential to positively impact research across the natural sciences.
+It focuses on investigating the history of shared ancestry of living and extinct organisms using biological data, and represents this evolutionary history with a diagram known as a phylogeny or phylogenetic tree (because it grows through time and appears to have branches; Figure \ref{fig:tree}).
+Phylogenies provide the basis to study and understand all biological processes in an evolutionary context \citep{dobzhansky1973nothing}.
+Hence, it follows that improving reproducibility rates in phylogenetics has the potential to positively impact research across the natural sciences.
 
@@
 
-To explore barriers to approachable phylogenetics, we develop a case study
-that touches on three common problems within the field: standardizing
-organism names in phylogenies, obtaining current phylogenetic knowledge for a group of organisms,
-and summarizing this phylogenetic knowledge in a meaningful way.
-To address these problems, we propose a research workflow that relies on resources from the Open
-Tree of Life (OpenTree), an open source project that provides
-digital availability of phylogenetic results from published, peer-reviewed research, which
-is considered as vetted and state-of-the-art knowledge in the field.
-OpenTree phylogenies are stored in a public database, the Phylesystem \citep{mctavish2015phylesystem},
-and are downloadable as various computer-readable file types, which is key for reusability
-and reproducible workflows \citep{wilson2017good}.
-OpenTree also provides access to a single standard for organism names (taxonomic standard)
-that is applied to the stored phylogenies \citep{rees2017automated}, which are
-then used to summarize a single phylogenetic tree encompassing all life \citep{opentreeoflife2019synth}.
+To explore barriers to approachable phylogenetics, we develop a case study that touches on three common problems within the field: standardizing organism names in phylogenies, obtaining current phylogenetic knowledge for a group of organisms, and summarizing this phylogenetic knowledge in a meaningful way.
+To address these problems, we propose a research workflow that relies on resources from the Open Tree of Life (OpenTree), an open source project that provides digital availability of phylogenetic results from published, peer-reviewed research, which is considered as vetted and state-of-the-art knowledge in the field.
+OpenTree phylogenies are stored in a public database, the Phylesystem \citep{mctavish2015phylesystem}, and are downloadable as various computer-readable file types, which is key for reusability and reproducible workflows \citep{wilson2017good}.
+OpenTree also provides access to a single standard for organism names (taxonomic standard) that is applied to the stored phylogenies \citep{rees2017automated}, which are then used to summarize a single phylogenetic tree encompassing all life \citep{opentreeoflife2019synth}.
 
@@
 are available for download and use from OpenTree free of cost to any user.
-One way to access OpenTree resources is through its Graphical User
-Interface (GUI; aka, a website or application that allows users to access computer
-functionalities, in this case OpenTree resources, with mouse or keyboard clicks).
-However, reducing as many manual steps as possible in research workflows is key
-for reproducibility, as manual data manipulation scales poorly and is prone to
-error \citep{bakken2019journey}.
-OpenTree's resources are also programmatically available through its Application
-Programming Interface services (APIs; aka, computer code that implements computer functionalities,
-in this case OpenTree resources, that can be used by programmers
-to build more functionalities),
-which provide scalability and reproducibility \citep{opentreeAPIv3}.
-However, this comes at a high cost for the user, which requires considerable more
-computer programming experience and literacy to be able to successfully use APIs.
+One way to access OpenTree resources is through its Graphical User Interface (GUI; aka, a website or application that allows users to access and use functionalities with mouse or keyboard clicks).
+However, reducing as many manual steps as possible in research workflows is key for reproducibility, as manual data manipulation scales poorly and is prone to error \citep{bakken2019journey}.
+OpenTree's resources are also programmatically available through its Application Programming Interface services (APIs; aka, computer code that implements computer functionalities, in this case OpenTree resources, that can be used by programmers
+to build more functionalities), which provide scalability and reproducibility \citep{opentreeAPIv3}.
+However, this comes at a high cost for the user, which requires considerably more computer programming experience and literacy to be able to successfully use APIs.
 %% In recent years, OpenTree API services have been wrapped into software packages for R and Python
@@
 %% OpenTree resources more accessible to a wider user audience.
-The \texttt{rotl} R package \citep{michonneau2016rotl} and the \texttt{opentree} Python module
-\citep{mctavish2021opentree} have been developed to wrap OpenTree's API services
+The \texttt{rotl} R package \citep{michonneau2016rotl} and the \texttt{opentree} Python module \citep{mctavish2021opentree} have been developed to wrap OpenTree's API services
 and make them available through the R and Python programming languages.
-R and Python are open source and free of cost, and represent
-two of the most widely used programming languages in the sciences today \citep{eglen2009quick,
-baker2017scientific}.
-As such, \texttt{rotl} and \texttt{opentree} software packages should contribute
-to making OpenTree resources more accessible to a wider programming user audience.
+R and Python are open source and free of cost, and represent two of the most widely used programming languages in the sciences today \citep{eglen2009quick, baker2017scientific}.
+As such, \texttt{rotl} and \texttt{opentree} software packages should contribute to making OpenTree resources more accessible to a wider programming user audience.
 
-However, while learners in the natural sciences have been engaging independently
-with R and Python programming languages, computer programming is not traditionally
-a core skill formally taught to biologists and naturalists
-\citep{sayres2018bioinformatics, wright2019the, williams2019barriers}.
-As computers continue to play a larger role in most scientific disciplines \citep{piccolo2016tools},
-higher baseline computational skills are required across all natural sciences not only
-to develop an original research workflow, but to be able to follow and reproduce
-research workflows from other researchers.
+However, while learners in the natural sciences have been engaging independently with R and Python programming languages, computer programming is not traditionally a core skill formally taught to biologists and naturalists \citep{sayres2018bioinformatics, wright2019the, williams2019barriers}.
+As computers continue to play a larger role in most scientific disciplines \citep{piccolo2016tools}, higher baseline computational skills are required across all natural sciences not only to develop an original research workflow, but to be able to follow and reproduce research workflows from other researchers.
 
-Thus, efforts to increase reproducibility rates long term in the natural sciences
-would benefit from addressing specific barriers for learners in the field, to
-support them in acquiring the skills needed to reproduce research workflows
-that rely heavily on computer code \citep{peng2011reproducible, sandve2013ten, powers2019open}.
+Thus, efforts to increase reproducibility rates long term in the natural sciences would benefit from addressing specific barriers for learners in the field, to support them in acquiring the skills needed to reproduce research workflows that rely heavily on computer code \citep{peng2011reproducible, sandve2013ten, powers2019open}.
 
-In the next section, we describe (in no particular order) the barriers to approachable research
-workflows that we identified on our case study.
-Then we develop a set of principles to address these barriers, and apply the latter
-to a set of teaching materials that are available at
-https://mctavishlab.github.io/R\_OpenTree\_tutorials/.
+In the next section, we describe (in no particular order) the barriers to approachable research workflows that we identified on our case study.
+Then we develop a set of principles to address these barriers, and apply the latter to a set of teaching materials that are available at https://mctavishlab.github.io/R\_OpenTree\_tutorials/.
 
@@
 
-The main goal of our case study is to obtain a single phylogeny summarizing
-data from a set of published phylogenies for the canids (the family of dogs, coyotes, wolves, etc.),
-our organism of study.
-All analysis for our case study can be completely accomplished using functions
-from the R package \texttt{rotl} or the Python module \texttt{opentree}. If
-a researcher were to use the proposed analysis workflow in a publication, they would typically
-describe it in the methodology section as ``The canid summary phylogeny was obtained using
-functions from X package, details are available as supplementary files".
-This is usual practice, mainly because journals do not have space to publish
-all code used for an analysis in the methods section. Yet, supplementary files and data have
-the misfortune to not be peer-reviewed as thoroughly (or at all) as the main manuscript
-\citep{pop2015use}.
-They are also prone to the dreaded promise ``available upon request", which has
-very low rates of fulfillment \citep{krawczyk2012available}.
-Without the primary data and code that was used to perform an analysis it is impossible
-to reproduce said analysis \citep{miyakawa2020no}. When the code is available, other issues can
-complicate reproduction of the analysis, to the point of completely obstructing
-reproducibility.
-For example, some questions that are often unaddressed in research workflows are:
-what software do I need to open the code files?
+The main goal of our case study is to obtain a single phylogeny summarizing data from a set of published phylogenies for the canids (the family of dogs, coyotes, wolves, etc.), our organism of study.
+All analysis for our case study can be completely accomplished using functions from the R package \texttt{rotl} or the Python module \texttt{opentree}. 
+If a researcher were to use the proposed analysis workflow in a publication, they would typically describe it in the methodology section as ``The canid summary phylogeny was obtained using functions from X package, details are available as supplementary files".
+This is usual practice, mainly because journals do not have space to publish all code used for an analysis in the methods section.
+Yet, supplementary files and data have the misfortune to not be peer-reviewed as thoroughly (or at all) as the main manuscript \citep{pop2015use}.
+They are also prone to the dreaded promise ``available upon request", which has very low rates of fulfillment \citep{krawczyk2012available}.
+Without the primary data and code that was used to perform an analysis it is impossible to reproduce said analysis \citep{miyakawa2020no}.
+When the code is available, other issues can complicate reproduction of the analysis, to the point of completely obstructing reproducibility.
+For example, some questions that are often unaddressed in research workflows are: what software do I need to open the code files?
 Can I run the code with the same software I used to open the files, or do I need a different one?
@@
 
-These questions can usually be answered by referring to the software documentation,
-which are usually publicly available and can be accessed by any potential user.
-As opposed to code, software documentation is written in natural language (i.e.,
-any known human language, e.g., English, Spanish, Chinese), and is considered
-a key element for successful adoption of software
-by target users \citep{karimzadeh2018top}. This might explain
-why documentation for software addressed to academic users is also
-usually written using highly specialized computational
-language or jargon (i.e., computationally specific concepts,
-words, and phrases) as well as formal scientific and academic language.
-We identify this as \textit{barrier 1} to approachable research workflows.--
-\textit{Specialized language is intimidating}.
-While scientific jargon might have an important role for formal acceptance
-of software by the scientific and academic community, it can be perceived as
-cold and/or intimidating language, that often slows down or even
-obstructs examination, application, and adoption of code by a wider audience
-\citep{ball2017its}.
-In contrast, introducing information without the use of jargon, supports learners'
-conceptual understanding of new ideas and concepts \citep{mcdonnell2016concepts, pan2019online}.
+These questions can usually be answered by referring to the software documentation, which are usually publicly available and can be accessed by any potential user.
+As opposed to code, software documentation is written in natural language (i.e., any known human language, e.g., English, Spanish, Chinese), and is considered a key element for successful adoption of software by target users \citep{karimzadeh2018top}.
+This might explain why documentation for software addressed to academic users is also usually written using highly specialized computational language or jargon (i.e., computationally specific concepts, words, and phrases) as well as formal scientific and academic language.
+We identify this as \textit{barrier 1} to approachable research workflows.-- \textit{Specialized language is intimidating}.
+While scientific jargon might have an important role for formal acceptance of software by the scientific and academic community, it can be perceived as cold and/or intimidating language, that often slows down or even obstructs examination, application, and adoption of code by a wider audience \citep{ball2017its}.
+In contrast, introducing information without the use of jargon, supports learners' conceptual understanding of new ideas and concepts \citep{mcdonnell2016concepts, pan2019online}.
 % jargon creates a hostile environment that does not foster learning \citep{CITATION}.
@@
 Another element of good software documentation is that it has to be thorough \citep{karimzadeh2018top}.
-Meaning that it should describe general usage of individual functions, as well as
-arguments and variables that said function can take \citep{karimzadeh2018top}.
-Individual documentation for each function is usually presented in alphabetic order,
-and does not have a specific analysis goal.
-Moreover, most software has numerous functions, so documentation is usually very
-lengthy and it is hard to navigate.
-This can have the effect of increasing the amount of information that needs to be
-simultaneously processed by the users, which can lead to overload of the finite amount
-of working memory any one possesses, known as cognitive load \citep{sweller1988cognitive}.
-In this context, identifying connections across functions that are meant to work
-on the same analysis workflow can become a very difficult task.
+Meaning that it should describe general usage of individual functions, as well as arguments and variables that said function can take \citep{karimzadeh2018top}.
+Individual documentation for each function is usually presented in alphabetic order, and does not have a specific analysis goal.
+Moreover, most software has numerous functions, so documentation is usually very lengthy and it is hard to navigate.
+This can have the effect of increasing the amount of information that needs to be simultaneously processed by the users, which can lead to overload of the finite amount of working memory any one possesses, known as cognitive load \citep{sweller1988cognitive}.
+In this context, identifying connections across functions that are meant to work on the same analysis workflow can become a very difficult task.
 We recognize this as \textit{barrier 2.-- Lack of specific goals leading to high cognitive load}.
-High cognitive load is know to have a negative effect in learning software
-\citep{chandler1996cognitive, van2005research, lambert2009student}.
+High cognitive load is know to have a negative effect in learning software \citep{chandler1996cognitive, van2005research, lambert2009student}.
 
-A third important aspect of software documentation are examples that demonstrate
-usage of individual functions \citep{karimzadeh2018top}.
-Examples presented in software documentation are usually
-worked to perfection, as they are intended to showcase the ideal or minimal case
-in which a function works well.
-Perfectly worked examples ignore the user experience by
-maintaining focus on the software content, and fail to provide users with expert
-and clear advice on how to troubleshoot if needed.
+A third important aspect of software documentation are examples that demonstrate usage of individual functions \citep{karimzadeh2018top}.
+Examples presented in software documentation are usually worked to perfection, as they are intended to showcase the ideal or minimal case in which a function works well.
+Perfectly worked examples ignore the user experience by maintaining focus on the software content, and fail to provide users with expert and clear advice on how to troubleshoot if needed.
 We identify this as \textit{barrier 3.-- Content-focused examples}.
-Error management training is an approach that focuses on framing mistakes as beneficial
-to learning complex tasks, to give learners the opportunity to
+Error management training is an approach that focuses on framing mistakes as beneficial to learning complex tasks, to give learners the opportunity to
 actively explore a task with a positive mindset \citep{frese1995error}.
-Providing examples that showcase potential errors supports users performance \citep{steele2014error},
-and can greatly improve learners ability to troubleshoot outside the classroom
-\citep{shannon2015live, nederbragt2020ten}.
+Providing examples that showcase potential errors supports users performance \citep{steele2014error}, and can greatly improve learners ability to troubleshoot outside the classroom \citep{shannon2015live, nederbragt2020ten}.
 
@@
 
-In sum, best practices for good software documentation are not enough to promote
-reproducibility of published research workflows that rely heavily on code. In the following
-section we describe some principles that can help to reduce or remove the
-identified barriers, to create
-research workflows that are more approachable and hence more reproducible by a
-larger audience.
+In sum, best practices for good software documentation are not enough to promote reproducibility of published research workflows that rely heavily on code. 
+In the following section we describe some principles that can help to reduce or remove the identified barriers, to create research workflows that are more approachable and hence more reproducible by a larger audience.
 
@@
 
-Besides avoiding formal language, and incorporating elements of pop culture, such as picture
-character icons known as ``emojis``, to make the language more familiar to a
-broader target audience (Figure \ref{fig:error}), we made an effort to specifically
-complement the primary documentation by identifying
-computational concepts that were assumed or were not explained in depth.
-We vetted the tutorials through feedback from workshop participants as well as
-individual users to identify such specialized concepts.
+Avoiding formal language, and incorporating elements of pop culture, such as picture character icons known as ``emojis``, make the language more familiar to a broader target audience (Figure \ref{fig:error}).
+We made an effort to specifically complement the primary documentation by identifying computational concepts that were assumed or were not explained in depth.
+We vetted the tutorials through feedback from workshop participants as well as individual users to identify such specialized concepts.
 
@@
 
-\subsubsection*{Principle 2. Reduce cognitive load by providing specific and clear goals
-with literate programming}
+\subsubsection*{Principle 2. Reduce cognitive load by providing specific and clear goals with literate programming}
 
-Cognitive load can be greatly reduced for learners by applying an active learning
-strategy such as linking usage to a ``real world'' or ``human'' application
-\citep{felder2009active}.
-Programming computer languages are by themselves quite abstract and represent a
-learning subject with a potentially high cognitive load for most learners.
-Pedagogical research shows that active learning practices are one of the most effective
-ways to take on abstract subjects \citep{freeman2014active}.
-A story-like narrative that links code usage in an integrative example, invites learners
-to try the code, which can lead them to remember what they are doing and
-why they are doing it.
-This ``literate programming'' paradigm \citep{knuth1984literate, fritzson2002mathmodelica}
-makes code more approachable, as it integrates narratives with computer code in
-the same document, supporting learners in actively following
-the code usage, supporting memory and understanding \citep{piccolo2016tools}.
+Cognitive load can be greatly reduced for learners by applying an active learning strategy such as linking usage to a ``real world'' or ``human'' application \citep{felder2009active}.
+Programming computer languages are by themselves quite abstract and represent a learning subject with a potentially high cognitive load for most learners.
+Pedagogical research shows that active learning practices are one of the most effective ways to take on abstract subjects \citep{freeman2014active}.
+A story-like narrative that links code usage in an integrative example, invites learners to try the code, which can lead them to remember what they are doing and why they are doing it.
+This ``literate programming'' paradigm \citep{knuth1984literate, fritzson2002mathmodelica} makes code more approachable, as it integrates narratives with computer code in the same document, supporting learners in actively following the code usage, supporting memory and understanding \citep{piccolo2016tools}.
 
-We propose that documents developed with ``literate programming'' can be made more
-accessible by choosing narratives that are relatable to a more general audience.
-An easy way to do this in biology is choosing a charismatic taxon as model organism.
-For a research group this can be the group they are studying. For the general audience,
-a highly charismatic group such as dinosaurs will do the trick.
-For example, when we presented our tutorial to the Amphibia Web Organization \citep{van2002amphibiaweb}
-in January 2020, we tailored all examples to frogs and their allies.
+We propose that documents developed with ``literate programming'' can be made more accessible by choosing narratives that are relatable to a more general audience.
+An easy way to do this in biology is choosing a charismatic taxon as a model organism.
+For a research group this can be the group they are studying. For the general audience, a highly charismatic group such as dinosaurs should work well.
+For example, when we presented our tutorial to the Amphibia Web Organization \citep{van2002amphibiaweb} in January 2020, we tailored all examples to frogs and their allies.
 % \citep{cite-aw-workshop}
 
-We examined available software documentation for the package \texttt{rotl},
-and designed a narrative that requires the usage of as many functions as possible.
-We demonstrate code applications that are commonly requested by OpenTree users,
-but that are not demonstrated in the documentation of the package.
-By framing the function workflow using highly requested uses, the documentation acquires a
-narrative arc that is easier to follow and remember by users. This can also facilitate
-translating the code application to other use cases of interest for learners in biology.
+We examined available software documentation for the package \texttt{rotl}, and designed a narrative that requires the usage of as many functions as possible.
+We demonstrate code applications that are commonly requested by OpenTree users, but that are not demonstrated in the documentation of the package.
+By framing the function workflow using highly requested uses, the documentation acquires a narrative arc that is easier to follow and remember by users.
+This can also facilitate translating the code application to other use cases of interest for learners in biology.
 
@@
 
-A practice that has become more and more widespread in programming-language pedagogical practices
-is live programming.
-One benefit of live programming is that typos and mistakes occur -- which normalizes
-them for learners, and shows them how to solve them when they are outside the
-classroom \citep{shannon2015live, nederbragt2020ten}.
-When co-author McTavish was a postdoc, teaching an introductory programming workshop as a volunteer
-with the Carpentries (a non-profit group that teaches foundational coding and
- data science skills to researchers worldwide) \citep{wilson2006swc, SWCwebsite},
-a senior faculty member taking the workshop
-complained that the typos were slowing things down and interfering with the pedagogy.
+An activity that has become increasingly widespread in programming-language education is live programming.
+During live programming, and instructor writes code and executes it in a way that is visible to all learners through a screen \citep{guzdial2013lure, selvaraj2021live}.
+One benefit of this practice is that typos and mistakes occur, normalizing them for learners. 
+Watching an instructor handling errors, demonstrates learners on how to solve them when they are outside theclassroom \citep{shannon2015live, nederbragt2020ten}.
+When co-author McTavish was a postdoc, teaching an introductory programming workshop as a volunteer with the Carpentries (a non-profit group that teaches foundational coding and data science skills to researchers worldwide) \citep{wilson2006swc, SWCwebsite},
+a senior faculty member taking the workshop complained that the typos were slowing things down and interfering with the pedagogy.
 McTavish replied "the typos ARE the pedagogy".
-This has become a slogan of sorts at the Carpentries, capturing the idea that embracing
-and discussing mistakes is essential to teaching programming \citep{wilson2019teaching}.
+This has become a slogan of sorts at the Carpentries, capturing the idea that embracing and discussing mistakes is essential to teaching programming \citep{wilson2019teaching}.
 Yet, working through mistakes is rarely done on written pedagogical materials.
-Software documentation focuses on demonstrating usage function with examples that
-work seamlessly, without errors. We argue that the opposite is needed to support
-adoption of reproducible workflows and support long term independence in learners
-and users performance \citep{gaspar2007restoring, steele2014error}.
+Software documentation focuses on demonstrating usage function with examples that work seamlessly, without errors. 
+We argue that the opposite is needed to support adoption of reproducible workflows and support long term independence in learners and users performance \citep{gaspar2007restoring, steele2014error}.
 
-In our tutorials, we apply this principle by demonstrating examples that do not work
-as expected, and exemplifying ways to address them (Figure \ref{fig:error}).
-For example, we identified inputs that would give
-a wide range of warnings and errors. We then focus on providing explanations for
-error and warning messages. We believe this supports users and learners to be less
-intimidated by the messages and practice taking useful information out of it.
+In our tutorials, we apply this principle by demonstrating examples that do not work as expected, and exemplifying ways to address them (Figure \ref{fig:error}).
+For example, we identified inputs that would give a wide range of warnings and errors.
+We then focus on providing explanations for these errors and warning messages.
+We believe this supports users and learners to be less intimidated by the messages, and to practice taking useful information out of them.
 % This is probably part of the computer science curriculum, but it is not at all
@@
 
-We also demonstrate ways to evaluate inputs to determine if they will trigger an
-error or warning, and design and demonstrate alternative analysis routes on what
-to do when faced with an error or warning.
-One of the most essential skills in programming is interpreting and moving forward
-from errors.
-This has two pedagogical benefits. First, it provides users and learners with the
-means to troubleshoot their own warnings and errors. Second, it allows them to
-understand with more depth what the function is doing.
+We also demonstrate ways to evaluate inputs to determine if they will trigger an error or warning, and design and demonstrate alternative analysis routes on what to do when faced with an error or warning.
+One of the most essential skills in programming is interpreting and moving forward from errors.
+This has two pedagogical benefits.
+First, it provides users and learners with the means to troubleshoot their own warnings and errors.
+Second, it allows them to understand with more depth what the function is doing.
 
@@
 
-Response from the community has been invaluable in gauging success of our teaching
-materials.
-Senior researchers often comment on the usefulness
-of the tutorials for their research, as well as how they have supported students in using the
-R packages more independently.
+Response from the community has been invaluable in gauging success of our teaching materials.
+Senior researchers often comment on the usefulness of the tutorials for their research, as well as how they have supported students in using the demosntrated R packages more independently.
 
@@
 but we followed basic recommendations for successful reproducibility \citep{sandve2013ten}.
-For example, we published the tutorials on a public, free license, free of cost, and free for
-use and reuse repository and persistent website \citep{RopentreeTutorials}.
-We also created a main version of the tutorial that is stable. Any updates to the tutorial are
-then published as new versions, or tutorials for new workshops \citep{wilson2006swc, SWCwebsite}
-Versions presented at workshops are a copy from the original repository, and
-constitute a temporally stable snapshot of functions and workflows presented
-during a workshop.
-This ensures that the tutorials are available for the users to return to any time
-they need it, and to be shared with other users and learners (Figure \ref{fig:schedule}).
+For example, we published the tutorials on a persistent and public website \citep{RopentreeTutorials} that adheres to the four r's of openness, by being free license, free of cost, and thus free for use, reuse, redistribute, revise and remix \citep{hilton2010four}.
+To make the website persistent, any updates to the tutorial are published as new versions.
+Versions presented at workshops are a copy from the original repository, and constitute a temporally stable snapshot of functions and workflows presented during a live workshop \citep{wilson2006swc, SWCwebsite}.
+This ensures that the tutorials are available for the users to return to any time they need it, and to be shared with other users and learners (Figure \ref{fig:schedule}).
+This approach not only teaches reproducibility, but also makes the teaching process itself reproducibile \citep{dogucu_tools_2022}.
 
@@
 
-We note that making approachable research workflows has all the advantages of
-reproducible research workflows.
-It saves explanation/training time when analyses are run again by students and
-collaborators.
-It can also save research time for yourself when analyses are run again with more
-data, a different dataset, a different organism or biological model;
+We note that making approachable research workflows has all the advantages of reproducible research workflows.
+It saves time during explanation and training, when analyses are run by new collaborators and students.
+It can also save research time for yourself, when analyses are run again with more data, a different dataset, a different organism or biological model.
 It contributes to scientific efforts that can build off of each other.
@@
 % \citep{roland2002think}
-Some universities have been incorporating reproducibility in their curriculum
-\citep{uwlibraries2022, nigms2022}.
-The focus of these resources has been for students to acquire and practice skills
-to document their work.
-The principles identified and outlined here can be used to set learning goals and
-outcomes for new reproducibility syllabi.
-Ultimately, the long term improvement of reproducibility rates in science will depend
-on our ability to intentionally integrate best practices for achieving reproducibility
-into the educational framework of future data scientists \citep{nasem2018data}.
-Inclusion of reproducibility into the data acumen of undergraduate curriculum will
-provide college learners and future researchers with the tools to develop
-the fundamental skills needed to successfully create reproducible scientific workflows
-and research products.
+Some universities and academic groups have started to incorporate reproducibility as a subject into their curriculum
+\citep{uwlibraries2022, nigms2022, psyteachr2022}.
+The focus of these resources has been for students to acquire and practice skills to document their work.
+The principles identified and outlined here can be used to set learning goals and outcomes for new reproducibility syllabi.
+Ultimately, the long term improvement of reproducibility rates in science will depend on our ability to intentionally integrate best practices for achieving reproducibility into the educational framework of future data scientists \citep{nasem2018data}.
+Inclusion of reproducibility into the data acumen of undergraduate curriculum will provide college learners and future researchers with the tools to develop the fundamental skills needed to successfully create reproducible scientific workflows and research products.
 
-The principles to create tutorials described here facilitate adoption of software
-and analysis workflows among researchers at different academic levels, from undergrads
-to established researchers.
-It will also help close the gap between students that had access to computational
-resources (and computational training) from an early age and students that did not.
-Late access to computational resources and training can occur due to lack of
-economic resources, often occurring in households from underrepresented communities
-and minorities \citep{google2016diversity, warner2021quantifying}.
-It can also be due to gender-biased parental and community pressures,
-in which male individuals are more often encouraged to perform activities related to computers,
-while female individuals are discouraged \citep{warner2021quantifying}.
-These principles can be used to improve not only reproducibility practices,
-but also software adoption in the natural sciences.
+The principles to create tutorials described here facilitate adoption of software and analysis workflows among researchers at different academic levels, from undergrads to established researchers.
+It will also help close the gap between students that had access to computational resources (and computational training) from an early age and students that did not.
+Late access to computational resources and training can occur due to lack of economic resources, often occurring in households from underrepresented communities and minorities \citep{google2016diversity, warner2021quantifying}.
+It can also be due to gender-biased parental and community pressures, in which male identifying individuals are more often encouraged to perform activities related to computers, while female identifying individuals are discouraged, starting from as early as elementray school \citep{master_gender_2021}.
+These principles can be used to improve not only reproducibility practices, but also software adoption in the natural sciences.
 

### 2022-06-10 jsdse/Manuscript.tex
fixing article and book title style on bib file

@@
 Research reproducibility is essential for scientific development. Yet, rates of reproducibility are low, especially in the natural sciences.
-As increasingly more research relies on computers and software, efforts for improving reproducibility rates have focused on making available research workflows as computer code, as well as raw and processed data in computer readable form.
+As increasingly more research relies on computers and software, efforts for improving reproducibility rates have focused on making research products digitally available, such as publishing analysis workflows as computer code, and raw and processed data in computer readable form.
 However, research products that are digitally available are not necessarily friendly for learners and interested parties with little to no experience in the field.
-This renders research products unapproachable, counteracts availability, and hinders reproducibility short and long term. 
-To improve long term adoption of reproducible workflows in research, they need to be made approachable for learners, the researchers of the future.
+This renders research products unapproachable, counteracts their availability, and hinders scientific reproducibility.
+To improve both short and long term adoption of reproducible practices in research, research products need to be made approachable for learners, the researchers of the future.
 
@@
 content-focused descriptions.
-We propose a set of principles to improve the unapproachable aspects of research workflows, and illustrate their application in a tutorial. 
-We elaborate on the general application of these principles for documenting research workflows and products, to provide present learners and future researchers with tools for successful scientific reproducibility.
+We propose a set of principles to improve the unapproachable aspects of research workflows, and illustrate their application using an online teaching resource.
+We elaborate on the general application of these principles for documenting not only research products but also teaching materials, to provide present learners and future researchers with tools for successful scientific reproducibility.
 %> ** add a sentence on implications for teaching reproducibility
-
-%% Projects that address availability of code and data in the natural sciences
-%% have focused on producing documentation that describes the nature and usage of the
-%% resources.
-%% However, it often communicates in a highly specialized language that is largely
-%% inaccessible for the average target user. Additionally, baseline
-%% computational knowledge and required skills to access scientific results are often
-%% not found among potential users, and keep on increasing.
 
@@
 \begin{center}
-\includegraphics[width=3in]{fig-tree.png}
+\includegraphics[width=3in]{fig_tree.png}
 \end{center}
@@
 When the code is available, other issues can complicate reproduction of the analysis, to the point of completely obstructing reproducibility.
-For example, some questions that are often unaddressed in research workflows are: what software do I need to open the code files?
-Can I run the code with the same software I used to open the files, or do I need a different one?
-Do I need additional software that the analysis depends on?
-What does the code even mean?
+For example, some questions that are often unaddressed in research workflows are:
+\textit{what software do I need to open the code files?}
+\textit{Can I run the code with the same software I used to open the files, or do I need a different one?}
+\textit{Do I need additional software that the analysis depends on?}
+\textit{What does the code even mean?}
 
@@
 While scientific jargon might have an important role for formal acceptance of software by the scientific and academic community, it can be perceived as cold and/or intimidating language, that often slows down or even obstructs examination, application, and adoption of code by a wider audience \citep{ball2017its}.
-In contrast, introducing information without the use of jargon, supports learners' conceptual understanding of new ideas and concepts \citep{mcdonnell2016concepts, pan2019online}.
+In contrast, introducing information without the use of jargon, supports learner's conceptual understanding of new ideas and concepts \citep{mcdonnell2016concepts, pan2019online}.
 % jargon creates a hostile environment that does not foster learning \citep{CITATION}.
@@
 actively explore a task with a positive mindset \citep{frese1995error}.
-Providing examples that showcase potential errors supports users performance \citep{steele2014error}, and can greatly improve learners ability to troubleshoot outside the classroom \citep{shannon2015live, nederbragt2020ten}.
+Providing examples that showcase potential errors supports user's performance \citep{steele2014error}, and can greatly improve learner's ability to troubleshoot outside the classroom \citep{shannon2015live, nederbragt2020ten}.
 
@@
 \begin{center}
-\includegraphics[width=3in]{fig-error.png}
+\includegraphics[width=3in]{fig_error.png}
 \end{center}
@@
 An easy way to do this in biology is choosing a charismatic taxon as a model organism.
-For a research group this can be the group they are studying. For the general audience, a highly charismatic group such as dinosaurs should work well.
+For a research group, this can be the biological group they are studying. For the general audience, a highly charismatic group --such as dinosaurs, should work well.
 For example, when we presented our tutorial to the Amphibia Web Organization \citep{van2002amphibiaweb} in January 2020, we tailored all examples to frogs and their allies.
@@
 An activity that has become increasingly widespread in programming-language education is live programming.
-During live programming, and instructor writes code and executes it in a way that is visible to all learners through a screen \citep{guzdial2013lure, selvaraj2021live}.
+During live programming, an instructor writes code and executes it in a way that is visible to learners through a screen \citep{guzdial2013lure, selvaraj2021live}.
 One benefit of this practice is that typos and mistakes occur, normalizing them for learners.
-Watching an instructor handling errors, demonstrates learners on how to solve them when they are outside theclassroom \citep{shannon2015live, nederbragt2020ten}.
+Watching an instructor handling errors, demonstrates learners on how to solve them when they are outside the classroom \citep{shannon2015live, nederbragt2020ten}.
 When co-author McTavish was a postdoc, teaching an introductory programming workshop as a volunteer with the Carpentries (a non-profit group that teaches foundational coding and data science skills to researchers worldwide) \citep{wilson2006swc, SWCwebsite},
@@
 Software documentation focuses on demonstrating usage function with examples that work seamlessly, without errors.
-We argue that the opposite is needed to support adoption of reproducible workflows and support long term independence in learners and users performance \citep{gaspar2007restoring, steele2014error}.
+We argue that the opposite is needed to support adoption of reproducible workflows and support long term independence in learner's and user's performance \citep{gaspar2007restoring, steele2014error}.
 
-In our tutorials, we apply this principle by demonstrating examples that do not work as expected, and exemplifying ways to address them (Figure \ref{fig:error}).
+In our tutorials, we apply this principle by demonstrating examples that do not work as expected, and exemplifying ways to address issues (Figure \ref{fig:error}).
 For example, we identified inputs that would give a wide range of warnings and errors.
@@
 \begin{center}
-\includegraphics[width=3in]{fig-schedule.png}
+\includegraphics[width=3in]{fig_schedule.png}
 \end{center}
 \caption{Snapshot of the home to our tutorial website, showing part of the schedule.
- Our tutorial website was constructed using The Software Carpentries workshop template \citep{swc-workshop-template}. \label{fig:schedule}}
+ Our tutorial website was constructed using the software workshop template from the Carpentries \citep{swc-workshop-template}. \label{fig:schedule}}
 \end{figure}

### 2022-06-10 jsdse/reviews/response_2022_06_09.Rmd
response to reviews

---
output: rmarkdown::pdf_document
header-includes:
    - \usepackage{setspace}\doublespacing
    - \usepackage{float}
geometry: margin=1.2in
fontsize: 12pt
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

January 10, 2022

University of California, Merced

&nbsp;  &nbsp;

Dear Dr. Horton, associate editor, and reviewers,

&nbsp;

We appreciate the helpful and thoughtful comments made to our revised manuscript. We are excited to learn that the manuscript is now suitable for publication, subject to minor revisions.

Please receive a newly revised version of our manuscript, formerly titled *“Promoting analysis reproducibility with accessibility: An example in evolutionary biology”*, and now titled **“Approachable case studies support learning and reproducibility in data science: An example from evolutionary biology”**.

We have incorporated into the revised manuscript all comments and suggestions from you, associate editor and 2 reviewers. We believe that all of them have greatly improved the manuscript. We hope that you find it ready for publication in the special issue about teaching reproducibility and responsible workflow.

We address individual comments line by line below.

&nbsp;  &nbsp;

Please receive our respectful regards,

&nbsp;  &nbsp;

The authors

\pagebreak

\singlespacing

*Comments from the Editors and Reviewers:*

*Editor*
*Comments to the Author*

*In addition to the comments from the reviewers and associate editor, I had the following additional points:*

*1. The abstract (in the metadata associated with the paper) is missing spacing in many places (e.g., "reproducibilityare". This is likely due to cut and paste issues. Can you please review and fix as this will promulgate through the production process.*

Thanks. We reviewed and fixed missing spaces across (hopefully) the entire manuscript.

&nbsp;  &nbsp;

*2. It might be appropriate to consider citing this in-process work (https://arxiv.org/abs/2202.09504)*

We agree. We have cited this work...

&nbsp;  &nbsp;

*3. page 2, end of third paragraph: would it be useful to cite the NASEM 2018 (Data Science for Undergraduates) report, which includes reproducibility and workflow as important components of "data acumen"? In addition, would citation of the NASEM 2019 Reproducibility report be appropriate?*

Yes, after pondering your question, we concluded that the NASEM 2018 report should also be cited in this section. The NASEM 2019, I am not completely sure yet...

&nbsp;  &nbsp;

*4. Note style for journal (Taylor & Francis) is to capitalize article and book titles*

Thanks. We went over the entire bib file and capitalized article and book titles. Hopefully they are all appropriately styled now.

&nbsp;  &nbsp;

*********************

*Associate Editor*
*Comments to the Author:*
*Thank you for reviewing the manuscript according to reviewers feedback. One issue which is pointed by one reviewer is mentioning package version - which is your first section in the online resource you discuss, so it will be easy to incorporate.*

Thank you! We addressed that on section...

&nbsp;  &nbsp;

*In the Conclusions when you provide examples of some institutions adopting reproducibility as a main teaching principle, you could also mention University of Glasgow School of Psychology and Neuroscience - https://psyteachr.github.io/*

Thanks for pointing us out to this resource. We thought it was appropriate, so we added psyteachR to the examples.

&nbsp;  &nbsp;

*********************

*Reviewer: 1*

*Comments to the Author*
*The authors have addressed all the issues I had raised about the initial submission.*

We really appreciate all comments and suggestions made on the initial review!

&nbsp;  &nbsp;

*********************

*Reviewer: 2*

*Comments to the Author*
*A significant barrier not discussed here is the deprecation of functions in open software packages like R and Python. It often requires considerable programming savvy to repair code that has fallen victim to deprecation.*

Thank you, the point that you raise is very important. Associate Editor correctly points out that this can be addressed by providing information on package version used. In general, old package versions remain accessible, and persist in time, so they can be used to reproduce an analysis, even long time after a function had been deprecated. 
The teaching materials that we use as example include a section that addresses documenting package versions used for an analysis.

We elaborate on this barrier on section...

&nbsp;  &nbsp;

*Line 38: considerably* - Fixed

*Line 54: “…as a model…”* - Fixed

*Line 30: Define “live programming”* - Added a sentence defining it.

*Line 37: This is called Carpentries here but Software Carpentries in Figure 3’s caption. Are these the same thing?*  - Software Carpentries is a component of the Carpentries, we clarified this in the Figure caption.

*Line 53 and elsewhere: “learner’s” and “user’s”.* - Fixed

*Line 12: Warner et al. 2021 do not make this claim in their paper.* - You are right, we meant...

*********************

### 2022-06-10 jsdse/reviews/response_2022_06_09.Rmd
response to reviews mostly done!

@@
 
-Yes, after pondering your question, we concluded that the NASEM 2018 report should also be cited in this section. The NASEM 2019, I am not completely sure yet...
+Yes, after pondering your question, we concluded that the NASEM 2018 report should also be cited in this section.
+
+The NASEM 2019 report is very interesting and addresses many important points relevant for this work.
+We cite it at the end of first paragraph, Page 5: "As computers continue to play a larger role in most scientific disciplines (Piccolo & Frampton 2016), higher baseline computational skills are required across all natural sciences not only to develop an original research workflow, but to be able to follow and reproduce research workflows from other researchers (NASEM, 2019)"
 
@@
 
-*Line 12: Warner et al. 2021 do not make this claim in their paper.* - You are right, we meant...
+*Line 12: Warner et al.
+2021 do not make this claim in their paper.* - You are right, we changed the sentence accordingly to: "[The principles described here] can also help close the academic gap that is generated by uneven access to computational resources across students belonging to different groups (KewalRamani et al.
+2018), where the most affected learners usually belong to underrepresented minorities and rural areas (Warner et al.
+2021)."
 

### 2022-06-10 jsdse/Manuscript.tex
response to reviews mostly done!

@@
 Research products that are unapproachable counteract availability, and hinder reproducibility short and long term.
-To support long term adoption of reproducible practices in the natural sciences, research workflows need to be made approachable for learners, the researchers of the future \citep{roland2002think}.
+To support long term adoption of reproducible practices in the natural sciences, research workflows need to be made approachable for learners, the researchers of the future \citep{roland2002think, nasem2018data}.
 
@@
 Then, we propose some principles for researchers to address these barriers, and create research workflows that are reproducible by a larger audience.
-% ** describe/introduce reproducibility curriculum here
+% ** describe/introduce reproducibility curriculum here:
 The principles proposed here can be generalized and integrated into the undergraduate and graduate school STEM curriculum, either for courses specialized in reproducibility or within other subject areas, as a necessary component of successful and impactful science.
@@
 However, while learners in the natural sciences have been engaging independently with R and Python programming languages, computer programming is not traditionally a core skill formally taught to biologists and naturalists \citep{sayres2018bioinformatics, wright2019the, williams2019barriers}.
-As computers continue to play a larger role in most scientific disciplines \citep{piccolo2016tools}, higher baseline computational skills are required across all natural sciences not only to develop an original research workflow, but to be able to follow and reproduce research workflows from other researchers.
+As computers continue to play a larger role in most scientific disciplines \citep{piccolo2016tools}, higher baseline computational skills are required across all natural sciences not only to develop an original research workflow, but to be able to follow and reproduce research workflows from other researchers \citep{nasem2019reproducibility}.
 
@@
 They are also prone to the dreaded promise ``available upon request", which has very low rates of fulfillment \citep{krawczyk2012available}.
-Without the primary data and code that was used to perform an analysis it is impossible to reproduce said analysis \citep{miyakawa2020no}.
+Without the primary data and code that was used to perform an analysis, it is impossible to reproduce said analysis \citep{miyakawa2020no}.
 When the code is available, other issues can complicate reproduction of the analysis, to the point of completely obstructing reproducibility.
@@
 actively explore a task with a positive mindset \citep{frese1995error}.
-Providing examples that showcase potential errors supports user's performance \citep{steele2014error}, and can greatly improve learner's ability to troubleshoot outside the classroom \citep{shannon2015live, nederbragt2020ten}.
+Providing examples that showcase potential errors, supports user's performance \citep{steele2014error}, and can greatly improve learner's ability to troubleshoot outside the classroom \citep{shannon2015live, nederbragt2020ten}.
 
@@
 
-The principles to create tutorials described here facilitate adoption of software and analysis workflows among researchers at different academic levels, from undergrads to established researchers.
-It will also help close the gap between students that had access to computational resources (and computational training) from an early age and students that did not.
-Late access to computational resources and training can occur due to lack of economic resources, often occurring in households from underrepresented communities and minorities \citep{google2016diversity, warner2021quantifying}.
-It can also be due to gender-biased parental and community pressures, in which male identifying individuals are more often encouraged to perform activities related to computers, while female identifying individuals are discouraged, starting from as early as elementray school \citep{master_gender_2021}.
-These principles can be used to improve not only reproducibility practices, but also software adoption in the natural sciences.
+The principles described here are relevant to facilitate adoption of software and analysis workflows among researchers at different academic levels, from undergrads to established researchers.
+They can also help close the academic gap that is generated by uneven access to computational resources across students belonging to different groups \citep{kewalramani2018student}, where the most affected learners usually belong to underrepresented minorities and rural areas \citep{warner2021quantifying}.
+Differences in access can also be due to gender-biased parental and community pressures, in which male identifying individuals are more likely to be encouraged to perform activities related to computers \citep{google2016diversity}, while female identifying individuals are discouraged, starting from as early as elementray school \citep{master_gender_2021}.
+These principles can be used to improve not only reproducibility practices, but also ease adoption of computational resources in the natural sciences.
 

### 2022-06-11 jsdse/reviews/response_2022_06_09.Rmd
addressing EJM's comments!

@@
 
-We agree. We have cited this work...
+We agree.
+We have cited this work at the end of the first paragraph of the Conlcusion section: "These approaches not only teach reproducibility, but also make the teaching process itself reproducible (Dogucu & Cetinkaya-Rundel 2022).
 
@@
 
-Thank you! We addressed that on section...
+Thank you! We addressed that on the first paragraph of the Conclusion section.
 
@@
 Thank you, the point that you raise is very important.
-Associate Editor correctly points out that this can be addressed by providing information on package version used.
+Associate Editor highlighted your comment and points out that this can be addressed by providing information on package version used.
 In general, old package versions remain accessible, and persist in time, so they can be used to reproduce an analysis, even long time after a function had been deprecated.
@@
 
-We elaborate on this barrier on section...
+We now discuss this barrier on the first paragraph of the Conclusion section.
 

### 2022-06-11 jsdse/Manuscript.tex
addressing EJM's comments!

@@
 \begin{abstract}
-Research reproducibility is essential for scientific development. Yet, rates of reproducibility are low, especially in the natural sciences.
+Research reproducibility is essential for scientific development. Yet, rates of reproducibility are low.
 As increasingly more research relies on computers and software, efforts for improving reproducibility rates have focused on making research products digitally available, such as publishing analysis workflows as computer code, and raw and processed data in computer readable form.
@@
 This renders research products unapproachable, counteracts their availability, and hinders scientific reproducibility.
-To improve both short and long term adoption of reproducible practices in research, research products need to be made approachable for learners, the researchers of the future.
+To improve both short and long term adoption of reproducible scientific practices, research products need to be made approachable for learners, the researchers of the future.
 
@@
 unclear goals and high cognitive load; and
-content-focused descriptions.
+lack of trouble-shooting examples.
 We propose a set of principles to improve the unapproachable aspects of research workflows, and illustrate their application using an online teaching resource.
@@
 To address these problems, we propose a research workflow that relies on resources from the Open Tree of Life (OpenTree), an open source project that provides digital availability of phylogenetic results from published, peer-reviewed research, which is considered as vetted and state-of-the-art knowledge in the field.
-OpenTree phylogenies are stored in a public database, the Phylesystem \citep{mctavish2015phylesystem}, and are downloadable as various computer-readable file types, which is key for reusability and reproducible workflows \citep{wilson2017good}.
+OpenTree phylogenies are stored in a public database, the Phylesystem \citep{mctavish2015phylesystem}, and are downloadable as various computer-readable file types, which is key for reusable and reproducible workflows \citep{wilson2017good}.
 OpenTree also provides access to a single standard for organism names (taxonomic standard) that is applied to the stored phylogenies \citep{rees2017automated}, which are then used to summarize a single phylogenetic tree encompassing all life \citep{opentreeoflife2019synth}.
@@
 %% tools developed to process and summarize stored phylogenies),
-are available for download and use from OpenTree free of cost to any user.
-One way to access OpenTree resources is through its Graphical User Interface (GUI; aka, a website or application that allows users to access and use functionalities with mouse or keyboard clicks).
+are available for download and use from OpenTree, free of financial cost to any user.
+One way to access OpenTree resources is manually, through its Graphical User Interface (GUI; aka, a website or application that allows users to access and use functionalities with mouse or keyboard clicks).
 However, reducing as many manual steps as possible in research workflows is key for reproducibility, as manual data manipulation scales poorly and is prone to error \citep{bakken2019journey}.
-OpenTree's resources are also programmatically available through its Application Programming Interface services (APIs; aka, computer code that implements computer functionalities, in this case OpenTree resources, that can be used by programmers
-to build more functionalities), which provide scalability and reproducibility \citep{opentreeAPIv3}.
-However, this comes at a high cost for the user, which requires considerably more computer programming experience and literacy to be able to successfully use APIs.
+OpenTree's resources are also programmatically available through its Application Programming Interface services (APIs; aka, computer code that automatically implements functionalities, that is usually used by programmers to build different or tailored functionalities).
+While APIs provide data processing scalability and reproducibility \citep{opentreeAPIv3}, they come at a high technical and cognitive cost for the user, whom requires considerably more computer programming experience and literacy to be able to successfully utilize APIs.
 %% In recent years, OpenTree API services have been wrapped into software packages for R and Python
@@
 %% OpenTree resources more accessible to a wider user audience.
-The \texttt{rotl} R package \citep{michonneau2016rotl} and the \texttt{opentree} Python module \citep{mctavish2021opentree} have been developed to wrap OpenTree's API services
-and make them available through the R and Python programming languages.
-R and Python are open source and free of cost, and represent two of the most widely used programming languages in the sciences today \citep{eglen2009quick, baker2017scientific}.
-As such, \texttt{rotl} and \texttt{opentree} software packages should contribute to making OpenTree resources more accessible to a wider programming user audience.
+OpenTree's API services have been wrapped by the \texttt{rotl} R package \citep{michonneau2016rotl} and the \texttt{opentree} Python module \citep{mctavish2021opentree}.
+R and Python programming languages are open source and free of cost, and represent two of the most widely used programming languages in the sciences today \citep{eglen2009quick, baker2017scientific}.
+As such, \texttt{rotl} and \texttt{opentree} software packages are contributing to approachability of OpenTree's resources to R and Python users, increasing availability to a wider user base.
 
@@
 
-In the next section, we describe (in no particular order) the barriers to approachable research workflows that we identified on our case study.
+In the next section, we describe (in no particular order) three barriers to approachable research workflows that we identify using our case study.
 Then we develop a set of principles to address these barriers, and apply the latter to a set of teaching materials that are available at https://mctavishlab.github.io/R\_OpenTree\_tutorials/.
@@
 
-The main goal of our case study is to obtain a single phylogeny summarizing data from a set of published phylogenies for the canids (the family of dogs, coyotes, wolves, etc.), our organism of study.
+The main goal of our case study is to obtain a single phylogeny summarizing data from a set of published phylogenies for the canids (the family of dogs, coyotes, wolves, etc.), our organisms of study.
 All analysis for our case study can be completely accomplished using functions from the R package \texttt{rotl} or the Python module \texttt{opentree}.
@@
 Yet, supplementary files and data have the misfortune to not be peer-reviewed as thoroughly (or at all) as the main manuscript \citep{pop2015use}.
-They are also prone to the dreaded promise ``available upon request", which has very low rates of fulfillment \citep{krawczyk2012available}.
+They are also prone to the dreaded promise ``available upon request", which has very low rates of fulfillment \citep{krawczyk2012available, gabelica2022many}.
 Without the primary data and code that was used to perform an analysis, it is impossible to reproduce said analysis \citep{miyakawa2020no}.
@@
 For example, some questions that are often unaddressed in research workflows are:
-\textit{what software do I need to open the code files?}
-\textit{Can I run the code with the same software I used to open the files, or do I need a different one?}
+\textit{what software do I need to read the code files?}
+\textit{What software can I use to actually run the code?}
 \textit{Do I need additional software that the analysis depends on?}
+\textit{What software versions ere used?}
 \textit{What does the code even mean?}
 
-These questions can usually be answered by referring to the software documentation, which are usually publicly available and can be accessed by any potential user.
+Some of these questions can be answered by referring to the software documentation, which is usually publicly available and can be accessed by any potential user.
 As opposed to code, software documentation is written in natural language (i.e., any known human language, e.g., English, Spanish, Chinese), and is considered a key element for successful adoption of software by target users \citep{karimzadeh2018top}.
@@
 Perfectly worked examples ignore the user experience by maintaining focus on the software content, and fail to provide users with expert and clear advice on how to troubleshoot if needed.
-We identify this as \textit{barrier 3.-- Content-focused examples}.
-Error management training is an approach that focuses on framing mistakes as beneficial to learning complex tasks, to give learners the opportunity to
-actively explore a task with a positive mindset \citep{frese1995error}.
+We identify this as \textit{barrier 3.-- Lack of trouble-shooting examples}.
+Error management training is an approach that focuses on framing mistakes as beneficial to learning complex tasks, to give learners the opportunity to actively explore a task with a positive mindset \citep{frese1995error}.
 Providing examples that showcase potential errors, supports user's performance \citep{steele2014error}, and can greatly improve learner's ability to troubleshoot outside the classroom \citep{shannon2015live, nederbragt2020ten}.
@@
 
-Avoiding formal language, and incorporating elements of pop culture, such as picture character icons known as ``emojis``, make the language more familiar to a broader target audience (Figure \ref{fig:error}).
+Avoiding formal language, and incorporating elements of pop culture, such as picture character icons known as ``emojis", make the language more familiar to a broader target audience (Figure \ref{fig:error}).
 We made an effort to specifically complement the primary documentation by identifying computational concepts that were assumed or were not explained in depth.
@@
 a senior faculty member taking the workshop complained that the typos were slowing things down and interfering with the pedagogy.
-McTavish replied "the typos ARE the pedagogy".
+McTavish replied ``the typos ARE the pedagogy".
 This has become a slogan of sorts at the Carpentries, capturing the idea that embracing and discussing mistakes is essential to teaching programming \citep{wilson2019teaching}.
@@
 First, it provides users and learners with the means to troubleshoot their own warnings and errors.
-Second, it allows them to understand with more depth what the function is doing.
+Second, it allows them to understand with more depth what the code is doing.
 
@@
 Response from the community has been invaluable in gauging success of our teaching materials.
-Senior researchers often comment on the usefulness of the tutorials for their research, as well as how they have supported students in using the demosntrated R packages more independently.
+Senior researchers often comment on the usefulness of the tutorials for their research, as well as how they have supported students in using the demonstrated R packages more independently.
+We note that making approachable research workflows has all the advantages of reproducible research workflows.
+It saves time during explanation and training, when analyses are run by new collaborators and students.
+It can also save research time for yourself, when analyses are run again with more data, a different dataset, a different organism or biological model.
+It contributes to making scientific efforts that can build off of each other.
 
-When developing our tutorials, we not only applied the principles elaborated here,
+When developing our tutorials, we not only applied the principles elaborated here to make them more approachable,
 but we followed basic recommendations for successful reproducibility \citep{sandve2013ten}.
@@
 This ensures that the tutorials are available for the users to return to any time they need it, and to be shared with other users and learners (Figure \ref{fig:schedule}).
-This approach not only teaches reproducibility, but also makes the teaching process itself reproducibile \citep{dogucu_tools_2022}.
+Finally, we dedicate a section of the tutorial to document the software versions that are demonstrated throughout. 
+A common issue in open source software packages written in R and Python is the deprecation of functions (i.e., functions that are no longer reccommended or maintained, and that are in the process of being phased out and replaced by new ones; \citealt{deprecation2017, vadlamani2021apiscanner}).
+Running code that has fallen victim to deprecation require considerable programming savvy, and it is a task that is hard even for experienced programmers \citep{vadlamani2021apiscanner}.
+Yet, old versions of open source software packages remain digitally accessible, and persist in time.
+If a package version is known, an analysis can be reproduced without investing additional resources in finding the appropriate functions to run it, even long time after a function had been deprecated.
+Applying these principles in a tutorial not only teaches reproducibility, but also makes the teaching process itself reproducible \citep{dogucu_tools_2022}.
 
@@
 
-We note that making approachable research workflows has all the advantages of reproducible research workflows.
-It saves time during explanation and training, when analyses are run by new collaborators and students.
-It can also save research time for yourself, when analyses are run again with more data, a different dataset, a different organism or biological model.
-It contributes to scientific efforts that can build off of each other.
+Incorporating the principles described here into teaching resources not onlu improves reproducibility practices, but it should facilitate adoption of software and analysis workflows in the natural sciences, among researchers at different academic levels, from undergrads to established researchers.
+It can also help close the academic gap that is generated by uneven access to computational resources across students belonging to different groups \citep{kewalramani2018student}, where the most affected learners usually belong to underrepresented minorities and rural areas \citep{warner2021quantifying}.
+Differences in access can also be due to gender-biased parental and community pressures, in which male identifying individuals are more likely to be encouraged to perform activities related to computers \citep{google2016diversity}, while female identifying individuals are discouraged, starting from as early as elementray school \citep{master_gender_2021}.
 
@@
 Inclusion of reproducibility into the data acumen of undergraduate curriculum will provide college learners and future researchers with the tools to develop the fundamental skills needed to successfully create reproducible scientific workflows and research products.
-
-The principles described here are relevant to facilitate adoption of software and analysis workflows among researchers at different academic levels, from undergrads to established researchers.
-They can also help close the academic gap that is generated by uneven access to computational resources across students belonging to different groups \citep{kewalramani2018student}, where the most affected learners usually belong to underrepresented minorities and rural areas \citep{warner2021quantifying}.
-Differences in access can also be due to gender-biased parental and community pressures, in which male identifying individuals are more likely to be encouraged to perform activities related to computers \citep{google2016diversity}, while female identifying individuals are discouraged, starting from as early as elementray school \citep{master_gender_2021}.
-These principles can be used to improve not only reproducibility practices, but also ease adoption of computational resources in the natural sciences.
 

### 2022-06-11 jsdse/reviews/response_2022_06_09.Rmd
updating manuscript pdf and replies to reviews

--- 2022-06-10 jsdse/reviews/response_2022_06_09.Rmd
@@
 
-We agree. We have cited this work...
+We agree.
+We have cited this work at the beginning of the second paragraph of the Conclusion section: "These approaches not only teach reproducibility, but also make the teaching process itself reproducible (Dogucu & Cetinkaya-Rundel 2022).
 
@@
 
-Thank you! We addressed that on section...
+Thank you! We addressed that on the second paragraph of the Conclusion section.
 
@@
 Thank you, the point that you raise is very important.
-Associate Editor correctly points out that this can be addressed by providing information on package version used.
+Associate Editor highlighted your comment and points out that this can be addressed by providing information on package version used; we agree.
 In general, old package versions remain accessible, and persist in time, so they can be used to reproduce an analysis, even long time after a function had been deprecated.
@@
 
-We elaborate on this barrier on section...
+We now discuss this barrier and suggest ways to address it at the end of the second paragraph of the Conclusion section.
 

### 2022-06-11 jsdse/Manuscript.tex
updating manuscript pdf and replies to reviews

@@
 Watching an instructor handling errors, demonstrates learners on how to solve them when they are outside the classroom \citep{shannon2015live, nederbragt2020ten}.
-When co-author McTavish was a postdoc, teaching an introductory programming workshop as a volunteer with the Carpentries (a non-profit group that teaches foundational coding and data science skills to researchers worldwide) \citep{wilson2006swc, SWCwebsite},
+When co-author McTavish was a postdoc, teaching an introductory programming workshop as a volunteer with the Carpentries (a non-profit group that teaches foundational coding and data science skills to researchers worldwide; \citealt{wilson2006swc, SWCwebsite}),
 a senior faculty member taking the workshop complained that the typos were slowing things down and interfering with the pedagogy.
@@
 
-When developing our tutorials, we not only applied the principles elaborated here to make them more approachable,
-but we followed basic recommendations for successful reproducibility \citep{sandve2013ten}.
+When developing our tutorials, we not only applied the principles elaborated here to make them more approachable, but we followed basic recommendations for successful reproducibility \citep{sandve2013ten}.
+Applying all these principles in a tutorial not only teaches reproducibility, but also makes the teaching process itself reproducible \citep{dogucu_tools_2022}.
 For example, we published the tutorials on a persistent and public website \citep{RopentreeTutorials} that adheres to the four r's of openness, by being free license, free of cost, and thus free for use, reuse, redistribute, revise and remix \citep{hilton2010four}.
@@
 If a package version is known, an analysis can be reproduced without investing additional resources in finding the appropriate functions to run it, even long time after a function had been deprecated.
-Applying these principles in a tutorial not only teaches reproducibility, but also makes the teaching process itself reproducible \citep{dogucu_tools_2022}.
 
@@
 
-Incorporating the principles described here into teaching resources not onlu improves reproducibility practices, but it should facilitate adoption of software and analysis workflows in the natural sciences, among researchers at different academic levels, from undergrads to established researchers.
+Incorporating the principles described here into teaching resources not only improves reproducibility practices, but it should facilitate adoption of software and analysis workflows in the natural sciences, among researchers at different academic levels, from undergrads to established researchers.
 It can also help close the academic gap that is generated by uneven access to computational resources across students belonging to different groups \citep{kewalramani2018student}, where the most affected learners usually belong to underrepresented minorities and rural areas \citep{warner2021quantifying}.

### 2022-06-11 jsdse/latex_submission/Manuscript.tex
jsdse resubmission files

--- 2022-06-11 jsdse/Manuscript.tex
@@
 lack of trouble-shooting examples.
-We propose a set of principles to improve the unapproachable aspects of research workflows, and illustrate their application using an online teaching resource.
-We elaborate on the general application of these principles for documenting not only research products but also teaching materials, to provide present learners and future researchers with tools for successful scientific reproducibility.
+We propose principles to improve the unapproachable aspects of research workflows, and illustrate their application using an online teaching resource.
+We elaborate on the general application of these principles for documenting research products and teaching materials, to provide present learners and future researchers with tools for successful scientific reproducibility.
 %> ** add a sentence on implications for teaching reproducibility
